---
title: "*Borrelia* Competence of *Peromyscus*"
subtitle: "A NEON data project"
# author: "Clay J. Morrow"
site: "bookdown::bookdown_site"
output:
 bookdown::gitbook:
   number_sections: false
   config:
     toc:
       collapse: section
params:
  run_scripts: FALSE
  force: FALSE
  cache_dir: "infection-modeling/R/rmarkdown/cache/resistance-tolerance"
---

```{r knit_setup, include=FALSE}
# These knitr settings are carried through the full book.
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE, error = FALSE,
  out.width = "80%"
)
```

```{r,  include=FALSE, warning=FALSE}
if (params$run_scripts){
  # run all the scripts for this project
  source("infection-modeling/R/scripts/00_map-sites.R")
  source("infection-modeling/R/scripts/01_prepare-mammal-data.R")
  source("infection-modeling/R/scripts/01-2_prepare-Peromyscus-data.R")
  source("infection-modeling/R/scripts/02_clean-immune-data.R")
  source("infection-modeling/R/scripts/03_summarize-weather.R") # gives warnings
  source("infection-modeling/R/scripts/04_quantify-behavior.R") # makes plots
  source("infection-modeling/R/scripts/05_prepare-model-data.R")
  
  # Model fits
  source("infection-modeling/R/scripts/06_fit-SEMs.R")
  # source("infection-modeling/R/scripts/06_fit-models.R")
  
  # remove the incidental objects from memory
  rm(list = ls())
  gc()
}
```


# Introduction:

This site follows the analyses of data from the *Borrelia* Competence of 
*Peromyscus* project. The project is a collaboration between 
University of Wisconsin - Madison (UW) and the University of South Florida
(USF). The primary goal of the project is to explain and disease competence
of *Peromyscus* mice regarding the *Borrelia burgdorferi* pathogen within the
NEON ecology network. The following sections explain the data (briefly), our
analytical decisions, and results of the analyses. 

Use the arrows at the bottom (or sides) of each page move to the next 
section, or use the side bar to jump to a specific section. 

**Note** that this project is a work-in-progress, so some pages may not be 
complete. Check back periodically for updates.

