```{r}
library(tidyverse)
```


# Neon sites and data overview

This project utilizes the US National Ecological 
Observatory Network ([NEON](https://www.neonscience.org/)), which collects data
about various ecological features and systems. The system that we are primarily
interested in, *Peromyscus* mice and tick-borne disease, is covered by NEON's 
[small mammal](https://data.neonscience.org/data-products/DP1.10072.001) 
trapping network.

## Site locations

Our work focused primarily on eight sites shown by Figure \@ref(fig:sitesMap) 
from 2022 to 2024. Information about the sites can be found on NEON's 
[Field Sites](https://www.neonscience.org/field-sites/explore-field-sites)
web page.

```{r sitesMap, out.width="100%", fig.width=6.5, fig.cap=caption}
knitr::include_graphics(
  path = R.utils::getAbsolutePath(
    "infection-modeling/graphics/eight_NEON_sites_map.jpg"
  ), rel_path = FALSE, error = FALSE
)

caption = paste(
  "Upper northeast terrestrial NEON site locations used for this project."
)
```

## Small mammal data

```{r}
# mammal data
mammals <- readRDS("infection-modeling/data/mammal-data.rds")
# peromyscus data
Peros <- readRDS("infection-modeling/data/peromyscus-data.rds")
```


```{r get_obs_stats}
# observations and individuals (total)
nobs <- mammals %>% filter(!is.na(taxonID)) %>% nrow()
nind <- mammals %>% pull(iid) %>% unique() %>% length()
# observations and individuals (2022-2024)
tnobs <- mammals %>% filter(!is.na(taxonID), year %in% 2022:2024) %>% nrow()
tnind <- mammals %>% filter(year %in% 2022:2024) %>% 
  pull(iid) %>% unique() %>% length()
# Peromyscus observations and individuals
perobs <- Peros %>% 
  filter(
    year %in% 2022:2024
  ) %>% nrow()
perind <- Peros %>% 
  filter(
    year %in% 2022:2024
  ) %>% 
  pull(iid) %>% unique() %>% length()

spec_inds <- mammals %>% 
  filter(!is.na(scientificName)) %>% 
  mutate(
    all_time = TRUE,
    now = year %in% 2022:2024
  ) %>% 
  group_by(scientificName, all_time) %>% 
  mutate(all_ind = length(unique(iid))) %>% 
  group_by(scientificName, now) %>% 
  mutate(now_ind = length(unique(iid))) %>% 
  select(scientificName, all_time, now, all_ind, now_ind) %>% 
  distinct() %>% 
  mutate(keep = now | (all_ind == now_ind)) %>% 
  filter(keep) %>% ungroup() %>% 
  select(-c(all_time, now, keep))

nspec <- nrow(spec_inds)
```

Through their capture efforts at these sites, NEON has made 
`r prettyNum(nobs, ",")` small mammal observations from 
`r prettyNum(nspec, ",")` species and `r prettyNum(nind, ",")` unique 
individuals since its inception (Table \@ref(tab:speciesIndividuals)). 
From 2022 to 2024, they made `r prettyNum(tnobs, ",")` observations of 
`r prettyNum(tnind, ",")` individuals, `r prettyNum(perind, ",")` of which 
are either *Peromyscus maniculatus* or *Peromyscus leucopus*. NEON collects 
information about each individual mouse it catches. Some key traits used by 
this project are body size, sex, maturity, reproductive 
status (Table \@ref(tab:mamTraitsSub)).

<br>
```{r speciesIndividuals}
spec_inds %>% arrange(desc(all_ind)) %>% 
  knitr::kable(
    digits = 2,
    col.names = c("Species", "all time", "2022-2024"),
    caption = paste(
      "Approximate counts of individuals^[Individuals were sometimes",
      "labeled as multiple species across their capture record",
      "and double-counted by this table. The totals given in the text",
      "above are accurate.]",
      "from each species^[There are occasional misidentifications.]",
      "captured by NEON from its",
      "full duration and from our focal time window (2022 - 2024)."
    )
  )
```

<br>
```{r mamTraitsSub}
set.seed(119977)
Peros %>% 
  filter(
    year %in% 2022:2024,
    !is.na(sex), sex != "U",
    !is.na(weight), !is.na(tailLength), #!is.na(totalLength),
    !is.na(testes) | testes == "unknown" | !is.na(vagina) | vagina == "unknown",
    !is.na(lifeStage), 
    sex_male != 0.5
  ) %>% 
  mutate(
    obs = as.numeric(factor(uid)),
    sex = factor(sex_male, levels = c(0, 1), labels = c("F", "M"))
  ) %>% 
  select(
    obs, ind = iid, date, age = lifeStage, sex, weight, tail = tailLength,  
    testes, vagina
  ) %>% 
  sample_n(10) %>% 
  arrange(obs) %>% 
  knitr::kable(
    caption = paste(
      "Ten random small mammal observations subset.",
      "Columns correspond to observation ID,",
      "individual ID, capture data, life stage, sex, weight (g),",
      "tail length (mm), testes status (scrotal/nonscrotal),",
      "vagina condition (swollen or plugged)."
    )
  )
set.seed(NULL) # unset seed
```

## Disease prevalence across sites

**TODO**: *Borrelia* prevalence in ticks... 

## Habitat differences across sites

**TODO** 

## Community differences across sites

* small-mammal communitites: **TODO**

* Other animal communities (e.g., deer, predators): **TODO**

## Other data

In addition to the 
[small mammal](https://data.neonscience.org/data-products/DP1.10072.001) 
data product, we also use other NEON-collected data including

* [temperature](https://data.neonscience.org/data-products/DP1.00003.001),

* [precipitation](https://data.neonscience.org/data-products/DP1.00006.001),
and 

* [humidity](https://data.neonscience.org/data-products/DP1.00098.001) from
NEON's weather stations; as well as 

* tick-borne 
[pathogen status](https://data.neonscience.org/data-products/DP1.10064.002).

Further, our working group collected supplemental data in 2022-2024 including

* Times at which individuals were captured, which is indicative of their 
activity/behavior; 

* Refined estimates of *Borrelia* infection, which are utilize a detection
method that is more sensitive than those used by NEON, including quantitative
pathogen burden; and

* Expression of six key genes, thought to be important for resistance to and
tolerance of *Borrelia* infection.

Much of this data will be explored in later sections.