---
title: "NEON variable associations"
author: "Clay Morrow"
date: "2025-05-02"
output: 
  bookdown::html_document2
---

# Introduction

This document will track the initial steps for building an infection SEM.

To enhance readability, `tidyverse` functions and syntax will be used often 
throughout this document.

# Setup 

This document was last rendered: `r lubridate::now() |> as.Date()`

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First, we load the R packages that we'll use and point to the directory
in which NEON data are saved. Data objects will be loaded from here as needed
throughout the rest of this document.

```{r load_packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate) # manage dates
library(lemon) # ggplot extension
```

```{r neon_data_dir}
# path to NEON data directory
neon_data_dir = "../neon-data-curation/data/eight_sites"

# check that the directory exists
stopifnot(dir.exists(neon_data_dir))
```

```{r}
my_token <- readLines(
  "../neon-data-curation/docs/NEON-API-token-morrow5.txt"
)
```


```{r}
NEON_sites <- read.csv("data/raw-data/NEON-SiteMap-Table.csv") |> 
  tibble()
```


```{r}
trap_locations <- readRDS(
  file.path(
    "../neon-data-curation/data/eight_sites",
    "location_data/small-mammal-trap-locations.rds"
  )
)
```


# Small mammal data

The small mammal data product object is a list of multiple data and metadata
tables. The `$readme*` object describes the product and `$variables*` table
describes the different fields of each table. The primary table we want is 
called `mam_pertrapnight`, which has one row for each capture (or lack thereof) 
per trap per night. We'll simply the name of this dataset to `mammals` from
here on out.

```{r mammal-data}
# load the small mammal trapping data products
mammal_prods <- readRDS(file.path(neon_data_dir, "mammal-trap-data.rds"))

# Extract a tibble of the "mammals per trap night" data object
mammals <- mammal_prods$mam_pertrapnight |> tibble()

# number of observations in the mammal table 
n_mammals <- nrow(mammals)
```

The field descriptions of the `mammals` data are contained within the 
`$variables*` object:

```{r mammal_variable_lookup}
# Get a variable lookup metadata table, for our mammal data
mammal_var_lookup <- mammal_prods$variables_10072 |> 
  filter(table == "mam_pertrapnight") |> 
  tibble()

# Print the variable lookup table
mammal_var_lookup |> select(fieldName, units, description)
```

We will next clean up `mammals` a bit to make it easier to use. Notably, 
many of the columns that we probably won't use get collapsed into a single
list-column (via `tibble`). This allows removal of some clutter without losing
any data.

```{r remove_mammal_prod, include = FALSE}
rm(mammal_prods)
gc()
```


```{r, collapse_mammals}
# collapse some variables that we are unlikely to use
mammals <- mammals |> 
  arrange(collectDate, plotID, nightuid, trapCoordinate, uid) |> 
  ungroup() |>
  mutate(cap_id_num = row_number()) |> # create a row ID to simplify uid.
  relocate(cap_id_num, .before = 1) |> 
  group_by(uid) |> 
  nest(
    other_mammal_vars = c(
      namedLocation, domainID, plotType, nlcdClass, geodeticDatum,
      coordinateUncertainty, elevationUncertainty, trapType, individualCode,
      taxonRank:nativeStatusCode, replacedTag, frozenSamplePrepMethod, 
      bloodSampleID, bloodSampleMethod, fecalSampleID, fecalSampleCondition, 
      earSampleID, hairSampleID, hairSampleContents, voucherSampleID:release
    )
  ) |> ungroup()

# ensure n hasn't changed - this check will be repeated periodically
stopifnot(nrow(mammals) == n_mammals)
```

We'll also get some extra information about year, month, day, and Julian day 
(`DOY`: day-of-year). We're going to hold off on formatting the actual date 
columns (`collectDate` and `endDate`) for now because they may be used for 
joining with other data tables.

```{r, mammal_DOY}
# add new time variables
mammals <- mammals |> 
  mutate(
    year = year(collectDate),
    month = month(collectDate),
    day = day(collectDate),
    DOY = yday(collectDate)
  ) |> 
  relocate(year, month, day, DOY, .before = collectDate)
```

And we'll also refactor some of the categorical variables to be more useful
in modeling

```{r relevel_columns}
# refactor the tick info
mammals <- mammals |> 
  mutate(
    tickNumber = if_else(
      larvalTicksAttached != "Y" & 
        nymphalTicksAttached != "Y" & 
        adultTicksAttached != "Y" & 
        tickNumber != "No count performed", true = "0", false = tickNumber
    ),
    tickNumber = factor(
      tickNumber, levels = c("0", "1-5", "6-20", ">20"), ordered = TRUE
    ),
    nymphalTicksAttached = factor(
      nymphalTicksAttached, levels = c("N", "Y"), ordered = TRUE
    ),
    larvalTicksAttached = factor(
      larvalTicksAttached, levels = c("N", "Y"), ordered = TRUE
    ),
    adultTicksAttached = factor(
      adultTicksAttached, levels = c("N", "Y"), ordered = TRUE
    )
  )

mammals$tickNumber |> table()
```


# Pathogens

### tick-borne

Next, we'll load the tick-borne pathogens neon data product. As with the small
mammal product, there are multiple tables contained within this product. The
main data is stored in `rpt2_pathogentesting`, which we will rename to 
`pathogens`.

```{r tickborne-data}
# read the NEON tick-borne pathogen product object
pathogen_prods <- readRDS(
  file.path(neon_data_dir, "rodent-tickborne-pathogen-status.rds")
)

# extract the main pathogen testing table
pathogens <- pathogen_prods$rpt2_pathogentesting |> tibble()

n_pathogens <- length(unique(pathogens$testPathogenName))

# get a variable lookup metadata table for this dataset
pathogen_var_lookup <- pathogen_prods$variables_10064 |> 
  filter(table == "rpt2_pathogentesting") |> tibble()

# view the lookup table
pathogen_var_lookup |> select(fieldName, units, description)
```

```{r rm_pathogen_prods, include=FALSE}
rm(pathogen_prods)
gc()
```


The `pathogens` data has 1 row per test (ear and/or blood) per pathogen 
(`r n_pathogens` possible) per individual. Linking these samples to an 
individual capture (`mammals$uid`) requires matching the `sampleCode` variable 
with either `bloodSampleBarcode` or `earSampleBarcode` from the `mammal` table,
depending on whether `sampleID` ends with ".E" (ear) or ".B" (blood). 

```{r match_pathogen_mammal}
# add in an indicator of "blood" or "ear" for sampling and matched mammal row
pathogens <- pathogens |> 
  mutate(
    year = year(.data$collectDate),
    sample_method = dplyr::case_when(
      grepl(".*\\.E$", sampleID) ~ "ear",
      grepl(".*\\.B$", sampleID) ~ "blood",
      .default = NA
    ),
    matched_mammal_index = dplyr::case_when(
      sample_method == "ear" ~ match(sampleCode, mammals$earSampleBarcode),
      sample_method == "blood" ~ match(sampleCode, mammals$bloodSampleBarcode),
      .default = NA
    ),
    mammal_uid = mammals$uid[.data$matched_mammal_index],
  )

# check that all samples had a match (ear subset)
pathogens |> filter(sample_method == "ear") |> 
  pull(.data$sampleCode) %in% mammals$earSampleBarcode |> all() |> 
  stopifnot()

# (blood subset)
pathogens |> filter(sample_method == "blood") |> 
  pull(.data$sampleCode) %in% mammals$bloodSampleBarcode |> all() |> 
  stopifnot()
```

The `pathogen` data is then transformed such that there is one row for each
individual (`mammal_uid`), with the results converted logical (`TRUE` if
"Positive" results from *either* blood or ear samples). We will also collapse
the original data into a list column to avoid clutter.

```{r}
# collapse unneeded columns into a list column, by mammal_uid, for later use
pathogens_nested <- pathogens |> 
  group_by(mammal_uid) |> 
  nest(all_tickborne_data = -c(mammal_uid))

# get a column for each pathogen, indicating whether it was found
pathogens <- pathogens |> 
  select(
    mammal_uid, sample_method, sampleID, sampleCode, testPathogenName, 
    testResult
  ) |> 
  group_by(mammal_uid, testPathogenName) |> 
  summarize(
    # n_tests = n(), 
    pathogen_positive = "Positive" %in% testResult,
    .groups = "drop"
  ) |> 
  group_by(mammal_uid) |> 
  pivot_wider(
    names_from = testPathogenName, values_from = pathogen_positive
  ) |> ungroup()

# get the observation number from the pathogen data
n_tickborne_obs <- nrow(pathogens)

# check that there's only 1 row per individual
stopifnot(n_tickborne_obs == length(unique(pathogens$mammal_uid)))

# re-join with the nested full data
pathogens <- left_join(pathogens, pathogens_nested, by = "mammal_uid") |> 
  relocate(all_tickborne_data, .after = mammal_uid)
```

```{r rm_nested_pathogens, include = FALSE}
rm(pathogens_nested)
gc()
```


## Hanta virus

We will also extract the hantavirus data -- which are temporally 
non-overlapping with the tick-borne data but still interesting -- and join it
into our `pathogen` dataset.

```{r hanta}
# get the hanta virus product
hanta_prods <- readRDS(
  file.path(neon_data_dir, "rodent-hanta-virus.rds")
)

# extract the hanta virus data and match the mammal uid
hanta <- hanta_prods$rpt_bloodtesting |> 
  mutate(
    matched_mammal_index = match(
      bloodSampleBarcode, mammals$bloodSampleBarcode
    ),
    mammal_uid = mammals$uid[matched_mammal_index]
  )

# collapse full hanta data into a list column, by mammal_uid, for later use
hanta_nested <- hanta |> 
  group_by(mammal_uid) |> 
  nest(all_hanta_data = -c(mammal_uid))

# transpose to the same shape as the pathogen table (for consistency)
hanta <- hanta |> 
  select(
    mammal_uid, bloodSampleBarcode, testPathogenName, testResult
  ) |> 
  group_by(mammal_uid, testPathogenName) |> 
  summarize(
    pathogen_positive = "Positive" %in% testResult,
    .groups = "drop"
  ) |> 
  group_by(mammal_uid) |> 
  pivot_wider(
    names_from = testPathogenName, values_from = pathogen_positive
  ) |> ungroup()

# number of individual observations for hanta data
n_hanta_obs <- nrow(hanta)

# check that the dimensions are correct
stopifnot(n_hanta_obs == nrow(hanta_nested))

# rejoin the nested full data into the hanta object
hanta <- left_join(hanta, hanta_nested, by = "mammal_uid")

stopifnot(nrow(hanta) == n_hanta_obs)

# join the hanta into the pathogen results
pathogens <- full_join(
  pathogens, hanta, by = "mammal_uid"
)  |> 
  relocate(all_hanta_data, .after = all_tickborne_data)

# note: hanta was only tested for in years prior to tick-borne pathogens:
stopifnot(nrow(pathogens) == n_tickborne_obs + n_hanta_obs)
```

```{r rm_hanta_prods, include = FALSE}
rm(hanta_prods)
rm(hanta)
gc()
```


## Collapse pathogen data

Before joining the updated `pathogens` data into `mammals`, we'll first 
collapse the data as we've done previously -- keeping only the key columns.
The main pathogens that we'll keep as individual columns are hanta virus 
(`hanta`), *Borrelia burgdorferi* (`B_burgdorferi`), and all *Borrelia*
combined (`any_borrelia`).

```{r main_pathogen_columns}
# copy of the "Hantaan virus" column
pathogens$hanta <- pathogens$`Hantaan virus`

# add a column indicating if there was any borrelia infection
pathogens$any_borrelia <- pathogens |> 
  select(contains("borrelia", ignore.case = TRUE)) |> 
  apply(1, function(x) {
    if (all(is.na(x))) {return(NA)} else {return(any(x, na.rm = TRUE))}
  })

# copy of the column indicating confirmed Borrelia burgdorferi
pathogens$B_burgdorferi <- pathogens |> 
  pull(contains("burgdorferi", ignore.case = TRUE))
```

```{r collapse_pathogens}
# collapse the larger results matrix into a list column  
pathogens <- pathogens |> 
  relocate(c(hanta:B_burgdorferi), .after = "mammal_uid") |> 
  group_by(mammal_uid) |> 
  nest(all_infections_matrix = -c(mammal_uid:all_hanta_data)) |> 
  relocate(all_infections_matrix, .before = all_tickborne_data)

# check dimensions
stopifnot(nrow(pathogens) == n_tickborne_obs + n_hanta_obs)
```

Joining the pathogen data into the mammal data results in our `full_data`
object, that will serve as our primary data for models. We will continue adding
data from various sources into this object.

```{r full_data}
# join the pathogen table into the mammal table
full_data <- mammals |> full_join(pathogens, by = c("uid" = "mammal_uid"))

# check that it is the same size as the mammals table
stopifnot(nrow(full_data) == nrow(mammals))

# and check that the non-null list-column values are equal to the pathogen size
n_tested = full_data |> 
  rowwise() |> 
  filter(!is.null(all_infections_matrix)) |> 
  nrow()
stopifnot(n_tested == nrow(pathogens))
```

```{r rm_previous_sets, include = FALSE}
rm(pathogens, mammals)
gc()
```



# Capture times

This section focuses on adding the capture time data that we calculated 
previously (see code in the  "../activity-timing/" folder).

```{r, fig.asp = 1, out.width="50%", fig.width=6}
# load and trim the capture time data
captimes <- readRDS(
  file.path(
    "../activity-timing/data/ibuttons",
    "capture-time-trapping-data_2022-2024.rds"
  )
) |> 
  hoist(neon_trap_data, "uid") |> # pull the uid column out for matching
  select(
    uid, bout_label, plot_bout, local_midnight, sunset, sunrise,
    trapping_interval, night_length:cap_prop_night,  ibutton_pair, 
    SN_In, SN_Out, ibutton_data, possible_cap_times:scientificName.update
  ) |> 
  filter(!is.na(uid))

# combine the captimes
full_data <- full_data |> 
  mutate(collectDate = as_datetime(.data$collectDate)) |> 
  left_join(captimes, by = "uid")

# check the merge was right
stopifnot(nrow(full_data) == n_mammals)
stopifnot(
  nrow(full_data |> filter(!is.na(cap_prop_night))) == 
    nrow(captimes |> filter(!is.na(cap_prop_night)))
)
```

```{r rm_captimes, include = FALSE}
rm(captimes)
gc()
```

## Key pathogen x Capture time summaries

Let's quickly look at the breakdown of the 3 key pathogens:

```{r}
# sample size for Bb infection
n_bb_samples = full_data |> filter(!is.na(B_burgdorferi)) |> nrow()

# sample size for capture times
n_captimes = full_data |> filter(!is.na(cap_prop_night)) |> nrow()

# sample size for both Bb infection and capture times
n_complete = full_data |> 
  filter(!is.na(B_burgdorferi), !is.na(cap_prop_night)) |> 
  nrow()

full_data |> 
  filter(!is.na(B_burgdorferi), !is.na(cap_prop_night), B_burgdorferi) |> 
  nrow()

c(Bb = n_bb_samples, captimes = n_captimes, complete = n_complete)

# Borrelia infection table, by year
full_data |> 
  filter(!is.na(B_burgdorferi)) |> 
  group_by(year, B_burgdorferi) |> 
  tally() |> pivot_wider(names_from = B_burgdorferi, values_from = n)
```


```{r, eval = FALSE, echo = FALSE}
# borrelia table
full_data |> 
  filter(!is.na(any_borrelia)) |> 
  group_by(year, siteID, any_borrelia) |> 
  tally() |> 
  pivot_wider(names_from = year, values_from = n) |> 
  arrange(siteID)

# B. burgdorferi table
full_data |> 
  filter(!is.na(B_burgdorferi)) |> 
  group_by(year, siteID, B_burgdorferi) |> 
  tally() |> 
  pivot_wider(names_from = year, values_from = n) |> 
  arrange(siteID)

# hanta table
full_data |> 
  filter(!is.na(hanta)) |> 
  group_by(year, siteID, hanta) |> 
  tally() |> 
  pivot_wider(names_from = year, values_from = n) |> 
  arrange(siteID)
```


```{r, fig.width = 4.5, fig.asp = 1}
tmp <- full_data |> 
  mutate(
    Bb_missing = is.na(B_burgdorferi),
    captime_missing = if_else(is.na(cap_prop_night), "no captime", "captime"),
    B_burgdorferi = case_when(
      is.na(B_burgdorferi) ~ "not sampled",
      B_burgdorferi == TRUE ~ "positive",
      B_burgdorferi == FALSE ~ "negative"
    )
  ) |> 
  group_by(year, siteID, captime_missing, B_burgdorferi) |> tally()

pd = position_dodge(0.8)

theme_set(theme_classic())

tmp |> 
  filter(year %in% 2022:2024, B_burgdorferi != "not sampled", captime_missing == "captime") |> 
  ggplot(aes(x = siteID, y = n, fill = B_burgdorferi)) + 
  facet_rep_wrap(~year, ncol = 1) + 
  geom_col(position = pd) + 
  theme(legend.position = "bottom", strip.background = element_blank()) +
  labs(x = "Site", y = "Count", title = "+captime +pathogens", fill = "Borrelia") +
  geom_text(aes(label = n), position = pd, vjust = 0) + 
  scale_y_continuous(limits = c(0, 90)) + 
  scale_fill_manual(values = c("orange", "cornflowerblue"))

ggsave("cap_path_counts.jpg", width = 4.5, height = 4.5)

tmp |> 
  filter(year %in% 2022:2024, B_burgdorferi != "not sampled", captime_missing != "captime") |> 
  ggplot(aes(x = siteID, y = n, fill = B_burgdorferi)) + 
  facet_rep_wrap(~year, ncol = 1) + 
  geom_col(position = pd) + 
  theme(legend.position = "bottom", strip.background = element_blank()) +
  labs(x = "Site", y = "Count", title = "-captime +pathogens", fill = "Borrelia") +
  geom_text(aes(label = n), position = pd, vjust = 0) + 
  scale_y_continuous(limits = c(0, 180)) +
  scale_fill_manual(values = c("orange", "cornflowerblue"))

ggsave("nocap_path_counts.jpg", width = 4.5, height = 4.5)

tmp |> 
  filter(year %in% 2022:2024, B_burgdorferi == "not sampled", captime_missing == "captime") |> 
  ggplot(aes(x = siteID, y = n, fill = B_burgdorferi)) + 
  facet_rep_wrap(~year, ncol = 1) + 
  geom_col(position = pd) + 
  theme(legend.position = "bottom", strip.background = element_blank()) +
  labs(x = "Site", y = "Count", title = "+captime -pathogens", fill = "Borrelia") +
  geom_text(aes(label = n), position = pd, vjust = 0) + 
  scale_y_continuous(limits = c(0, 450)) +
  scale_fill_manual(values = "darkgreen")

ggsave("cap_nopath_counts.jpg", width = 4.5, height = 4.5)

tmp |> 
  filter(year %in% 2022:2024, B_burgdorferi == "not sampled", captime_missing != "captime") |> 
  ggplot(aes(x = siteID, y = n, fill = B_burgdorferi)) + 
  facet_rep_wrap(~year, ncol = 1) + 
  geom_col(position = pd) + 
  theme(legend.position = "bottom", strip.background = element_blank()) +
  labs(x = "Site", y = "Count", title = "-captime -pathogens", fill = "Borrelia") +
  geom_text(aes(label = n), position = pd, vjust = 0) + 
  scale_y_continuous(limits = c(0, 8500)) + 
  scale_fill_manual(values = "darkgreen")

ggsave("nocap_nopath_counts.jpg", width = 4.5, height = 4.5)


```


```{r}
full_data |> filter(year %in% 2022) |> 
  mutate(
    Bb_missing = is.na(B_burgdorferi), 
    captime_missing = is.na(cap_prop_night),
    state = case_when(
      !captime_missing & !Bb_missing ~ "timing_pathogens",
      !captime_missing & Bb_missing ~ "timing_noPathogens",
      captime_missing & !Bb_missing ~ "noTiming_pathogens",
      captime_missing & Bb_missing ~ "noTiming_noPathogens"
    )
  ) |> 
  group_by(year, siteID, state) |> 
  tally() |> 
  arrange(state) |> 
  pivot_wider(names_from = state, values_from = n) |> 
  write.csv(file = "state_table_2022.csv", row.names = FALSE)

full_data |> filter(year %in% 2023) |> 
  mutate(
    Bb_missing = is.na(B_burgdorferi), 
    captime_missing = is.na(cap_prop_night),
    state = case_when(
      !captime_missing & !Bb_missing ~ "timing_pathogens",
      !captime_missing & Bb_missing ~ "timing_noPathogens",
      captime_missing & !Bb_missing ~ "noTiming_pathogens",
      captime_missing & Bb_missing ~ "noTiming_noPathogens"
    )
  ) |> 
  group_by(year, siteID, state) |> 
  tally() |> 
  arrange(state) |> 
  pivot_wider(names_from = state, values_from = n)|> 
  write.csv(file = "state_table_2023.csv", row.names = FALSE)
```


# Weather

```{r}

```


# Ticks 

```{r ticks}
# read in tick object
tick_obj <- readRDS(
  file.path(neon_data_dir, "tick-drag-cloth-sampling.rds")
)

# Get the tick table
ticks <- tick_obj$tck_taxonomyProcessed |> tibble() |> 
  mutate(
    year = year(collectDate), month = month(collectDate),
    date = date(collectDate), week = week(collectDate),
    siteID = gsub("(.*)_.*", "\\1", plotID)
  ) |> 
  group_by(siteID, plotID, year, month, sexOrAge) |> 
  mutate(total_tickgrp_count = sum(individualCount)) |> 
  group_by(siteID, plotID, year, month) |> 
  arrange(siteID, plotID, year, month, collectDate) |> 
  mutate(
    tck_samp_interval = interval(first(date), last(date)),
    tck_samp_days = time_length(tck_samp_interval, "days") + 1
  ) |> 
  ungroup() |> 
  mutate(
    tck_density = individualCount / tck_samp_days
  )

# Get the variable lookup table
tick_vars <- tick_obj$variables_10093 |> tibble() |> 
  filter(table == "tck_taxonomyProcessed") |> 
  select(fieldName, units, description)
```

```{r tick_counts}
# total ticks by site by year (all species)
ticks |> 
  filter(year %in% 2021:2024) |> 
  select(
    siteID, plotID, year, month, sexOrAge, genus, scientificName,
    individualCount
  ) |> 
  group_by(year, siteID) |> 
  summarise(n_ticks = sum(individualCount, na.rm = TRUE), .groups = "keep") |> 
  pivot_wider(names_from = year, values_from = n_ticks) |> 
  mutate(total = rowSums(across(`2021`:`2024`))) |> 
  arrange(desc(total))
```

```{r ixodes}
# get monthly plot densities of Ixodes (or larval) ticks from 2021 - 2024 for each group
ixodes <- ticks |> 
  filter(sexOrAge == "Larva" | genus == "Ixodes", year %in% 2021:2024) |> 
  group_by(siteID, plotID, year, month, sexOrAge) |> 
  mutate(
    count = sum(individualCount),
    ixodes_prop = if_else(
      sexOrAge == "Larva", NA, count/total_tickgrp_count
    )
  ) |> 
  select(
    siteID, plotID, year, month, sexOrAge, tck_samp_interval, 
    tck_samp_days, count, total_tickgrp_count, ixodes_prop
  ) |> 
  distinct() |> 
  ungroup(sexOrAge) |> 
  mutate(
    # density = count / tck_samp_days,
    # # trying to estimate larval tick ixodes props.
    mean_ixodes_prop = mean(ixodes_prop, na.rm = TRUE),
  ) |> 
  rowwise() |> 
  mutate(
    count = round(total_tickgrp_count * mean_ixodes_prop),
    # count = if_else(
    #   sexOrAge == "Larva", 
    #   round(count * mean_ixodes_prop),
    #   round(total_tickgrp_count * mean_ixodes_prop)
    # ),
    density = count / tck_samp_days
  ) |> 
  ungroup()

tick_table <- ixodes |> 
  pivot_wider(
    names_from = sexOrAge, values_from = c(count, density)
  ) |> 
  mutate(
    across(count_Larva:density_Female, ~replace_na(.x, 0)), # NA actually 0
    count_Ixodes = rowSums(across(contains("count_"))),
    density_Ixodes = count_Ixodes / tck_samp_days,
    count_Adult = count_Male + count_Female,
    density_Adult = count_Adult / tck_samp_days,
    count_AllTicks = total_tickgrp_count,
    density_AllTicks = total_tickgrp_count / tck_samp_days
  ) |> 
  relocate(c(count_Adult, count_Ixodes), .after = count_Female) |> 
  relocate(c(density_Adult, density_Ixodes), .after = density_Female) |> 
  select(-total_tickgrp_count)

# Total counts
tick_table |> 
  group_by(year) |> 
  summarize(across(contains("count"), ~sum(.x, na.rm = TRUE)))

# Total densities
tick_table |> 
  group_by(year) |> 
  summarize(across(contains("density"), ~mean(.x, na.rm = TRUE)))

## NOTE These larvae are from ALL ticks. Should I assume the proportion of Ixodes
## adults is comparable to the proportion of ixodes larvae?
```
## Tick pathogens

```{r tick_pathogens}
tick_pathogen_obj <- readRDS(
  file.path(neon_data_dir, "tick-pathogen-status.rds")
)

tick_pathogens <- tick_pathogen_obj$tck_pathogen |> tibble() |> 
  mutate(year = year(collectDate), month = month(collectDate))

# get only B.b. results from 2021-2024
borrelia <- tick_pathogens |> 
  filter(
    grepl("burgdorferi", testPathogenName),
    year %in% 2021:2024
  ) |> select(
    year, month, collectDate, plotID, individualCount, testResult
  )

# calculate some basic values
borrelia <- borrelia |> group_by(year, month, plotID) |> 
  mutate(
    n_testticks_plotmonth = sum(individualCount, na.rm = TRUE),
    n_Bbpos_plotmonth = sum(testResult == "Positive", na.rm = TRUE),
    Bb_tickrate_plotmonth = n_Bbpos_plotmonth / n_testticks_plotmonth
  ) |> 
  group_by(year) |> 
  mutate(
    n_testticks_plotyear = sum(individualCount, na.rm = TRUE),
    n_Bbpos_plotyear = sum(testResult == "Positive", na.rm = TRUE),
    Bb_tickrate_plotyear = n_Bbpos_plotyear / n_testticks_plotyear
  ) |> 
  ungroup()

# join the new Borrelia tick columns into the 
full_data <- left_join(
  full_data,
  borrelia |> 
    select(year, month, plotID, n_testticks_plotmonth:Bb_tickrate_plotyear) |> 
    distinct(),
  by = c("plotID", "year", "month")
)

# check the merge
stopifnot(nrow(full_data) == n_mammals)
```

# Genetics (resistance)

# Plant communities

```{r}
tree_prods <- readRDS(
  file.path(neon_data_dir, "litterfall-data.rds")
)
```


# Litterfall

```{r}
litter_prod <- readRDS(
  file.path(neon_data_dir, "litterfall-data.rds")
)

names(litter_prod)

litter_lookup <- litter_prod$variables_10033 |> tibble()

litter_lookup |> 
  filter(table == "ltr_massdata") |> 
  select(fieldName, description)

litter_loc_path = "data/litter-trap-locations.rds"
if (!file.exists(litter_loc_path)) {
  litter_locations <- litter_prod$ltr_pertrap |> 
    select(
      trapID, namedLocation, decimalLatitude, decimalLongitude, elevation
    ) |> 
    geoNEON::getLocByName(token = my_token) |> 
    mutate(elevation = as.numeric(elevation))
  saveRDS(litter_locations, litter_loc_path)
} else {litter_locations <- readRDS(litter_loc_path)}

litter <- litter_prod$ltr_massdata |> 
  tibble() |> 
  left_join(litter_locations) |>
  mutate(
    year = year(collectDate), month = month(collectDate),
    trapping_days = time_length(collectDate - setDate, "days"),
    mass_per_trapdays = dryMass / trapping_days
  ) 
```

```{r, fig.asp = 0.7, fig.width = 6}
# plot average monthly seed mass across years
litter |> 
  filter(functionalGroup == "Seeds", year %in% 2021:2024) |> 
  ggplot(aes(x = month, y = dryMass, col = siteID)) + 
  facet_wrap(~year, scales = "free_y") +
  stat_summary(fun.data = mean_se, size = 0.25, show.legend = FALSE) +
  stat_summary(fun = mean, geom = "line", linewidth = 0.75) + 
  labs(
    y = "Mean seed mass in litterfall per trap day", 
    x = "Month collected",
    col = NULL
  ) + 
  scale_color_manual(values = NEON_sites$color_hex) + 
  theme_bw() + 
  theme(legend.margin = margin(l = -2)) + 
  scale_x_continuous(breaks = 1:12)

ggsave(
  file.path("graphics", "siteMonthlySeedMass_2021-2024.jpg"), 
  width = 6, height = 0.7*6
)
```

```{r, fig.asp=0.66, fig.width = 8}
ggplot() +
  facet_wrap(~siteID, scales = "free") + 
  geom_point(
    data = trap_locations, 
    aes(x = adjDecimalLongitude, y = adjDecimalLatitude, col = "mammal")
  ) +
  geom_point(
    data = litter_locations, 
    aes(x = decimalLongitude, y = decimalLatitude, col = "litter")
  ) + 
  theme_bw() +
  theme(
    legend.position = "inside", legend.position.inside = c(0.9, 0),
    legend.justification = c(1, 0),
    # axis.text = element_blank()
  ) +
  labs(x = "Longitude", y = "Latitude", col = "Plot") #+ 
# scale_x_continuous(
#   breaks = seq(-100, -50, by = 0.1), 
#   minor_breaks = seq(-100, -50, by = 0.05)
# )

ggsave(
  file.path("graphics", "litterfall-mammal-trap-proximity.jpg"),
  width = 8, height = 0.66*8
)
```

```{r collapse_litter}
# summarize the seed mass by site, plot, year, and month and add it to full data
full_data <- litter |> 
  group_by(siteID, year, month) |> 
  mutate(
    nSeedSamples_siteMean = n(),
    seedMass_perTrapDays_siteMean = mean(mass_per_trapdays, na.rm = TRUE),
    seedTrapDays_siteMean = mean(trapping_days, na.rm = TRUE)
  ) |> 
  # group_by(plotID, .add = TRUE) |> 
  # mutate(
  #   seedMass_perTrapDays_plotMean = mean(mass_per_trapdays, na.rm = TRUE),
  #   seedTrapDays_plotMean = mean(trapping_days, na.rm = TRUE)
  # ) |> 
  group_by(siteID, year, .add = FALSE) |> 
  mutate(
    nSeedSamples_siteMeanYear = n(),
    seedMass_perTrapDays_siteMeanYear = mean(mass_per_trapdays, na.rm = TRUE),
    seedTrapDays_siteMeanYear = mean(trapping_days, na.rm = TRUE)
  ) |> 
  # group_by(plotID, .add = TRUE) |> 
  # mutate(
  #   seedMass_perTrapDays_plotMeanYear = mean(mass_per_trapdays, na.rm = TRUE),
  #   seedTrapDays_plotMeanYear = mean(trapping_days, na.rm = TRUE)
  # ) |> 
  select(
    siteID, year, month,
    seedMass_perTrapDays_siteMean, seedTrapDays_siteMean,
    seedMass_perTrapDays_siteMeanYear, seedTrapDays_siteMeanYear,
    nSeedSamples_siteMean, nSeedSamples_siteMeanYear
  ) |> 
  distinct() |> 
  right_join(full_data)
```

```{r seedmass_captime, fig.asp = 1, fig.width=6}
full_data |> 
  filter(!is.na(cap_prop_night), !is.na(seedMass_perTrapDays_siteMean)) |> 
  ggplot(
    aes(
      x = seedMass_perTrapDays_siteMean, y = cap_prop_night, col = siteID
    )
  ) +
  facet_wrap(~siteID, scales = "free") +
  geom_smooth(
    method = "lm", formula = "y ~ x", se = FALSE, #show.legend = FALSE,
    size = 0.5, # linetype = "dashed", 
    aes(group = interaction(siteID, year), linetype = factor(year))
  ) + 
  geom_smooth(
    method = "lm", formula = "y ~ x", fill = "grey80", size = 1,
    #show.legend = FALSE,
    aes(group = siteID)
  ) + 
  stat_summary(
    fun.data = mean_se, fill = "white", # size = 0.4, # show.legend = FALSE, 
    aes(group = interaction(siteID, year, month), shape = factor(year))
  ) + 
  # geom_smooth(method = "lm", aes(group = 1), col = "black") + 
  labs(
    x = "Seed mass per trapping day (monthly average)", 
    y = "Relative capture time (mean ± SE within site and year)"
  ) + 
  scale_color_manual(guide = "none", values = NEON_sites$color_hex) + 
  scale_shape_manual(
    name = "Year", labels = c(2022:2024), values = c(21, 22, 24)
  ) + 
  scale_linetype_manual(name = "Year", values = c(2:4), guide = FALSE) + 
  theme_bw() + 
  theme(
    legend.position.inside = c(4/5, 0), legend.position = "inside", 
    legend.justification = c(0.5, 0),
    strip.background = element_blank()
  )

ggsave(
  file.path("graphics", "captureTime-vs-seedMass.jpg"),
  width = 6, height = 6, dpi = 300
)
```

```{r fig.asp = 1, fig.width=6}
pd = position_dodge(0.8)

tmp <- full_data |> 
  filter(!is.na(cap_prop_night)) |> 
  group_by(siteID, year) |> arrange(year, month) |> 
  summarize(first_month = first(month), last_month = last(month))


litter |> 
  filter(year %in% 2022:2024) |> 
  group_by(siteID, year, month) |> 
  summarize(n = length(unique(collectDate)), .groups = "drop") |>
  complete(siteID, year, month) |> 
  mutate(n = replace_na(n, 0)) |> 
  ggplot(aes(x = month, y = n, fill = factor(year))) + 
  geom_rect(
    data = tmp, inherit.aes = FALSE, alpha = 0.25, col = "grey50",
    aes(
      xmin = first_month, xmax = last_month, 
      ymin = year - 2019, ymax = year - 2018, 
      fill = factor(year)
    )
  ) + 
  # geom_line(aes(col = factor(year)), position = pd, linetype = "dotted") +
  geom_col(position = pd, alpha = 1, width = 0.8) +
  # geom_point(aes(col = factor(year)), position = pd) + 
  facet_wrap(~siteID) + 
  scale_x_continuous(breaks = 1:12) + 
  labs(
    y = "Litter collections per month", x = "Month", fill = "Year", color = "Year"
  ) + 
  theme_bw() + 
  theme(
    legend.position.inside = c(0.9, 0), legend.position = "inside",
    legend.justification = c(1, 0),
    strip.background = element_blank()
  )

ggsave(
  file.path("graphics", "monthly-litter-captures_with-capture-time-windows.jpg")
)
```

```{r, fig.asp = 0.35, fig.width=7}
full_data |> 
  filter(!is.na(cap_prop_night), !is.na(seedMass_perTrapDays_siteMean)) |> 
  ggplot(
    aes(
      x = seedMass_perTrapDays_siteMean, y = cap_prop_night
    )
  ) + 
  facet_wrap(~year, scales = "free_x", ncol = 3) + 
  geom_point(
    shape = 16, size = 0.75, #col = "grey50"
    aes(col = siteID)
  ) + 
  geom_smooth(
    method = "lm", formula = "y ~ x", aes(group = 1), col = "grey50", 
    fill = "grey80", linewidth = 0.75
  ) + 
  stat_summary(
    fun.data = mean_se, shape = 21, size = 0.5, alpha = 1, stroke = 0.5, 
    linewidth = 0.5, 
    aes(fill = siteID)
  ) + 
  theme_bw() + 
  theme(strip.background = element_blank()) +
  scale_fill_manual(values = NEON_sites$color_hex) + 
  scale_color_manual(values = NEON_sites$color_hex, guide = FALSE) + 
  labs(
    fill = NULL, x = "Average seed mass (per trapping day per bout)",
    y = "Relative capture time"
  )

ggsave(
  file.path(
    "graphics", "Capture-time-by-seedmass_within-year.jpg"
  ),
  width = 7, height = 0.35*7, dpi = 300
)
```


# Beetles

# Updated *Borrelia* infection status (USF)

# Distances

```{r, fig.asp = 0.5, fig.width=7}
location_dir = file.path(neon_data_dir, "location_data")

plot_locs <- readRDS(
  file.path(location_dir, "plot-centroid-locations.rds")
) |> tibble()

plot_locs |> 
  select(
    siteID, plotID, decimalLatitude, decimalLongitude, elevation,
    xOffset, yOffset, zOffset
  ) 

# use BLAN as an example to get distances among plots
BLAN_dists <- plot_locs |> 
  filter(siteID == "BLAN") |> 
  select(decimalLongitude, decimalLatitude) |> 
  geosphere::distm()

colnames(BLAN_dists) <- plot_locs |> filter(siteID == "BLAN") |> pull(plotID)
rownames(BLAN_dists) <- plot_locs |> filter(siteID == "BLAN") |> pull(plotID)

BLAN_dists

max(BLAN_dists) # 2.2 km
min(BLAN_dists[lower.tri(BLAN_dists)]) # 0.4 km

# look at the plot distribution within sites
plot_locs |> 
  ggplot(aes(x = decimalLongitude, y = decimalLatitude, col = elevation)) + 
  geom_point() + 
  facet_wrap(~siteID, scales = "free", ncol = 4) + 
  theme(
    legend.position = "none", strip.background = element_blank(),
    axis.text.x = element_blank()
  )
```


# Summary variables

```{r mouse_population}
# estimate population size and
full_data <- full_data |> 
  group_by(year, taxonID, plotID) |> 
  mutate(
    year_popsize = n(),
    Bb_mouserate_yr = sum(B_burgdorferi, na.rm = TRUE) / year_popsize
  ) |> 
  group_by(year, month, plotID) |> 
  mutate(
    month_popsize = n(),
    Bb_mouserate_mnth = sum(B_burgdorferi, na.rm = TRUE) / month_popsize
  )
```


# Preliminary infection modeling

This should probably be in a new file...

```{r}
names(full_data)
```

```{r}
glm.Bb <- glm(
  B_burgdorferi ~ weight + sex + 
    # decimalLatitude + decimalLongitude + # not affected by lat, long
    # taxonID + # not affected by species (among Peromyscus)
    # larvalTicksAttached + nymphalTicksAttached +
    # tickNumber +
    # (as.numeric(tickNumber) > 2) + # more than 1-5 ticks 
    # month_popsize + 
    # Bb_mouserate_mnthformula + 
    year_popsize +
    Bb_mouserate_yr +
    year_popsize:Bb_mouserate_yr +
    # Bb_tickrate_plotmonth +
    # Bb_tickrate_plotyear +
    year +
    month, # + # + hindfootLength + DOY + month + year,
  # DOY, 
  data = full_data |> 
    filter(taxonID %in% c("PELE", "PEMA", "PELEPEMA"), sex != "U"), 
  family = "binomial"
)

# model results
summary(glm.Bb)

# sample size of fit
glm.Bb$model |> nrow()
```

```{r}
# which columns are the limitting factors (most NAs)?
glm.Bb$data[, names(glm.Bb$model)] |> 
  filter(year %in% 2021:2024) |> 
  summary()

# which plots are not represented in the tick pathogen data?
tibble(
  plotID = unique(full_data$plotID), 
  in_borrelia_dat = plotID %in% borrelia$plotID
)
```

The above model still needs:
* [ ] weather (temp and precip)
* [x] *Borrelia* prevelance (tick and mice populations)
* [ ] genetic resistance
* [ ] population density
* [ ] movement

<!-- # OLDER -->

<!-- ```{r} -->
<!-- full_data |>  -->
<!--   filter(!is.na(B_burgdorferi), !is.na(cap_prop_night)) |>  -->
<!--   mutate(year = factor(year(collectDate))) |>  -->
<!--   group_by(siteID, B_burgdorferi, year) |>  -->
<!--   count(.drop = FALSE) |>  -->
<!--   ggplot(aes(x = B_burgdorferi, y = n, group = year)) + -->
<!--   facet_wrap(~ siteID, scales = "free_y") + -->
<!--   geom_col(aes(fill = year), position = pd) + -->
<!--   geom_text(aes(label = n)) +  -->
<!--   theme_bw() +  -->
<!--   theme(strip.background = element_blank()) +  -->
<!--   labs(x = "B. burgdorferi infection", y = "count") -->
<!-- ``` -->

<!-- From the above, it appears that there is only overlap between tested -->
<!-- individuals and activity timings for 2022 - with very few samples (93). -->

<!-- There appear to be only 603 small mammal pathogen samples in total at the 8  -->
<!-- sites. Of those, only 93 correspond to trappings where ibuttons were present.  -->
<!-- Among these, all the overlap was in 2022 (note that 2024 data is not yet  -->
<!-- available) and only 14 mice tested positive for *B. burgdorferi*. In total,  -->
<!-- only 136 of the 603 small mammals (22%) tested positive for *B. burgdorferi*. -->

<!-- ```{r} -->
<!-- library("GGally") -->
<!-- mammals |>  -->
<!--   filter(taxonID %in% c("PELE", "PEMA", "PELEPEMA")) |>  -->
<!--   ungroup() |>  -->
<!--   # group_by(siteID) |>  -->
<!--   select( -->
<!--     bb_result, weight, hindfootLength, tailLength, earLength, sex -->
<!--   ) |>  -->
<!--   ggpairs() -->
<!-- ``` -->


<!-- ```{r} -->
<!-- library(lme4) -->

<!-- mammals <- mammals |>  -->
<!--   mutate(row_ID = row_number()) -->

<!-- model_data <- mammals |>  -->
<!--   filter( -->
<!--     taxonID %in% c("PELE", "PEMA", "PELEPEMA"),  -->
<!--   ) |>  -->
<!--   select( -->
<!--     row_ID, B_burgdorferi, weight, sex, taxonID, plotID, siteID, decimalLatitude, -->
<!--     decimalLongitude, elevation, larvalTicksAttached, recapture, fate,  -->
<!--     collectDate -->
<!--   ) %>% -->
<!--   mutate( -->
<!--     lat = scale(decimalLatitude), lon = scale(decimalLongitude),  -->
<!--     elev = scale(elevation), captureMonth = month(collectDate),  -->
<!--     year = year(collectDate) -->
<!--   ) %>% -->
<!--   filter(complete.cases(.)) -->

<!-- gfm <- glmer( -->
<!--   B_burgdorferi ~ weight + sex + taxonID + captureMonth + larvalTicksAttached + # lon + lat + -->
<!--     (1|plotID),  -->
<!--   family = "binomial", -->
<!--   data = model_data -->
<!-- ) -->
<!-- model_data$resid = residuals(gfm) -->
<!-- model_data$fitted = fitted(gfm) -->
<!-- model_data$resid_group = model_data$resid > 0  -->

<!-- summary(gfm) -->

<!-- hist(residuals(gfm)) -->

<!-- ggplot(model_data, aes(x = resid, y = fitted, col = resid_group)) + -->
<!--   geom_point() -->

<!-- tmp <- mammals |> left_join(model_data) -->

<!-- tmp |> filter(resid_group) -->

<!-- tmp |> filter(!resid_group) -->

<!-- tmp |>  -->
<!--   filter(!is.na(resid_group), !is.na(B_burgdorferi)) |>  -->
<!--   mutate(collectMonth = month(collectDate)) |>  -->
<!--   select( -->
<!--     resid_group, decimalLatitude, decimalLongitude, elevation, collectMonth,  -->
<!--     sex:fate -->
<!--   ) |>  -->
<!--   ggpairs() -->
<!-- ``` -->

