---
params: 
  force: FALSE
  cache_dir: "infection-modeling/R/rmarkdown/cache/resistance-tolerance"
---

# Immunity

```{r}
library(GGally)
library(tidyverse)
library(vegan)
```


```{r, include = FALSE}
# load some important functions for caching
source("infection-modeling/R/functions/cache_funs.R")
```

```{r}
# source("infection-modeling/R/scripts/clean_immune_data.R") # saves immune
immunity <- readRDS("infection-modeling/data/immune-data.rds")

ipca_list <- readRDS("infection-modeling/data/immune-PCA.rds")
```

In addition to the NEON

## Genetic correlations

First, let's look at how the genetic traits are correlated with each other. 
Figure \@ref(fig:genCorPlot) shows that the genetic traits are nearly all significantly
intercorrelated (the lone exception being TGF-B and IL-6, which are 
uncorrelated). Notably, the resistance traits (IL-10, IFN-y, IL-6, TLR-2) are
most strongly correlated with each other and, similarly, the tolerance traits
(TGF-B and GATA-3) are most strongly correlated with each each other. Both
TGF-B and GATA-3 appear to be zero inflated and all the genetic traits 
are highly right-skewed (i.e., many low values).

```{r genCorPlot, fig.width=7.5, fig.asp=0.95, warning=FALSE, message=FALSE, fig.cap=caption}
library(GGally)

plot_cache(
  immunity %>% 
    select(one_of(ipca_list$species_scores$gene)) %>% 
    ggpairs(progress = FALSE) + 
    scale_x_continuous(breaks = c(500, 1500)),
  plot_path = file.path(params$cache_dir, "res-tol-pairplot.png"), 
  force = params$force
)

caption = paste(
  "Correlations among genetic expression traits.",
  "IL-10, IFN-y, IL-6, and TLR-2 are thought to be resistance-associated genes",
  "and TGF-B and GATA-3 are throught to be tolerance-associated."
)
```

## PCA

The resistance and tolerance genetic traits also broadly group together
in a principal component analysis (PCA; Figure \@ref(fig:genPCPlot)), with "resistance" expression aligning with 
the first principal component (PC) and "tolerance" expression aligning with the
second PC. Overall, these first two PCs explain 87% of the variation among the
genetic expression varaibles (Table \@ref(tab:genPCImport)). It also appears
that *Borrelia* infection status is not strongly associated with either 
resistance or tolerance, but *Borrelia* burden may be associated with
tolerance expression (Figure Figure \@ref(fig:genPCPlot), blue lines).

```{r genPCPlot, fig.asp = 1, fig.cap=caption}
ordiplot(ipca_list$pca_obj) %>% 
  orditorp(display = "species")
  plot(ipca_list$envfit_obj, add = TRUE)

# TODO: recreate this figure with ggplot: 
## https://jkzorz.github.io/2020/04/04/NMDS-extras.html

caption = paste(
  "Principal component analysis of genetic expression traits.",
  "Traits were log(x+1) transformed prior to the analysis."
)
```

```{r genPCImport}
summary(ipca_list$pca_obj)$cont$importance %>% 
  knitr::kable(
    digits = 2, 
    caption = paste(
      "Importance of genetic expression principal components."
    )
  )
```

```{r genEnvFitSig}
vecs <- ipca_list$envfit_obj$vectors
facs <- ipca_list$envfit_obj$factors

vecs$arrows %>% 
  data.frame() %>% 
  bind_rows(
    facs$centroids %>% data.frame()
  ) %>% 
  rownames_to_column("variable") %>% 
  mutate(
    type = c("vector", rep("factor", 2)), 
    perms = c(vecs$permutations, rep(facs$permutations, 2)),
    rsquared = c(vecs$r, rep(facs$r, 2)),
    pvals = c(vecs$pvals, rep(facs$pvals, 2))
  ) %>% 
  knitr::kable(
    digits = 2,
    caption = paste(
      "Association of *Borrelia* burden and infection status",
      "with genetic expression (principal components)."
    )
  )
```

Multivariate regression of burden and infection against the first 
two genetic expression PCs implies that only infection status is 
significantly associated with the resistance-tolerance PCs (\@ref(tab:BbMvr)).
This approach is admittedly simplistic, since we aren't accounting for spatial
or temporal varation. These relationships will be assessed more vigorously
in later sections.

```{r BbMvr}
Bb_mvr <- immunity %>% 
  full_join(ipca_list$site_scores, by = "uid") %>% 
  select(Bb_infected, Bb_burden, expr_PC1, expr_PC2) %>% 
  filter(complete.cases(.)) %>% 
  lm(cbind(Bb_infected, Bb_burden) ~ expr_PC1 + expr_PC2, data = .)

mvr_smry <- summary(Bb_mvr)

inf_smry <- mvr_smry[["Response Bb_infected"]]

bur_smry <- mvr_smry[["Response Bb_burden"]]

bind_rows(
  inf_smry %>% coefficients() %>% data.frame() %>% 
    rownames_to_column("Predictor") %>% 
    mutate(response = "Bb_infected", rsqr = inf_smry$r.squared), 
  bur_smry %>% coefficients() %>% data.frame() %>%
    rownames_to_column("Predictor") %>% 
    mutate(response = "Bb_burden", rsqr = bur_smry$r.squared)
) %>% 
  relocate(response, before = .0) %>% 
  knitr::kable(
    col.names = c("Response", "Predictor", "Coef", "SE", "t", "P", "r-squared"),
    digits = 2,
    caption = paste(
      "Multivariate principal component regression (PCR) of *Borrelia*",
      "infection status and burden."
    )
  )
```

```{r}
## TODO: Possibly add in the info about transformations from the 
## resistance-tolerance.Rmd file.
```

