---
params: 
  force: FALSE
  cache_dir: "infection-modeling/R/rmarkdown/cache/resistance-tolerance"
---

```{r}
library(GGally)
library(tidyverse)
library(vegan)
```

# Immunity

```{r, include = FALSE}
# load some important functions for caching
source("infection-modeling/R/functions/cache_funs.R")
```

```{r}
# source("infection-modeling/R/scripts/clean_immune_data.R") # saves immune
immunity <- readRDS("infection-modeling/data/immune-data.rds")

nsamps <- nrow(immunity)

ipca_list <- readRDS("infection-modeling/data/immune-PCA.rds")
```

A key component of this project is the resistance to and tolerance of 
*Borrelia* infection in *Peromyscus* mice. As such, we obtained from NEON ear 
tissue samples from `r prettyNum(nsamps, ",")` individuals across three seasons
of their collection efforts (2022 - 2024). From these samples, we measured the
expression of six important genes thought to be associated with resistance
and tolerance (Table \@ref(tab:geneLookupTab)). We also quantified *Borrelia*
infection and burden, using Droplet Digital Polymerase Chain Reaction (ddPCR) 
techniques. The ddPCR variant is more sensitive than the traditional PCR that 
NEON uses to determine infection status^[**Citation needed**]. 
Gene expression and *Borrelia* infection information were collected at the 
University of South Florida (USF). Due to limitations in the amount of tissue
available in 2022, ddPCR was not conducted in this year. Additional tissue from
each individual was obtained in 2023 and 2024 to facilitate this. Because
the expression (and burden) data are highly right-skewed (many low-values), 
we present and analyze $log(x + 1)$ transforms of these variables throughout
this project.

<br> 
```{r geneLookupTab}
gene_lookup <- readRDS("infection-modeling/data/expression-lookup.rds")

gene_lookup %>% 
  knitr::kable(
    caption = paste(
      "Resistance and tolerance genes, the triplex batch on which they were",
      "analyzed, and their expected immune function."
    )
  )
```

## Genetic correlations

First, we look at how the expression of these genes traits intercorrelated 
among these individuals. 
Figure \@ref(fig:genCorPlot) shows that the genetic traits are nearly all 
significantly intercorrelated (the lone exception being TGF-B and IL-6, which 
are uncorrelated). Notably, the resistance traits (IL-10, IFN-y, IL-6, TLR-2) 
are most strongly correlated with each other and, similarly, the tolerance 
traits (TGF-B and GATA-3) are most strongly correlated with each each other. 

```{r genCorPlot, fig.width=7.5, fig.asp=0.95, warning=FALSE, message=FALSE, fig.cap=caption}
library(GGally)

plot_cache(
  immunity %>% 
    select(one_of(ipca_list$species_scores$gene)) %>% 
    filter(complete.cases(.)) %>% 
    ggpairs(progress = FALSE),
  plot_path = file.path(params$cache_dir, "res-tol-pairplot.png"), 
  force = params$force
)

caption = paste(
  "Correlations among log-transformed genetic expression traits.",
  "IL-10, IFN-y, IL-6, and TLR-2 are thought to be resistance-associated genes",
  "and TGF-B and GATA-3 are throught to be tolerance-associated."
)
```

## PCA

The resistance and tolerance genetic traits also broadly group together
in a principal component analysis (PCA; Figure \@ref(fig:genPCPlot)), with 
"resistance" expression aligning with 
the first principal component (PC) and "tolerance" expression aligning with the
second PC. Overall, these first two PCs explain 87% of the variation among the
genetic expression varaibles (Table \@ref(tab:genPCImport)). It also appears
that *Borrelia* infection status is not strongly associated with either 
resistance or tolerance, but *Borrelia* burden may be associated with
tolerance expression (Figure Figure \@ref(fig:genPCPlot), blue lines).

```{r genPCPlot, fig.asp = 1, fig.cap=caption}
plot(ipca_list$pca_obj, type = "n") %>%
  points("sites", col = "grey80", pch = 20) %>% 
  points("species", col = "red", pch = "+", cex = 1.2) %>% 
  orditorp(display = "species", font = 2, cex = 0.8)
plot(ipca_list$envfit_obj, add = TRUE, col = "blue")

# TODO: recreate this figure with ggplot: 
## https://jkzorz.github.io/2020/04/04/NMDS-extras.html

caption = paste(
  "Principal component analysis of genetic expression traits.",
  "Traits were log(x+1) transformed prior to the analysis."
)
```

<br>
```{r genPCImport}
summary(ipca_list$pca_obj)$cont$importance %>% 
  knitr::kable(
    digits = 2, 
    caption = paste(
      "Importance of genetic expression principal components."
    )
  )
```

<br>
```{r genEnvFitSig}
vecs <- ipca_list$envfit_obj$vectors
facs <- ipca_list$envfit_obj$factors

vecs$arrows %>% 
  data.frame() %>% 
  bind_rows(
    facs$centroids %>% data.frame()
  ) %>% 
  rownames_to_column("variable") %>% 
  mutate(
    type = c("vector", rep("factor", 2)), 
    perms = c(vecs$permutations, rep(facs$permutations, 2)),
    rsquared = c(vecs$r, rep(facs$r, 2)),
    pvals = c(vecs$pvals, rep(facs$pvals, 2))
  ) %>% 
  knitr::kable(
    digits = 2,
    caption = paste(
      "Association of *Borrelia* burden and infection status",
      "with genetic expression (principal components)."
    )
  )
```
<br>

Multivariate regression of burden and infection against the first 
two genetic expression PCs implies that only the "tolerance" genes (PC2) are  
significantly associated with the *Borrelia* infection or burden
(\@ref(tab:BbMvr)). This approach is admittedly simplistic, since we aren't 
accounting for spatial or temporal variation. These relationships will be 
assessed more vigorously in later sections.

<br>
```{r BbMvr}
Bb_mvr <- immunity %>% 
  full_join(ipca_list$site_scores, by = "uid") %>% 
  select(Bb_infected, log_burden, expr_PC1, expr_PC2) %>% 
  filter(complete.cases(.)) %>% 
  lm(cbind(Bb_infected, log_burden) ~ expr_PC1 + expr_PC2, data = .)

mvr_smry <- summary(Bb_mvr)

inf_smry <- mvr_smry[["Response Bb_infected"]]

bur_smry <- mvr_smry[["Response log_burden"]]

bind_rows(
  inf_smry %>% coefficients() %>% data.frame() %>% 
    rownames_to_column("Predictor") %>% 
    mutate(response = "Bb_infected", rsqr = inf_smry$r.squared), 
  bur_smry %>% coefficients() %>% data.frame() %>%
    rownames_to_column("Predictor") %>% 
    mutate(response = "log_burden", rsqr = bur_smry$r.squared)
) %>% 
  relocate(response, before = .0) %>% 
  knitr::kable(
    col.names = c("Response", "Predictor", "Coef", "SE", "t", "P", "r-squared"),
    digits = 2,
    caption = paste(
      "Multivariate principal component regression (PCR) of *Borrelia*",
      "infection status and burden."
    )
  )
```

```{r}
## TODO: Possibly add in the info about transformations from the 
## resistance-tolerance.Rmd file.
```

```{r}
Peros <- readRDS("infection-modeling/data/peromyscus-model-data.rds")
```

## Infection sample sizes

This project measured gene expression, but not *Borrelia* in 2022.
NEON did measure infection during that year, but we can only compare
*Borrelia*-positive samples across methods (we can be confident that a 
"positive" is correct, but a "negative" is not certain with NEON's PCR 
technique). Adding these samples would only net us 5 total samples 
(\@ref(tab:infectMethodSamps)), which is likely not worth the cost of the new
"method" variable that would be required to differentiate between the two
infection screening methodologies.

```{r infectMethodSamps}
methodSamps <- Peros %>% 
  select(
    year, siteID, plotID, expr_PC1, expr_PC2, Bb_infected, NEON_Bb_infected,
    cap_prop_night
  ) %>% 
  filter(year %in% c(2020:2024)) %>% 
  group_by(year, siteID, plotID) %>% 
  summarize(
    expression = sum(!is.na(expr_PC1) | !is.na(expr_PC2), na.rm = TRUE),
    captime = sum(!is.na(cap_prop_night), na.rm = TRUE),
    USF = sum(!is.na(Bb_infected), na.rm = TRUE),
    NEON = sum(!is.na(NEON_Bb_infected) & 
                        NEON_Bb_infected == 1, na.rm = TRUE),
    without_NEON = sum(
      (!is.na(expr_PC1)|!is.na(expr_PC2)) &
        !is.na(cap_prop_night) &
        (!is.na(Bb_infected)),
      na.rm = TRUE
    ),
    with_NEON = sum(
      (!is.na(expr_PC1)|!is.na(expr_PC2)) &
        !is.na(cap_prop_night) &
        (!is.na(Bb_infected) | 
           (!is.na(NEON_Bb_infected) & NEON_Bb_infected == 1)),
      na.rm = TRUE
    ), .groups = "drop"
  ) %>% 
  mutate(added = with_NEON - without_NEON) %>% 
  filter(expression > 0) %>% 
  arrange(year, siteID, plotID, expression)

methodSamps %>% 
  select(-c(siteID, without_NEON, with_NEON)) %>% 
  knitr::kable(
    digits = 2,
    caption = paste(
      "Tabulation of expression, behavior, and infection samples across NEON",
      "plots. Columns correspond to the year, sand plot by which counts",
      "are grouped; the number of samples for which we have expression,",
      "capture timing, and infection data (either from USF or NEON);",
      "and the samples gained by including NEON infection data."
    ),
    format = "pipe"
  )
```