---
params: 
  force: FALSE
  cache_dir: "infection-modeling/R/rmarkdown/cache/resistance-tolerance"
---

```{r}
library(tidyverse)
library(vegan)
```

# Climate and Weather

One important component of our ultimate models will be measures of short-term and 
long-term climate conditions. For short-term variables, we consider
the minimum, mean, and maximum temperature, total precipitation, and mean
relative humidity on the day the individual was captured as well as the
cumulative daily precipitation for the year up to the capture date. Long term
variables include the average monthly (mean) temperature, average monthly
precipitation, and the average monthly rain days (as a proportion) across 7
years (2017 - 2024) at a site. Additionally, coefficients of variation are 
considered in the long-term variables.

To simplify these variables, we conducted PCA for both short-term (weather) 
and long-term (climate) data. Positive values of the first weather PC are 
associated with warmer and wetter observations (Fig \@ref(fig:wthrPCA)) and the
first two PCs account for 73% of the variation in the weather data 
(\@ref(tab:wthrPCimport)).

```{r}
wpca_list <- readRDS("infection-modeling/data/weather-PCA.rds")
```

```{r wthrPCA, fig.asp = 1, fig.cap=caption}
ordiplot(wpca_list$pca_obj) %>% 
  orditorp(display = "species")

caption = paste(
  "Principal component ordination of weather on the day of an",
  "individual capture."
)
```

```{r wthrPCimport}
summary(wpca_list$pca_obj)$cont$importance %>% 
  knitr::kable(
    digits = 2,
    caption = paste(
      "Importance of weather principal components."
    )
  )
```

Positive values of the first climate PC are associated with cooler, dryer, 
and more variable sites (Fig \@ref(fig:climPCA)) and the first two PCs
explain 94% of the climatic variation among sites (\@ref(tab:climPCimport)).


```{r}
# climate pca list
cpca_list <- readRDS("infection-modeling/data/climate-PCA.rds")
```

```{r climPCA, fig.asp = 1, fig.cap=caption}
# rough plot of the PCA
ordiplot(cpca_list$pca_obj) %>%
  orditorp(display = "species")

caption = paste(
  "Principal component ordination of 7-year climate at a NEON site."
)
```

```{r climPCimport}
summary(cpca_list$pca_obj)$cont$importance %>% 
  knitr::kable(
    digits = 2,
    caption = "Importance of climate principal components."
  )
```

