---
params: 
  force: FALSE
  cache_dir: "infection-modeling/R/rmarkdown/cache/resistance-tolerance"
---

```{r}
library(tidyverse)
library(vegan)
```

# Climate and Weather

Another important component of our models will be measures of short-term and 
long-term weather conditions. For short-term variables (weather), we consider
the minimum, mean, and maximum temperature, total precipitation, and mean
relative humidity on the day an individual was captured as well as the
cumulative daily precipitation for the year up to the capture date. Long-term
variables (climate) include the average monthly (mean) temperature, average 
monthly precipitation, and the average monthly rain days (as a proportion) 
across 7 years (2017 - 2024) at a site. Additionally, coefficients of 
variation are for the climate variables were also considered.

To reduce complexity in our weather and climate information, we conducted 
PCA for both short-term (weather) and long-term (climate) data. Positive 
values of the first weather PC are associated with warmer and wetter
observations (Fig \@ref(fig:wthrPCA)) and the first two PCs account for 
73% of the variation in the weather data (Table \@ref(tab:wthrPCimport)). 
Positive values of the first climate PC are associated with cooler, dryer, 
and more variable sites (Fig \@ref(fig:climPCA)) and the first two PCs
explain 94% of the climatic variation among sites 
(Table \@ref(tab:climPCimport)).

```{r}
wpca_list <- readRDS("infection-modeling/data/weather-PCA.rds")
```

```{r wthrPCA, fig.asp = 1, fig.cap=caption}
plot(wpca_list$pca_obj, type = "n") %>%
  points("sites", col = "grey80", pch = 20) %>% 
  points("species", col = "red", pch = "+", cex = 1.2) %>% 
  orditorp(display = "species", font = 2, cex = 0.66)

caption = paste(
  "Principal component ordination of weather on the day of an",
  "individual capture."
)
```

<br>
```{r wthrPCimport}
summary(wpca_list$pca_obj)$cont$importance %>% 
  knitr::kable(
    digits = 2,
    caption = paste(
      "Importance of weather principal components."
    )
  )
```

```{r}
# climate pca list
cpca_list <- readRDS("infection-modeling/data/climate-PCA.rds")
```

```{r climPCA, fig.asp = 1, fig.cap=caption}
# rough plot of the PCA
plot(cpca_list$pca_obj, type = "n") %>%
  points("sites", col = "grey80", pch = 20, cex = 2) %>% 
  points("species", col = "red", pch = "+", cex = 1.2) %>% 
  orditorp(display = "species")

caption = paste(
  "Principal component ordination of 7-year climate at a NEON site."
)
```
<br> 

```{r climPCimport}
summary(cpca_list$pca_obj)$cont$importance %>% 
  knitr::kable(
    digits = 2,
    caption = "Importance of climate principal components."
  )
```

