```{r}
library(tidyverse)
```

# Behavior

Individual behavior of an individual is expected to be associated with disease 
dynamics in a number of ways. The timing and location of movements can 
influences exposure to pathogens and how an individual chooses to expend 
energy can influence its ability to fight off infection. We chose to assess 4 
metrics of behavior for this project: 1) Activity timing, 2) range size,
3) exploration, and 4) learning

## Activity timing

From 2022 - 2024, our group added ibutton temperature sensor pairs to all traps 
at one of each plots within the eight NEON sites of interest. These sensors,
which log temperatures every five minutes, allowed us to estimate the time
at which an individual entered a 
trap^[**Citation to original Orrock paper needed**] and, thus, the time at
which the individual is active. This technique works because when an individual 
is inside a trap, the air inside increases above the ambient air. 
Figure \@ref(fig:ibutCapTimeExamp) shows an example of these data. For most
analyses, capture time is presented as the fraction of the night 
(sunset to sunrise) that had elapsed when the capture occurred. This 
relative capture time scaling allows for more equitable comparisons 
across latitudinal gradients and photoperiods 
(Figure \@ref(fig:capTimeByLat)).

Figure \@ref(fig:capTimeOverlap) shows the distribution of those relative
capture times for *Peromyscus*, how they differ across sites and through time.

```{r ibutCapTimeExamp, fig.cap=caption}
knitr::include_graphics(
  path = R.utils::getAbsolutePath(
    paste0(
      "activity-timing/graphics/trapping-bout_temperature-data/",
      "capture-paired-temperatures_UNDE_2.jpg"
    )
  ), rel_path = FALSE, error = FALSE
)

caption = paste(
  "Capture time estimate example: ibutton data (green and blue lines)",
  "from a three-day trapping event at plot #2 from the UNDE NEON site in 2024.",
  "The dotted vertical lines correspond to the algorithmically-derived estimate",
  "of the time an individual entered the trap. Purple rectangles correspond to",
  "the segment of time during which the temperature difference between",
  "inner and outer buttons is above the trigger threshold (1°C).",
  "No data are shown for trapnights in which there was no capture."
)
```
```{r capTimeByLat, out.width="35%", fig.cap=caption}
knitr::include_graphics(
  path = R.utils::getAbsolutePath(
    paste0(
      "activity-timing/graphics/capture-time-exploration/",
      "captime-by-latitude.jpg"
    )
  ), rel_path = FALSE, error = FALSE
)

caption = paste(
  "Linear relationship between relative capture time (y) and latitude (x)."
)
```

```{r capTimeOverlap, fig.cap=caption}
knitr::include_graphics(
  path = R.utils::getAbsolutePath(
    paste0(
      "activity-timing/graphics/capture-time-exploration/",
      "capture-overlap_site-by-year.jpg"
    )
  ), rel_path = FALSE, error = FALSE
)

caption = paste(
  "*Peromyscus* relative capture time density distributions across eight NEON",
  "sites over three years."
)
```

## Range Size

## Exploration

## Learning

## Recapture dynamics

* trap diversity

* average trap distances

* "trapability" 

* recapture timing shifts (extra)

## Tick contact