```{r}
library(tidyverse)

Peros <- readRDS("infection-modeling/data/peromyscus-data.rds")
behavior <- readRDS("infection-modeling/data/individual-behavior-data.rds")

Peros <- left_join(Peros, behavior, by = "iid")
```

# Behavior

Behavior of individual mice is expected to be associated with disease 
dynamics in a number of ways. The timing and location of movements can 
influences exposure to pathogens and how an individual chooses to expend 
energy can influence its ability to fight off infection. We chose to assess 4 
metrics of behavior for this project: 1) Activity timing, 2) range size,
3) exploration, and 4) learning. Since this is a field study and we weren't
able to follow each individual, we estimated proxies for these behavioral 
metrics

```{r}
recaps <- Peros %>% filter(!is.na(iid), year %in% c(2022:2024)) %>% 
  group_by(iid) %>% tally()
ninds <- nrow(recaps)
nrecaps <- recaps %>% filter(n > 1) %>% nrow()
recap_pct = round((nrecaps / ninds) * 100, 1)
avg_caps <- recaps %>% ungroup() %>% filter(n > 1) %>% 
  pull(n) %>% mean() %>% round(1)
```

Many of these behavioral metrics rely on recaptures. From 2022 to 2024, there
were `r prettyNum(nrecaps, ",")` *Peromyscus* individuals that 
were recaptured at least once (`r prettyNum(recap_pct)`% of total individuals).
On average, individuals that were captured more than once were captured 
`r prettyNum(avg_caps)` times on average.

## Activity timing

Our indicator for activity timing in this study is capture time.
From 2022 - 2024, our group added ibutton temperature sensor pairs to all traps 
at one of each plots within the eight NEON sites of interest. These sensors,
which log temperatures every five minutes, allowed us to estimate the time
at which an individual entered a 
trap^[**Citation to original Orrock paper needed**] and, thus, the time at
which the individual is active. 

This technique of deriving capture timing works because 
the air inside increases above the ambient air when an animal is in it. 
Figure \@ref(fig:ibutCapTimeExamp) shows an example of these data. For most
analyses, capture time is presented as the fraction of the night 
(sunset to sunrise) that had elapsed when the capture occurred. This 
relative capture time scaling allows for more equitable comparisons 
across latitudinal gradients and photoperiods 
(Figure \@ref(fig:capTimeByLat)).

Figure \@ref(fig:capTimeOverlap) shows the distribution of those relative
capture times for *Peromyscus*, how they differ across sites and through time.

```{r ibutCapTimeExamp, fig.cap=caption}
knitr::include_graphics(
  path = R.utils::getAbsolutePath(
    paste0(
      "activity-timing/graphics/trapping-bout_temperature-data/",
      "capture-paired-temperatures_UNDE_2.jpg"
    )
  ), rel_path = FALSE, error = FALSE
)

caption = paste(
  "Capture time estimate example: ibutton data (green and blue lines)",
  "from a three-day trapping event at plot #2 from the UNDE NEON site in 2024.",
  "The dotted vertical lines correspond to the algorithmically-derived",
  "estimate",
  "of the time an individual entered the trap. Purple rectangles correspond to",
  "the segment of time during which the temperature difference between",
  "inner and outer buttons is above the trigger threshold (1°C).",
  "No data are shown for trapnights in which there was no capture."
)
```
```{r capTimeByLat, out.width="25%", fig.cap=caption}
knitr::include_graphics(
  path = R.utils::getAbsolutePath(
    paste0(
      "activity-timing/graphics/capture-time-exploration/",
      "captime-by-latitude.jpg"
    )
  ), rel_path = FALSE, error = FALSE
)

caption = paste(
  "Linear relationship between relative capture time (y) and latitude (x)."
)
```

```{r capTimeOverlap, fig.cap=caption}
knitr::include_graphics(
  path = R.utils::getAbsolutePath(
    paste0(
      "activity-timing/graphics/capture-time-exploration/",
      "capture-overlap_site-by-year.jpg"
    )
  ), rel_path = FALSE, error = FALSE
)

caption = paste(
  "*Peromyscus* relative capture time density distributions across eight NEON",
  "sites over three years."
)
```

## Range size

Our indicator of range size is the average distance between traps that an 
individual was captured between each consecutive capture. As it depends upon 
recaptures, this metric is biased towards individuals that have been captured
more times times (Figure \@ref(fig:MoveDistHist)); individuals captured 
only once always have an average trap distance of 0. But for those 
individuals that were recaptured, it provides some information about the 
size of the space that the individual occupies. Figure 
\@ref(fig:recapMoveDistro) shows the distribution of movement distances among
*Peromyscus* across NEON sites.

```{r MoveDistHist, out.width="30%", fig.cap=caption, fig.asp = 1, fig.width=3}
Peros %>% 
  filter(year %in% c(2022:2024)) %>% 
  select(iid, avg_move_dist, n_captures) %>% distinct() %>% 
  ggplot(aes(x = log(avg_move_dist + 1), fill = n_captures > 1)) + 
  geom_histogram(position = "identity", bins = 30, col = "black") +
  labs(y = "Individuals", x = "ln(Trap distance + 1)") + 
  scale_fill_manual(
    name = "Recaptured?",
    values = c("cornflowerblue", "orange"),
    labels = c("no", "yes")
  ) +
  theme(
    legend.position = "inside", legend.position.inside = c(1, 1),
    legend.justification = c(1, 1), legend.background = element_blank()
  )
  
caption = paste(
  "Distribution of the log-transformed distance (meters) between",
  "consecutive captures, averaged across all recaptures, for an individual",
  "*Peromyscus*."
)
```

```{r recapMoveDistro, fig.asp=0.8, fig.width=5, fig.cap=caption, out.width="50%"}
Peros %>% 
  filter(year %in% c(2022:2024), n_captures > 1, sex_male != 0.5) %>% 
  select(siteID, iid, sex_male, sex_mature, avg_move_dist) %>% 
  distinct() %>% 
  ggplot(aes(x = log(avg_move_dist + 1), fill = factor(sex_male))) +
  facet_wrap(~siteID) + 
  geom_density(alpha = 0.5) +
  labs(x = "ln(Trap disance + 1)", y = "Density", fill = "Sex") +
  scale_fill_discrete(
    labels = c("F", "M")
  ) +
  theme(
    legend.position = "inside", legend.position.inside = c(0.85, 0),
    legend.justification = c(0.5, 0), legend.background = element_blank(),
    panel.spacing = unit(0.1, "lines")
  )

caption = paste(
  "Distribution of log-transformed distance (D, meters) between",
  "consecutive captures of recaptured individuals, seperated by site and sex."
)
```

## Exploration

Our proxy for exploration is trap diversity: the number of unique traps an
individual was captured in. We scaled trap diversity by the total number of 
captures (relative to the population) to alleviate some of the bias towards
recaptures 
(Figures \@ref(fig:trapDivByCaps), \@ref(fig:wTrapDivTrapDiv)). 
While this (weighted) trap diversity is somewhat associated with trap 
distance, they are not equivalent (Figure \@ref(fig:trapDistTrapDIv)). 
An individual that travels far along a consistent path is probably functionally 
different from one that stays within a relatively small range, but uses more 
of that range. The former individual would have a higher trap distance and a 
lower trap diversity than the latter.

```{r trapDivByCaps, fig.asp = 1, fig.width = 3, out.width="25%", fig.cap = caption}
behavior %>% 
  ggplot(aes(x = n_captures, y = unique_traps)) + 
  geom_point() + geom_smooth(method = "lm", formula = y~x) +
  labs(x = "Captures", y = "Trap diversity")

caption = paste(
  "Relationship between individual capture count (x) and number of unique",
  "traps captured in (y)."
)
```

```{r wTrapDivTrapDiv, fig.asp = 1, fig.width = 3, out.width="25%", fig.cap=caption}
behavior %>% 
  ggplot(aes(x = unique_traps, y = weighted_trap_diversity)) +
  geom_point() + 
  geom_smooth(method = "lm", formula = y~poly(x, 2)) +
  labs(x = "Trap diversity (raw)", y = "Weighted trap diversity")

caption = paste(
  "Relationship between raw (x) and weighted (y) trap diversity.",
  "The regression line is a second-order polynomial."
)
```

```{r trapDistTrapDIv, fig.asp = 0.8, fig.width = 5, out.width="50%", fig.cap=caption}
Peros %>% 
  group_by(iid) %>% 
  mutate(plots_visited = length(unique(plotID))) %>% ungroup() %>% 
  filter(year %in% c(2022:2024)) %>% 
  select(
    iid, avg_move_dist, n_captures, weighted_trap_diversity, unique_traps,
    plots_visited
  ) %>% 
  distinct() %>% 
  ggplot(
    aes(x = log(avg_move_dist + 1), y = (weighted_trap_diversity + 1))
  ) +
  geom_point(aes(col = factor(plots_visited))) + 
  geom_smooth(method = "lm") +
  labs(
    x = "ln(Trap distance + 1)", y = "Weighted trap diversity",
    col = "Plots\nvisited"
  ) + theme(
    legend.position = "inside", legend.position.inside = c(1, 1), 
    legend.justification = c(1, 1), legend.background = element_blank()
  )

caption = paste(
  "Relationship between (log-transformed) trap distance and weighted trap",
  "diversity. Red points represent individuals that were only caught at",
  "at one plot and blue dot individuals were captured at 2 plots."
)
```

## Learning

## Recapture dynamics

* trap diversity

* average trap distances

* "trapability" 

* recapture timing shifts (extra)


### Outliers?

```{r}
most_trapped <- behavior %>% 
  left_join(
    Peros %>%
      group_by(iid) %>% 
      mutate(
        nplots = length(unique(plotID)), 
        # avg_captime = mean(cap_prop_night, na.rm = TRUE)
      ) %>% 
      select(iid, sex_male, new_taxonID, nplots) %>% distinct(),
    by = "iid"
  ) %>% 
  # arrange(desc(unique_traps)) %>% 
  arrange(desc(n_captures)) %>% 
  slice(1:10) %>% 
  mutate(
    sex = factor(sex_male, levels = c(0, 0.5, 1), labels = c("F", "U", "M"))
  ) %>% 
  # filter(iid %in% outl_trapdiv$iid) %>% 
  select(
  iid, sex, n_captures, unique_traps, nplots,
  residence_trapnights, known_alive_days,
  weighted_trapability, avg_move_dist, 
  max_travel_dist
  )
  
avg_traps <- most_trapped$unique_traps %>% mean(na.rm = TRUE)
max_traps <- most_trapped$unique_traps %>% max(na.rm = TRUE)
corresp_caps <- most_trapped$n_captures[which.max(most_trapped$unique_traps)]
corresp_days <- most_trapped$known_alive_days[
  which.max(most_trapped$unique_traps)
]
corresp_yrs <- corresp_days/365
avg_dist <- most_trapped$avg_move_dist %>% mean(na.rm = TRUE)
max_dist <- most_trapped$max_travel_dist %>% max(na.rm = TRUE)

big_maxd_km <- max(Peros$max_travel_dist)/1000
```

A handful of mice were captured in many traps 
(Table \@ref(tab:mostTrapped)) at the eight sites and across all of NEON's 
duration. One individual was trapped `r prettyNum(corresp_caps)` times in 
`r prettyNum(max_traps)` traps over 
`r prettyNum(corresp_yrs, digits = 1)` years! On average, these trap-happy
individuals stuck
to relatively small areas (average of `r prettyNum(avg_dist)` meters), with
a maximum distance of `r prettyNum(max_dist, digits = 1)` meters and 
`r prettyNum(avg_traps, digits = 1)` traps. Compare that to the very 
furthest distance between observations of an individual in the larger
*Peromyscus* population of `r prettyNum(big_maxd_km, digits = 1)` 
kilometers^[this is not unusual for a *Peromyscus* dispersal event 
(**Citation needed**)]! 

We evaluated the validity of these observations to the best of our abilities
and believe that the records for these individuals are corrected and not,
as we initially suspected, erroneous grouping of multiple individuals that 
share the same tag number^[NEON's `tagID` values are not entirely unique to 
individuals. Tags must be combined with site and recapture information to
differentiate individual animals (i.e., `iid`).]

<br>
```{r mostTrapped}
most_trapped %>% 
  knitr::kable(
    digits = 2,
    col.names = c(
      "iid", "sex", "caps", "traps", "plots", "trap nights", "days",
      "trapability", "avg D", "max D"
    ),
    caption = paste(
      "Most trapped *Peromyscus*: top ten most captured",
      "individuals. Columns represent individual ID, sex,",
      "number of times captured, number of unique traps captured in",
      "(i.e., trap diversity), number of unique plots caught on,",
      "number of trap nights between first and last capture",
      "(across all plots visited),", 
      "calendar days between first and last capture,",
      "trapability (weighted by times captured),", 
      "average distance (meters) between consecutive captures,", 
      "and maximum distance (meters) among observations of the individual."
    )
  )
```


## Tick contact