---
params: 
  force: FALSE
  cache_dir: "infection-modeling/R/rmarkdown/cache/resistance-tolerance"
---

# Infection and gene expression

```{r}
library(tidyverse)
```

```{r}
Peros <- readRDS("infection-modeling/data/peromyscus-model-data.rds")
```

## Infection

Our colleagues at USF did measured gene expression, but not *Borrelia* in 2022.
NEON did measure infection during that year, but we can only compare
*Borrelia*-positive samples across methods (we can be confident that a 
"positive" is correct, but a "negative" is not certain with NEON's PCR 
technique). Adding these samples would only net us 5 total samples 
(\@ref(tab:infectMethodSamps)), which is likely not worth the cost of the new
"method" variable that would be required to differentiate between the two
infection screening methodologies.

```{r infectMethodSamps}
methodSamps <- Peros %>% 
  select(
    year, siteID, plotID, expr_PC1, expr_PC2, Bb_infected, NEON_Bb_infected,
    cap_prop_night
  ) %>% 
  filter(year %in% c(2020:2024)) %>% 
  group_by(year, siteID, plotID) %>% 
  summarize(
    expression = sum(!is.na(expr_PC1) | !is.na(expr_PC2), na.rm = TRUE),
    captime = sum(!is.na(cap_prop_night), na.rm = TRUE),
    USF = sum(!is.na(Bb_infected), na.rm = TRUE),
    NEON = sum(!is.na(NEON_Bb_infected) & 
                        NEON_Bb_infected == 1, na.rm = TRUE),
    without_NEON = sum(
      (!is.na(expr_PC1)|!is.na(expr_PC2)) &
        !is.na(cap_prop_night) &
        (!is.na(Bb_infected)),
      na.rm = TRUE
    ),
    with_NEON = sum(
      (!is.na(expr_PC1)|!is.na(expr_PC2)) &
        !is.na(cap_prop_night) &
        (!is.na(Bb_infected) | 
           (!is.na(NEON_Bb_infected) & NEON_Bb_infected == 1)),
      na.rm = TRUE
    ), .groups = "drop"
  ) %>% 
  mutate(added = with_NEON - without_NEON) %>% 
  filter(expression > 0) %>% 
  arrange(year, siteID, plotID, expression)

methodSamps %>% 
  select(-c(siteID, without_NEON, with_NEON)) %>% 
  knitr::kable(
    digits = 2,
    caption = paste(
      "Tabulation of expression, behavior, and infection samples across NEON",
      "plots. Columns correspond to the year, sand plot by which counts",
      "are grouped; the number of samples for which we have expression,",
      "capture timing, and infection data (either from USF or NEON);",
      "and the samples gained by including NEON infection data."
    ),
    format = "pipe"
  )

# methodSamps %>% 
  # write.csv(
  #   "infection-modeling/data/added-samples-with-NEON-infections.csv",
  #   row.names = FALSE
  # )
#   
# methodSamps %>% 
#   select(without_NEON:added) %>% 
#   colSums()
```




## Resistance-Tolerance

```{r exprOrthog, fig.cap = caption, fig.asp = 0.8, fig.width = 6}
Peros %>% 
  select(expr_PC1, expr_PC2, siteID, year) %>% 
  filter(complete.cases(.)) %>% 
  ggplot(aes(x = expr_PC1, y = expr_PC2)) + 
  geom_hline(yintercept = 0, linetype = "dotted") + 
  geom_point(aes(col = , col = siteID, shape = factor(year))) + 
  geom_smooth(method = "lm", formula = y~x) + 
  labs(
    x = "expr_PC1 (resistance)", y = "expr_PC2 (tolerance)", col = "Site",
    shape = "Year"
  ) # +
  # theme(
  #   legend.position = "inside", legend.position.inside = c(1, 0.5),
  #   legend.justification = c(1, 0.5), legend.background = element_blank()
  # )

caption = paste(
  "Gene expression principal component values of individual *Peromyscus*.",
  "The trend line demonstrates the orthogonality of these axes."
)
```

```{r exprBurd, fig.cap = caption, fig.asp = 0.8, fig.width = 6}
Peros %>% 
  select(expr_PC1, expr_PC2, Bb_burden) %>% 
  filter(complete.cases(.)) %>% 
  # cor()
  ggplot(aes(x = log(Bb_burden + 1))) + 
  geom_point(aes(y = expr_PC1, col = "resistance"), alpha = 0.5) + 
  geom_smooth(aes(y = expr_PC1, col = "resistance"), method = "lm") + 
  geom_point(aes(y = expr_PC2, col = "tolerance"), alpha = 0.5) + 
  geom_smooth(aes(y = expr_PC2, col = "tolerance"), method = "lm")
  # ggplot(aes(x = expr_PC1, y = expr_PC2, col = log(Bb_burden + 1))) + 
  # geom_point(alpha = 0.5, size = 3) +
  # labs(
  #   x = "expr_PC1 (resistance)", y = "expr_PC2 (tolerance)", col = "ln(Burden)",
  # ) +
  # theme(
  #   legend.position = "inside", legend.position.inside = c(1, 0.5),
  #   legend.justification = c(1, 0.5), legend.background = element_blank()
  # ) + 
  # scale_color_gradient(low = "cornflowerblue", high = "red")

caption = paste(
  ""
)
```

