---
params: 
  force: FALSE
  cache_dir: "infection-modeling/R/rmarkdown/cache/resistance-tolerance"
---

# Complete data SEM

```{r}
library(tidyverse)
library(DiagrammeR)
```

Figure \@ref(fig:smallSelectSEM) visualizes a SEM fit to a "complete" data set
(i.e., only observations for which *all* variables were measured). As such, 
it has some limitations in its power, but it is the most "defensible" 
explanatory model because it does not draw conclusions beyond the scope of its
data. Models using more of the data will be explored in the next section.

While this model is very interesting, one of the primary paths of interest
is not significant (behavior -> ticks -> resistance -> burden -> tolerance). It is possible
that the "behavior -> ticks" path is improved by a model which is not limited
by the completeness restriction. Table \@ref(tab:smallSelSEMr2) summarizes the
fit (r^2) of each of the component models.

```{r smallSelectSEM, out.width="100%", fig.asp = 1, fig.cap=caption}
mod <- readRDS(
  "infection-modeling/data/model-objects/selected-alt-SEM_complete-cases.rds"
)

pbreaks = c(0, 0.05, 0.1, 1)#, 
pcols = c("black", "grey70", "grey70")#, 
linetypes = c("solid", "solid", "dotted")#, 
penwidth_range = c(1, 8)

# coefficients
coef_tab <- mod$coefficients %>% 
  data.frame() %>% 
  rename(sig.star = "Var.9") %>% 
  mutate(
    pval_group = cut(
      P.Value, breaks = pbreaks, include.lowest = TRUE,
      labels(seq_len(length(pbreaks) - 1))
    ),
    pval_col = pcols[pval_group],
    pval_linetype = linetypes[pval_group],
    pval_penwidth = scales::rescale(Std.Estimate, to = penwidth_range)
  )

# all variables
the_vars <- unique(c(coef_tab$Response, coef_tab$Predictor))

path_nodes <- data.frame(var = the_vars, id = seq_len(length(the_vars)))

create_graph(attr_theme = "bt") %>% 
  add_global_graph_attrs("layout", "dot", "graph") %>%
  # add_global_graph_attrs("rankdir", "LR", "graph") %>% 
  add_nodes_from_table(path_nodes, label_col = var) %>% 
  add_edges_from_table(
    coef_tab, from_col = "Predictor", to_col = "Response", 
    from_to_map = "label"
  ) %>% 
  set_node_attrs(width, 1.5) %>% 
  set_node_attrs(fontsize, 20) %>% 
  set_edge_attrs(color, coef_tab$pval_col) %>% 
  set_edge_attrs(style, coef_tab$pval_linetype) %>% 
  set_edge_attrs(penwidth, coef_tab$pval_penwidth) %>% 
  render_graph()


caption = paste(
  "Acyclycal graph diagram of selected SEM model, fit with complete cases.",
  "The goodness-of-fit P-value is", paste0(round(mod$Cstat$P.Value, 2), ","),
  "which indicates that this model is *not* statistically different from an",
  "'ideal' model of these data."
)
```

```{r smallSelSEMr2}
mod$R2 %>% 
  knitr::kable(
    digits = 2,
    caption = paste(
      "R-squared table of the selected SEM, fit with complete cases.",
      "Marginal R-squared values represent the proportion of variation in",
      "the resposne explained by the fixed effects and",
      "Conditional values represent the proportion explained by the full model.",
      "Models without random effects (i.e., simple linear regression) do not",
      "have a conditional R-squared value."
    )
  )
```

