
```{r}
library(tidyverse)
library(DiagrammeR)
```

```{r}
# Lookup table to recode variables
recode_tab <- tibble::tribble(
  ~old, ~new,
  "sex_male", "male",
  "sex_mature", "mature",
  "weight", "weight",
  "sex_male:weight", "male:weight",
  "sex_male:sex_mature", "male:mature",
  "sex_male:sex_mature:weight", "male:mature:weight",
  "ticks_attached", "ticks", 
  "expr_PC2", "tol PC",
  "expr_PC1", "res PC",
  "Bb_burden", "burden",
  "log_burden", "log burden",
  "Bb_infected", "infected",
  "cap_prop_night", "captime",
  "weighted_trapability", "trapable",
  "weighted_trap_diversity", "trapdiv",
  "avg_move_dist", "movement",
  "wthr_PC1", "wthr PC1",
  "wthr_PC2", "wthr PC2",
  "clim_PC1", "clim PC1",
  "clim_PC2", "clim PC2"
)
# Function to...
#' Replace values from key-value pairs
#'
#' @param x vector of values
#' @param old vector of values to be replaced
#' @param new vector of values to replace with
#'
#' @examples
#' key_val_recode(
#'   x = c("ham", "eggs", "eggs", "cheese"), 
#'   old = c("ham", "eggs"),
#'   new = c("green ham", "green eggs")
#' )
key_val_recode <- function(x, old = recode_tab$old, new = recode_tab$new){
  inx = match(x, old)
  dplyr::if_else(condition = is.na(inx), true = x, false = new[inx])
}
```


# Structural equation models

Structural equation models are a special case of path analysis, which describe
the directed dependence among sets of variables, often viewed as directed 
acyclic graphs (e.g., Figure \@ref(fig:pathExmpl)). These are an ideal tool
for understanding the relationships within and among many levels of
complex system with many variables. We will use this approach to model 
*Borrelia* infection competence of *Peromyscus*.

```{r pathExmpl, fig.width = 3, out.width = "50%", fig.cap = caption}
ndf <- create_node_df(
  n = 4, label = c("Y1", "Y2", "X1", "X2"), color = "black"
)
pdf <- create_edge_df(
  from = c(4, 4, 3, 1),
  to =   c(2, 1, 1, 2),
  # penwidth = c(1, 1, 3, 5),
  color = "black"
)

create_graph(attr_theme = "bt") %>% 
  # add_global_graph_attrs("layout", "dot", "graph") %>%
  add_global_graph_attrs("rankdir", "LR", "graph") %>%
  add_node_df(ndf) %>% 
  add_edge_df(pdf) %>% 
  render_graph()

caption = paste(
  "Example directed acyclic graph, with two response (Y) variables and two",
  "predictor (X) variables. In this case, Y1 is both a resonse (X1, X2) and",
  "a predictor (Y2)."
)
```

## Modeling approach

Towards finding a good explanatory model of *Borrelia* infection competence
among *Peromyscus*, we will fit a few different SEMs with different objectives.
The key data for each approach will be *Peromyscus* data for eight NEON sites.
The variables we will consider in each model are

0. Study design variables including NEON site and plot identity (`siteID`, 
`plotID`), and `year` of the observations as random effects;

1. Mouse general characteristics including individual identity (`iid`), sex 
(`male`), reproductive status (`sex_mature`), and weight (`weight`);

2. Mouse behavioral characteristics including capture time (`captime`),
trapability (`trapable`), trap diversity 
(`trapdiv`), and movement (`movement`);

3. Contact with ticks, specifically whether or not ticks were attached
at capture (`ticks`).

4. Immunity information including PCA for gene expression of resistance and
tolerance genes (`res PC`, `tol PC`) as well as *Borrelia* infection status
and burden (`infected`, `log burden`); and

5. Climatic information including principal components for short-term weather
(`wthr PC1`, `wthr PC2`) and long-term average site climate (`clim PC1`, 
`clim PC2`).

In general, the kinds of models we want to fit are as follows:

* [x] Fit with "complete cases" wherein all paths use only observations for
which every variable is available

* [x] Fit with maximal data, wherein all paths use as many observations as
possible

* [x] Model both infection status and burden:

  - [x] independent model, with complete cases
  
  - [x] independent model, with maximal data
  
  - [x] both responses in one model, with maximal data
  
* [ ] Compare model effects between *P leucopus* vs. *P maniculatus*

## Model selection

The form of each SEM presented below were arrived at with the following 
procedure:

1. Omnibus models were created for each component model, which were combined
into an omnibus SEM.

2. Model selection was performed on each component model by:

  i. removing any "singular" (i.e., variance = 0) random effects,
  
  ii. recursively removing variables with non-significant (fixed) effects, 
  one by one.   if any higher-order effects (interactions) were significant, 
  all constituent first-order (main) effects were retained regardless of 
  significance.
  
  iii. some non-significant fixed effects were added back into the model if 
  either a) the effect was considered "crucial" to our hypothesis. In practice
  this meant including links along the path behavior -> ticks -> resistance ->
  infection -> tolerance *or*; b) the removal resulted in other non-removed
  factors losing significance. 
  
3. A reduced SEM was fit by combining the selected component models from
step (2).

4. Additional paths were added to the SEM according tests of directed
separation (dsep), which indicates that a missing path is important for
explaining the observed data.

5. Some non-significant paths were removed if: a) they were not considered
critical to our understanding of the system *and*; b) the removal was not 
contraindicated by a follow-up dsep test.
  
Due to the complexity of these models, and at the recommendation of the 
`piecewiseSEM` package authors, This study considers $P < 0.1$ as the
p-value cutoff for statistical significance in the SEMs.

### Latent and Composite variables

The key feature that distinguishes SEMs from path analyses are the ability
to include latent (i.e., unmeasured) and composite variables. We have not yet , 
included these types of variables into our models, but some interesting options
might be:

* "Behavior" is an obvious **latent** variable to explore. It would 
be constructed such that it depends on the behavior metrics that we have 
measured (i.e.,, activity time, movement, learning) as well as sex and 
reproductive status.

* A "body condition" **composite** variable, constructed from all the 
size traits recorded by NEON may be a better predictor variable than weight
alone, though it would require using a smaller data set since weight was
by far the most measured physical trait.

* "Immunity" would be a natural **latent** or **composite** variable 
constructed from the six gene expression variables. This may capture more 
variation than the principal component proxies we are currently using.

It is unclear, and not well-documented, how to include either latent or 
composite variables in `piecewiseSEM`.

---- 

**Update**: After further investigation, `piecewiseSEM` *does not* support 
either of these use cases. If needed, the Bayesian implementation in the 
[`brms` package](https://jebyrnes.github.io/bayesian_sem/bayesian_sem.html)
supposedly supports it.

## Complete cases

The first set of models is fit with only the subset of observations for which
information for *all* variables is available. As such, 
it has some limitations in its power, but it is the most "defensible" 
explanatory model because it does not draw conclusions beyond the scope of its
data. Models using more of the data will be explored in the next section.

### Burden

Figure \@ref(fig:selBurdCompSem) visualize a burden SEM fit to a "complete" 
data set. 
While this model is very interesting, one of the primary paths of interest
is not significant (behavior -> ticks -> resistance -> burden -> tolerance). It 
is possible that the "behavior -> ticks" path is improved by a model which is 
not limited by the completeness restriction. Table \@ref(tab:selBurdCompSemR2)
summarizes the fit ($R^2$) of each of the component models.

```{r selBurdCompSem, out.width="100%", fig.asp = 1, fig.cap=caption}
selBurdComp <- readRDS(
  "infection-modeling/data/model-objects/selected-burden-SEM_complete-cases.rds"
)

pbreaks = c(0, 0.05, 0.1, 1)#, 
pcols = c("black", "grey70", "grey70")#, 
linetypes = c("solid", "solid", "dotted")#, 
penwidth_range = c(1, 8)

# coefficients
coef_tab <- selBurdComp$coefficients %>% 
  data.frame() %>% 
  rename(sig.star = "Var.9") %>% 
  mutate(
    pval_group = cut(
      P.Value, breaks = pbreaks, include.lowest = TRUE,
      labels(seq_len(length(pbreaks) - 1))
    ),
    pval_col = pcols[pval_group],
    pval_linetype = linetypes[pval_group],
    pval_penwidth = scales::rescale(Std.Estimate, to = penwidth_range),
    color = case_when(
      P.Value > 0.1 ~ "grey50",
      Std.Estimate < 0 ~ "cornflowerblue", 
      Std.Estimate >= 0 ~ "orange"
    )
  ) %>% 
  mutate(
    Predictor = key_val_recode(Predictor),
    Response = key_val_recode(Response)
  )

# all variables
the_vars <- unique(c(coef_tab$Response, coef_tab$Predictor))

path_nodes <- data.frame(var = the_vars, id = seq_len(length(the_vars)))

create_graph(attr_theme = "bt") %>% 
  add_global_graph_attrs("layout", "dot", "graph") %>%
  # add_global_graph_attrs("rankdir", "LR", "graph") %>%
  add_nodes_from_table(path_nodes, label_col = var) %>% 
  add_edges_from_table(
    coef_tab, from_col = "Predictor", to_col = "Response", 
    from_to_map = "label"
  ) %>% 
  set_node_attrs(width, 2) %>% 
  set_node_attrs(fontsize, 20) %>% 
  set_node_attrs(shape, "rectangle") %>% 
  # set_edge_attrs(color, coef_tab$pval_col) %>% 
  set_edge_attrs(arrowsize, 1) %>% 
  set_edge_attrs(color, coef_tab$color) %>% 
  set_edge_attrs(style, coef_tab$pval_linetype) %>% 
  set_edge_attrs(penwidth, coef_tab$pval_penwidth) %>% 
  render_graph()

caption = paste(
  "Directed acyclycal graph of selected burden SEM model, fit with complete cases.",
  "The goodness-of-fit P-value is", paste0(round(selBurdComp$Cstat$P.Value, 2), ","),
  "which indicates that this model is *not* statistically different from an",
  "'ideal' model of these data.", 
  "Orange vectors represent positive effects, blue represent negative effects",
  "and grey represents non-significant (P > 0.1) effects.",
  "The width of a vector is proportional to the magnitude of its effect."
)
```

<br>
```{r selBurdCompSemR2}
selBurdComp$R2 %>%
    mutate(
    Response = key_val_recode(Response)
  ) %>%
  knitr::kable(
    digits = 2,
    caption = paste(
      "R-squared table of the selected SEM, fit with complete cases.",
      "Marginal R-squared values represent the proportion of variation in",
      "the response explained by the fixed effects and",
      "Conditional values represent the proportion explained by the full model.",
      "Models without random effects (i.e., simple linear regression) do not",
      "have a conditional R-squared value."
    )
  )
```

### Infection status

Figure \@ref(fig:selInfCompSem) and Table \@ref(tab:selInfCompSemR2) visualize 
a the infection SEM fit to a "complete" data set. Again, while this model
is interesting, there are no significant paths between our primary variables
of interest (behavior -> ticks -> resistance -> infection -> tolerance).

```{r selInfCompSem, out.width="100%", fig.asp = 1, fig.cap=caption}
selInfComp <- readRDS(
  "infection-modeling/data/model-objects/selected-infection-SEM_complete-cases.rds"
)

pbreaks = c(0, 0.05, 0.1, 1)#, 
pcols = c("black", "grey70", "grey70")#, 
linetypes = c("solid", "solid", "dotted")#, 
penwidth_range = c(1, 8)

# coefficients
coef_tab <- selInfComp$coefficients %>% 
  data.frame() %>% 
  rename(sig.star = "Var.9") %>% 
  mutate(
    pval_group = cut(
      P.Value, breaks = pbreaks, include.lowest = TRUE,
      labels(seq_len(length(pbreaks) - 1))
    ),
    pval_col = pcols[pval_group],
    pval_linetype = linetypes[pval_group],
    pval_penwidth = scales::rescale(Std.Estimate, to = penwidth_range),
    color = case_when(
      P.Value > 0.1 ~ "grey50",
      Std.Estimate < 0 ~ "cornflowerblue", 
      Std.Estimate >= 0 ~ "orange"
    )
  ) %>% 
  mutate(
    Predictor = key_val_recode(Predictor),
    Response = key_val_recode(Response)
  )


# all variables
the_vars <- unique(c(coef_tab$Response, coef_tab$Predictor))

path_nodes <- data.frame(var = the_vars, id = seq_len(length(the_vars)))

create_graph(attr_theme = "bt") %>% 
  add_global_graph_attrs("layout", "dot", "graph") %>%
  # add_global_graph_attrs("rankdir", "LR", "graph") %>%
  add_nodes_from_table(path_nodes, label_col = var) %>% 
  add_edges_from_table(
    coef_tab, from_col = "Predictor", to_col = "Response", 
    from_to_map = "label"
  ) %>% 
  set_node_attrs(width, 2.2) %>% 
  set_node_attrs(fontsize, 20) %>% 
  set_node_attrs(shape, "rectangle") %>% 
  # set_edge_attrs(color, coef_tab$pval_col) %>% 
  set_edge_attrs(arrowsize, 1) %>% 
  set_edge_attrs(color, coef_tab$color) %>% 
  set_edge_attrs(style, coef_tab$pval_linetype) %>% 
  set_edge_attrs(penwidth, coef_tab$pval_penwidth) %>% 
  render_graph()


caption = paste(
  "Directed acyclycal graph of selected infection SEM model, fit with complete cases.",
  "The goodness-of-fit P-value is", paste0(round(selInfComp$Cstat$P.Value, 2), ","),
  "which indicates that this model is *not* statistically different from an",
  "'ideal' model of these data.", 
  "Orange vectors represent positive effects, blue represent negative effects",
  "and grey represents non-significant (P > 0.1) effects.",
  "The width of a vector is proportional to the magnitude of its effect."
)
```

<br>
```{r selInfCompSemR2}
selInfComp$R2 %>% 
  mutate(
    Response = key_val_recode(Response)
  ) %>% 
  knitr::kable(
    digits = 2,
    caption = paste(
      "R-squared table of the selected infection SEM, fit with complete cases.",
      "Marginal R-squared values represent the proportion of variation in",
      "the responne explained by the fixed effects and",
      "Conditional values represent the proportion explained by the full model.",
      "Models without random effects (i.e., simple linear regression) do not",
      "have a conditional R-squared value."
    )
  )
```

## Maximal cases

In contrast to the (#complete-cases) section above, these next SEMs are fit
with component models that use all available observations. This means that some
paths (i.e., those involving resistance and tolerance expression) are fit with
fewer samples than other. This gives the both the components and overall model
more power to estimate the effects, but does assume that the patterns among
individuals with observed characteristics are representative of individuals 
with characteristics not directly observed.

### Burden

Figure \@ref(fig:selBurdMaxSem) shows the selected burden model, fit with the
maximal data for each path. And Table \@ref(tab:selBurdMaxSemR2) shows the 
corresponding $R^2$ table. Still the paths between burden, gene expression,
and ticks, and behavior are not significant.

```{r selBurdMaxSem, out.width="100%", fig.asp = 1, fig.cap=caption}
selBurdMax <- readRDS(
  "infection-modeling/data/model-objects/selected-burden-SEM_maximal-cases.rds"
)

pbreaks = c(0, 0.05, 0.1, 1)#, 
pcols = c("black", "grey70", "grey70")#, 
linetypes = c("solid", "solid", "dotted")#, 
penwidth_range = c(1, 8)

# coefficients
coef_tab <- selBurdMax$coefficients %>% 
  data.frame() %>% 
  rename(sig.star = "Var.9") %>% 
  mutate(
    pval_group = cut(
      P.Value, breaks = pbreaks, include.lowest = TRUE,
      labels(seq_len(length(pbreaks) - 1))
    ),
    pval_col = pcols[pval_group],
    pval_linetype = linetypes[pval_group],
    pval_penwidth = scales::rescale(Std.Estimate, to = penwidth_range),
    color = case_when(
      P.Value > 0.1 ~ "grey50",
      Std.Estimate < 0 ~ "cornflowerblue", 
      Std.Estimate >= 0 ~ "orange"
    )
  ) %>% 
  mutate(
    Predictor = key_val_recode(Predictor),
    Response = key_val_recode(Response)
  )

# all variables
the_vars <- unique(c(coef_tab$Response, coef_tab$Predictor))

path_nodes <- data.frame(var = the_vars, id = seq_len(length(the_vars)))

create_graph(attr_theme = "bt") %>% 
  add_global_graph_attrs("layout", "dot", "graph") %>%
  # add_global_graph_attrs("rankdir", "LR", "graph") %>%
  add_nodes_from_table(path_nodes, label_col = var) %>% 
  add_edges_from_table(
    coef_tab, from_col = "Predictor", to_col = "Response", 
    from_to_map = "label"
  ) %>% 
  set_node_attrs(shape, "rectangle") %>% 
  set_node_attrs(width, 1.8) %>% 
  set_node_attrs(fontsize, 20) %>% 
  # set_node_attrs(width, 1.5) %>% 
  # set_node_attrs(fontsize, 20) %>% 
  # set_edge_attrs(color, coef_tab$pval_col) %>% 
  set_edge_attrs(arrowsize, 1) %>% 
  set_edge_attrs(color, coef_tab$color) %>% 
  set_edge_attrs(style, coef_tab$pval_linetype) %>% 
  set_edge_attrs(penwidth, coef_tab$pval_penwidth) %>% 
  render_graph()


caption = paste(
  "Directed acyclycal graph of selected burden SEM model, fit with maximal cases.",
  "The goodness-of-fit P-value is", paste0(round(selBurdMax$Cstat$P.Value, 2), ","),
  "which indicates that this model is *not* statistically different from an",
  "'ideal' model of these data.",
  "Orange vectors represent positive effects, blue represent negative effects",
  "and grey represents non-significant (P > 0.1) effects.",
  "The width of a vector is proportional to the magnitude of its effect."
)
```

<br>
```{r selBurdMaxSemR2}
selBurdMax$R2 %>% 
  mutate(
    Response = key_val_recode(Response)
  ) %>% 
  knitr::kable(
    digits = 2,
    caption = paste(
      "R-squared table of the selected SEM, fit with complete cases.",
      "Marginal R-squared values represent the proportion of variation in",
      "the response explained by the fixed effects and",
      "Conditional values represent the proportion explained by the full model.",
      "Models without random effects (i.e., simple linear regression) do not",
      "have a conditional R-squared value."
    )
  )
```

### Infection Status

Figure \@ref(fig:selInfMaxSem) and Table (\@ref(tab:selInfMaxSemR2)) summarize 
the infection status
SEM, fit with maximal data. This is the first model that suggests a link from
ticks -> infection -> tolerance expression. 

```{r selInfMaxSem, out.width="100%", fig.asp = 1, fig.cap=caption}
selInfMax <- readRDS(
  "infection-modeling/data/model-objects/selected-infection-SEM_maximal-cases.rds"
)

pbreaks = c(0, 0.05, 0.1, 1)#, 
pcols = c("black", "grey70", "grey70")#, 
linetypes = c("solid", "solid", "dotted")#, 
penwidth_range = c(1, 8)

# coefficients
coef_tab <- selInfMax$coefficients %>% 
  data.frame() %>% 
  rename(sig.star = "Var.9") %>% 
  mutate(
    pval_group = cut(
      P.Value, breaks = pbreaks, include.lowest = TRUE,
      labels(seq_len(length(pbreaks) - 1))
    ),
    pval_col = pcols[pval_group],
    pval_linetype = linetypes[pval_group],
    pval_penwidth = scales::rescale(Std.Estimate, to = penwidth_range),
    color = case_when(
      P.Value > 0.1 ~ "grey50",
      Std.Estimate < 0 ~ "cornflowerblue", 
      Std.Estimate >= 0 ~ "orange"
    )
  ) %>% 
  mutate(
    Predictor = key_val_recode(Predictor),
    Response = key_val_recode(Response)
  )

# all variables
the_vars <- unique(c(coef_tab$Response, coef_tab$Predictor))

path_nodes <- data.frame(var = the_vars, id = seq_len(length(the_vars)))

create_graph(attr_theme = "bt") %>% 
  add_global_graph_attrs("layout", "dot", "graph") %>%
  # add_global_graph_attrs("rankdir", "LR", "graph") %>% 
  add_nodes_from_table(path_nodes, label_col = var) %>% 
  add_edges_from_table(
    coef_tab, from_col = "Predictor", to_col = "Response", 
    from_to_map = "label"
  ) %>% 
  set_node_attrs(shape, "rectangle") %>% 
  set_node_attrs(width, 2.2) %>% 
  set_node_attrs(fontsize, 20) %>% 
  # set_node_attrs(width, 1.5) %>% 
  # set_node_attrs(fontsize, 20) %>% 
  # set_edge_attrs(color, coef_tab$pval_col) %>% 
  set_edge_attrs(arrowsize, 1) %>% 
  set_edge_attrs(color, coef_tab$color) %>% 
  set_edge_attrs(style, coef_tab$pval_linetype) %>% 
  set_edge_attrs(penwidth, coef_tab$pval_penwidth) %>% 
  render_graph()


caption = paste(
  "Directed acyclycal graph of selected burden SEM model, fit with maximal cases.",
  "The goodness-of-fit P-value is", paste0(round(selInfMax$Cstat$P.Value, 2), ","),
  "which indicates that this model is *not* statistically different from an",
  "'ideal' model of these data.",
  "Orange vectors represent positive effects, blue represent negative effects",
  "and grey represents non-significant (P > 0.1) effects.",
  "The width of a vector is proportional to the magnitude of its effect."
)
```

<br>
```{r selInfMaxSemR2}
selInfMax$R2 %>% 
  mutate(
    Response = key_val_recode(Response)
  ) %>% 
  knitr::kable(
    digits = 2,
    caption = paste(
      "R-squared table of the selected infection SEM, fit with complete cases.",
      "Marginal R-squared values represent the proportion of variation in",
      "the response explained by the fixed effects and",
      "Conditional values represent the proportion explained by the full model.",
      "Models without random effects (i.e., simple linear regression) do not",
      "have a conditional R-squared value."
    )
  )
```
### Burden and Infection

Figure \@ref(fig:selBothMaxSem) and Table \@ref(tab:selBothMaxSemR2) 
summarize what is probably the most
interesting model of this system; it includes both infection and burden and is
fit with the maximal data set. This SEM suggests that:

* *Borrelia* infection in an individual *Peromyscus* is most explained by 
the three-way interaction between the individual's sex, reproductive status,
and weight; contact with ticks; and weather conditions, in
addition to random spatial and temporal effects (35% of variation).

* The quantitative *Borrelia* burden within an individual, beyond its infection
status, is best explained by the individuals sex (there is also a marginal
interaction between sex and weight: $P=0.12$), in addition to random spatial
effects (1% of total variation). Unsurprisingly, most of the variation in burden is 
explained by infection status.

* Expression of tolerance genes was best explained by individual reproductive 
status, behavior (trap diversity), and *Borrelia* infection status, in addition 
to random spatial and temporal effects. The burden of *Borrelia* apparently had
no bearing on tolerance expression (removed during model selection).

* Expression of resistance genes was best explained by individual sex, weight,
and contact with ticks (there is also a marginal effect of weather: 
$P=0.13$),
in addition to random spatial effects. Resistance expression apparently
had no bearing on either **Borrelia** infection status or burden ($P>0.75$). 

* Our metrics of behavior (capture time, trapability, trap diversity,
movement distance) did not explain significant variation in *Borrelia*
infection, burden, or contact with ticks -- beyond that already
explained by sex, reproductive status, and weight. 

```{r selBothMaxSem, out.width="100%", fig.asp = 1, fig.cap=caption}
selBothMax <- readRDS(
  file.path(
    "infection-modeling/data/model-objects",
    "selected-BurdenInfection-SEM_maximal-cases.rds"
  )
)

pbreaks = c(0, 0.05, 0.1, 1)#, 
pcols = c("black", "grey70", "grey70")#, 
linetypes = c("solid", "solid", "dotted")#, 
penwidth_range = c(1, 8)

# coefficients
coef_tab <- selBothMax$coefficients %>% 
  data.frame() %>% 
  rename(sig.star = "Var.9") %>% 
  mutate(
    pval_group = cut(
      P.Value, breaks = pbreaks, include.lowest = TRUE,
      labels(seq_len(length(pbreaks) - 1))
    ),
    pval_col = pcols[pval_group],
    pval_linetype = linetypes[pval_group],
    pval_penwidth = scales::rescale(Std.Estimate, to = penwidth_range),
    color = case_when(
      P.Value > 0.1 ~ "grey50",
      Std.Estimate < 0 ~ "cornflowerblue", 
      Std.Estimate >= 0 ~ "orange"
    )
  ) %>% 
  mutate(
    Predictor = key_val_recode(Predictor),
    Response = key_val_recode(Response)
  )

# all variables
the_vars <- unique(c(coef_tab$Response, coef_tab$Predictor))

path_nodes <- data.frame(var = the_vars, id = seq_len(length(the_vars)))

create_graph(attr_theme = "bt") %>% 
  add_global_graph_attrs("layout", "dot", "graph") %>%
  # add_global_graph_attrs("rankdir", "LR", "graph") %>%
  add_nodes_from_table(path_nodes, label_col = var) %>% 
  add_edges_from_table(
    coef_tab, from_col = "Predictor", to_col = "Response", 
    from_to_map = "label"
  ) %>% 
  set_node_attrs(shape, "rectangle") %>% 
  set_node_attrs(width, if_else(path_nodes$var == "sex:mature:weight", 2.2, 2)) %>% 
  set_node_attrs(fontsize, 20) %>% 
  # set_node_attrs(width, 1.5) %>% 
  # set_node_attrs(fontsize, 20) %>% 
  # set_edge_attrs(color, coef_tab$pval_col) %>% 
  set_edge_attrs(arrowsize, 1) %>% 
  set_edge_attrs(color, coef_tab$color) %>% 
  set_edge_attrs(style, coef_tab$pval_linetype) %>% 
  set_edge_attrs(penwidth, coef_tab$pval_penwidth) %>% 
  render_graph()

caption = paste(
  "Directed acyclycal graph of selected burden SEM model, fit with maximal cases.",
  "The goodness-of-fit P-value is", paste0(round(selBothMax$Cstat$P.Value, 2), ","),
  "which indicates that this model is *not* statistically different from an",
  "'ideal' model of these data.",
  "Orange vectors represent positive effects, blue represent negative effects",
  "and grey represents non-significant (P > 0.1) effects.",
  "The width of a vector is proportional to the magnitude of its effect."
)
```

<br>
```{r selBothMaxSemR2}
selBothMax$R2 %>% 
  mutate(
    Response = key_val_recode(Response)
  ) %>% 
  knitr::kable(
    digits = 2,
    caption = paste(
      "R-squared table of the selected burden-infection SEM, fit with complete cases.",
      "Marginal R-squared values represent the proportion of variation in",
      "the response explained by the fixed effects and",
      "Conditional values represent the proportion explained by the full model.",
      "Models without random effects (i.e., simple linear regression) do not",
      "have a conditional R-squared value."
    )
  )
```

## Alternate Models

Here, we'll look at some olternate versions of a model

### Tolerance as predictor

Swapping tolerance from a response to a predictor of infection/burden
(compared with the model in Figure \@ref(fig:selBothMaxSem)) yields a
model where the link between tolerance and infection is not significant 
(Figure \@ref(fig:altBothMaxSem)).

```{r altBothMaxSem, out.width="100%", fig.asp = 1, fig.cap=caption}
altBothMax <- readRDS(
  file.path(
    "infection-modeling/data/model-objects",
    "alternate-BurdenInfection-SEM_maximal-cases.rds"
  )
)

pbreaks = c(0, 0.05, 0.1, 1)#, 
pcols = c("black", "grey70", "grey70")#, 
linetypes = c("solid", "solid", "dotted")#, 
penwidth_range = c(1, 8)

# coefficients
coef_tab <- altBothMax$coefficients %>% 
  data.frame() %>% 
  rename(sig.star = "Var.9") %>% 
  mutate(
    pval_group = cut(
      P.Value, breaks = pbreaks, include.lowest = TRUE,
      labels(seq_len(length(pbreaks) - 1))
    ),
    pval_col = pcols[pval_group],
    pval_linetype = linetypes[pval_group],
    pval_penwidth = scales::rescale(Std.Estimate, to = penwidth_range),
    color = case_when(
      P.Value > 0.1 ~ "grey50",
      Std.Estimate < 0 ~ "cornflowerblue", 
      Std.Estimate >= 0 ~ "orange"
    )
  ) %>% 
  mutate(
    Predictor = key_val_recode(Predictor),
    Response = key_val_recode(Response)
  )

# all variables
the_vars <- unique(c(coef_tab$Response, coef_tab$Predictor))

path_nodes <- data.frame(var = the_vars, id = seq_len(length(the_vars)))

create_graph(attr_theme = "bt") %>% 
  add_global_graph_attrs("layout", "dot", "graph") %>%
  add_nodes_from_table(path_nodes, label_col = var) %>% 
  add_edges_from_table(
    coef_tab, from_col = "Predictor", to_col = "Response", 
    from_to_map = "label"
  ) %>% 
  set_node_attrs(shape, "rectangle") %>% 
  set_node_attrs(width, if_else(path_nodes$var == "sex:mature:weight", 2.2, 2)) %>% 
  set_node_attrs(fontsize, 20) %>% 
  set_edge_attrs(arrowsize, 1) %>% 
  set_edge_attrs(color, coef_tab$color) %>% 
  set_edge_attrs(style, coef_tab$pval_linetype) %>% 
  set_edge_attrs(penwidth, coef_tab$pval_penwidth) %>% 
  render_graph()

caption = paste(
  "Directed acyclycal graph of an alternate (tolerance as a response)",
  "burden SEM model, fit with maximal cases.",
  "The goodness-of-fit P-value is", 
  paste0(round(altBothMax$Cstat$P.Value, 2), ","),
  "which indicates that this model is *not* statistically different from an",
  "'ideal' model of these data.",
  "Orange vectors represent positive effects, blue represent negative effects",
  "and grey represents non-significant (P > 0.1) effects.",
  "The width of a vector is proportional to the magnitude of its effect."
)
```

### Condensed behaviors (PCA)

Figure \@ref(fig:behaviorPCA) shows the principal component analysis of our 
four behavioral traits. These behavior traits were transformed with the function
`log(x + min(x) + 1)`. Here, the primary PC is associated with trapability,
trap diversity, and movement distance, while the the second axis is associated 
capture timing. As Figure \@ref(fig:BehaveBothMaxMod), an SEM wherein the four
individual behavior traits are condensed into their principal components 
exhibits no relationships with those principal components and any of our
response variables of interest. This pattern also remains when tolerance is 
treated as a predictor of infection (not shown).

```{r behaviorPCA, fig.asp = 1, fig.width = 6, fig.cap = caption}
bpca_list <- readRDS("infection-modeling/data/peromyscus-behavior-pca.rds")

# ordiplot(bpca_list$pca_obj, display = "species", type = "text", arrows = TRUE)
vegan::ordiplot(bpca_list$pca_obj, xlim = c(-5.5, 2.5), cex = .2); vegan::orditorp(bpca_list$pca_obj, display = "species")

bpca_import <- summary(bpca_list$pca_obj)$cont$importance

caption = paste(
  "Peromyscus behavior PCA. PC1 explains 47% of the variation and PC2 explains",
  "an additional 25%."
)
```

```{r BehaveBothMaxMod, out.width="100%", fig.asp = 1, fig.cap=caption}
BehaveBothMax <- readRDS(
  file.path(
    "infection-modeling/data/model-objects",
    "behaviorPCA-BurdenInfection-SEM_maximal-cases.rds"
  )
)

pbreaks = c(0, 0.05, 0.1, 1)#, 
pcols = c("black", "grey70", "grey70")#, 
linetypes = c("solid", "solid", "dotted")#, 
penwidth_range = c(1, 8)

# coefficients
coef_tab <- BehaveBothMax$coefficients %>% 
  data.frame() %>% 
  rename(sig.star = "Var.9") %>% 
  mutate(
    pval_group = cut(
      P.Value, breaks = pbreaks, include.lowest = TRUE,
      labels(seq_len(length(pbreaks) - 1))
    ),
    pval_col = pcols[pval_group],
    pval_linetype = linetypes[pval_group],
    pval_penwidth = scales::rescale(Std.Estimate, to = penwidth_range),
    color = case_when(
      P.Value > 0.1 ~ "grey50",
      Std.Estimate < 0 ~ "cornflowerblue", 
      Std.Estimate >= 0 ~ "orange"
    )
  ) %>% 
  mutate(
    Predictor = key_val_recode(Predictor),
    Response = key_val_recode(Response)
  )

# all variables
the_vars <- unique(c(coef_tab$Response, coef_tab$Predictor))

path_nodes <- data.frame(var = the_vars, id = seq_len(length(the_vars)))

create_graph(attr_theme = "bt") %>% 
  add_global_graph_attrs("layout", "dot", "graph") %>%
  add_nodes_from_table(path_nodes, label_col = var) %>% 
  add_edges_from_table(
    coef_tab, from_col = "Predictor", to_col = "Response", 
    from_to_map = "label"
  ) %>% 
  set_node_attrs(shape, "rectangle") %>% 
  set_node_attrs(width, if_else(path_nodes$var == "sex:mature:weight", 2.2, 2)) %>% 
  set_node_attrs(fontsize, 20) %>% 
  set_edge_attrs(arrowsize, 1) %>% 
  set_edge_attrs(color, coef_tab$color) %>% 
  set_edge_attrs(style, coef_tab$pval_linetype) %>% 
  set_edge_attrs(penwidth, coef_tab$pval_penwidth) %>% 
  render_graph()

caption = paste(
  "Directed acyclycal graph of an alternate",
  "infection SEM model, fit with maximal cases and behavioral PCs as predictors.",
  "The goodness-of-fit P-value is", 
  paste0(round(BehaveBothMax$Cstat$P.Value, 2), ","),
  "which indicates that this model is *not* statistically different from an",
  "'ideal' model of these data.",
  "Orange vectors represent positive effects, blue represent negative effects",
  "and grey represents non-significant (P > 0.1) effects.",
  "The width of a vector is proportional to the magnitude of its effect."
)
```


## *P. leucopus* vs. *P. maniculatus*

TBD

## Next steps

* Prediction: how can the expression module (smaller data) inform the larger 
things

  - can we map expression to something else? Proxy variables for expression.
  
  - use 200 mice to fine tune estimation
  
* Maybe look at proposal for machine learning for more general disease 
(all kinds) prediction?
