
```{r}
library(tidyverse)
library(DiagrammeR)
```

# Structural equation models

Structural equation models are a special case of path analysis, which describe
the directed dependence among sets of variables, often viewed as directed 
acyclic graphs (e.g., Figure \@ref(fig:pathExmpl)). These are an ideal tool
for understanding the relationships within and among many levels of
complex system with many variables. We will use this approach to model 
*Borrelia* infection competence of *Peromyscus*.

```{r pathExmpl, fig.width = 3, out.width = "25%", fig.cap = caption}
ndf <- create_node_df(
  n = 4, label = c("Y1", "Y2", "X1", "X2"), color = "black"
)
pdf <- create_edge_df(
  from = c(4, 4, 3, 1),
  to =   c(2, 1, 1, 2),
  # penwidth = c(1, 1, 3, 5),
  color = "black"
)

create_graph(attr_theme = "bt") %>% 
  add_global_graph_attrs("layout", "dot", "graph") %>%
  add_node_df(ndf) %>% 
  add_edge_df(pdf) %>% 
  render_graph()

caption = paste(
  "Example directed acyclic graph, with two response (Y) variables and two",
  "predictor (X) variables. In this case, Y1 is both a resonse (X1, X2) and",
  "a predictor (Y2)."
)
```

## Modeling approach

Towards finding a good explanatory model of *Borrelia* infection competence
among *Peromyscus*, we will fit a few different SEMs with different objectives.
The key data for each approach will be *Peromyscus* data for eight NEON sites.
The variables we will consider in each model are

0. Study design variables including NEON site and plot identity (`siteID`, 
`plotID`), and `year` of the observations;

1. Mouse general characteristics including individual identity (`iid`), sex 
(`sex_male`), reproductive status (`sex_mature`), and weight (`weight`);

2. Mouse behavioral characteristics including capture time (`cap_prop_night`),
trapability (`weighted_trapability`), trap diversity 
(`weighted_trap_diversity`), and movement (`avg_move_dist`);

3. Interactions with ticks specifically whether or not ticks were attactched
at capture (`ticks_attached`, `nymphalTicksAttached`, `larvalTicksAttached`).

4. Immunity information including PCA for gene expression of resistance and
tolerance genes (`expr_PC1`, `expr_PC2`) as well as *Borrelia* infection status
and burden (`Bb_infected`, `Bb_burden`); and

5. Climatic information including principal components for short-term weather
(`wthr_PC1`, `wthr_PC2`) and long-term average site climate (`clim_PC`, 
`clim_PC2`).

In general, the kinds of models we want to fit are as follows:

* [x] Fit with "complete cases" wherein all paths use only observations for
which every variable is available

* [ ] Fit with maximal data, wherein all paths use as many observations as
possible

* [ ] Model both infection status and burden:

  - [ ] independent model with complete cases
  
  - [ ] independent model with maximal data
  
  - [ ] "hurdle" style model (both responses in one model) with complete cases
  
  - [ ] "hurdle" style model (both responses in one model) with maximal data
  
* [ ] Compare model effects between *P leucopus* vs. *P maniculatus*

  <!-- Should this be for **all** models, or just a select model/subset? -->
  

### Latent and Composite variables

The key feature that distinguishes SEMs from path analyses are the ability
to include latent (i.e., unmeasured) and composite variables. We have not yet , 
included these types of variables into our models, but some interesting options
might be:

* [ ] "Behavior" is an obvious **latent** variable to explore. It would 
be constructed such that it depends on the behavior metrics that we have 
measured (i.e.,, activity time, movement, learning) as well as sex and 
reproductive status.

* [ ] A "body condition" **composite** variable, constructed from all the 
size traits recorded by NEON may be a better predictor variable than weight
alone, though it would require using a smaller data set since weight was
by far the most measured physical trait.

* [ ] "Immunity" would be a natural **latent** or **composite** variable 
constructed from the six gene expression variables. This may capture more 
variation than the principal component proxies we are currently using.

It is unclear, and not well-documented, how to include either latent or 
composite variables in `piecewiseSEM`.

## Complete cases

### Burden

Figure \@ref(fig:selBurdCompSem) visualizes a SEM fit to a "complete" data set
(i.e., only observations for which *all* variables were measured). As such, 
it has some limitations in its power, but it is the most "defensible" 
explanatory model because it does not draw conclusions beyond the scope of its
data. Models using more of the data will be explored in the next section.

While this model is very interesting, one of the primary paths of interest
is not significant (behavior -> ticks -> resistance -> burden -> tolerance). It 
is possible
that the "behavior -> ticks" path is improved by a model which is not limited
by the completeness restriction. Table \@ref(tab:selBurdCompSemR2) summarizes the
fit ($R^2$) of each of the component models.

```{r selBurdCompSem, out.width="100%", fig.asp = 1, fig.cap=caption}
selBurdComp <- readRDS(
  "infection-modeling/data/model-objects/selected-burden-SEM_complete-cases.rds"
)

pbreaks = c(0, 0.05, 0.1, 1)#, 
pcols = c("black", "grey70", "grey70")#, 
linetypes = c("solid", "solid", "dotted")#, 
penwidth_range = c(1, 8)

# coefficients
coef_tab <- selBurdComp$coefficients %>% 
  data.frame() %>% 
  rename(sig.star = "Var.9") %>% 
  mutate(
    pval_group = cut(
      P.Value, breaks = pbreaks, include.lowest = TRUE,
      labels(seq_len(length(pbreaks) - 1))
    ),
    pval_col = pcols[pval_group],
    pval_linetype = linetypes[pval_group],
    pval_penwidth = scales::rescale(Std.Estimate, to = penwidth_range),
    color = case_when(
      P.Value > 0.1 ~ "grey50",
      Std.Estimate < 0 ~ "cornflowerblue", 
      Std.Estimate >= 0 ~ "orange"
    )
  )


# all variables
the_vars <- unique(c(coef_tab$Response, coef_tab$Predictor))

path_nodes <- data.frame(var = the_vars, id = seq_len(length(the_vars)))

create_graph(attr_theme = "bt") %>% 
  add_global_graph_attrs("layout", "dot", "graph") %>%
  # add_global_graph_attrs("rankdir", "LR", "graph") %>% 
  add_nodes_from_table(path_nodes, label_col = var) %>% 
  add_edges_from_table(
    coef_tab, from_col = "Predictor", to_col = "Response", 
    from_to_map = "label"
  ) %>% 
  set_node_attrs(width, 3) %>% 
  set_node_attrs(fontsize, 20) %>% 
  set_node_attrs(shape, "rectangle") %>% 
  # set_edge_attrs(color, coef_tab$pval_col) %>% 
  set_edge_attrs(arrowsize, 1) %>% 
  set_edge_attrs(color, coef_tab$color) %>% 
  set_edge_attrs(style, coef_tab$pval_linetype) %>% 
  set_edge_attrs(penwidth, coef_tab$pval_penwidth) %>% 
  render_graph()


caption = paste(
  "Directed acyclycal graph of selected burden SEM model, fit with complete cases.",
  "The goodness-of-fit P-value is", paste0(round(selBurdComp$Cstat$P.Value, 2), ","),
  "which indicates that this model is *not* statistically different from an",
  "'ideal' model of these data.", 
  "Orange vectors represent positive effects, blue represent negative effects",
  "and grey represents non-significant (P > 0.1) effects.",
  "The width of a vector is proportional to the magnitude of its effect."
)
```

<br>
```{r selBurdCompSemR2}
selBurdComp$R2 %>% 
  knitr::kable(
    digits = 2,
    caption = paste(
      "R-squared table of the selected SEM, fit with complete cases.",
      "Marginal R-squared values represent the proportion of variation in",
      "the resposne explained by the fixed effects and",
      "Conditional values represent the proportion explained by the full model.",
      "Models without random effects (i.e., simple linear regression) do not",
      "have a conditional R-squared value."
    )
  )
```

### Infection status

TBD

### Infection-burden hurdle

TBD

## Maixmal cases

Figure \@ref(fig:selBurdMaxSem) shows the selected burden model, fit with the
maximal data for each path. And Table \@ref(tab:selBurdMaxSemR2) shows the 
corresponding $R^2$ table

```{r selBurdMaxSem, out.width="100%", fig.asp = 1, fig.cap=caption}
selBurdMax <- readRDS(
  "infection-modeling/data/model-objects/selected-burden-SEM_maximal-cases.rds"
)

pbreaks = c(0, 0.05, 0.1, 1)#, 
pcols = c("black", "grey70", "grey70")#, 
linetypes = c("solid", "solid", "dotted")#, 
penwidth_range = c(1, 8)

# coefficients
coef_tab <- selBurdMax$coefficients %>% 
  data.frame() %>% 
  rename(sig.star = "Var.9") %>% 
  mutate(
    pval_group = cut(
      P.Value, breaks = pbreaks, include.lowest = TRUE,
      labels(seq_len(length(pbreaks) - 1))
    ),
    pval_col = pcols[pval_group],
    pval_linetype = linetypes[pval_group],
    pval_penwidth = scales::rescale(Std.Estimate, to = penwidth_range),
    color = case_when(
      P.Value > 0.1 ~ "grey50",
      Std.Estimate < 0 ~ "cornflowerblue", 
      Std.Estimate >= 0 ~ "orange"
    )
  )

# all variables
the_vars <- unique(c(coef_tab$Response, coef_tab$Predictor))

path_nodes <- data.frame(var = the_vars, id = seq_len(length(the_vars)))

create_graph(attr_theme = "bt") %>% 
  add_global_graph_attrs("layout", "dot", "graph") %>%
  # add_global_graph_attrs("rankdir", "LR", "graph") %>% 
  add_nodes_from_table(path_nodes, label_col = var) %>% 
  add_edges_from_table(
    coef_tab, from_col = "Predictor", to_col = "Response", 
    from_to_map = "label"
  ) %>% 
  set_node_attrs(shape, "rectangle") %>% 
  set_node_attrs(width, 3) %>% 
  set_node_attrs(fontsize, 20) %>% 
  # set_node_attrs(width, 1.5) %>% 
  # set_node_attrs(fontsize, 20) %>% 
  # set_edge_attrs(color, coef_tab$pval_col) %>% 
  set_edge_attrs(arrowsize, 1) %>% 
  set_edge_attrs(color, coef_tab$color) %>% 
  set_edge_attrs(style, coef_tab$pval_linetype) %>% 
  set_edge_attrs(penwidth, coef_tab$pval_penwidth) %>% 
  render_graph()


caption = paste(
  "Directed acyclycal graph of selected burden SEM model, fit with maximal cases.",
  "The goodness-of-fit P-value is", paste0(round(selBurdMax$Cstat$P.Value, 2), ","),
  "which indicates that this model is *not* statistically different from an",
  "'ideal' model of these data.",
  "Orange vectors represent positive effects, blue represent negative effects",
  "and grey represents non-significant (P > 0.1) effects.",
  "The width of a vector is proportional to the magnitude of its effect."
)
```

<br>
```{r selBurdMaxSemR2}
selBurdMax$R2 %>% 
  knitr::kable(
    digits = 2,
    caption = paste(
      "R-squared table of the selected SEM, fit with complete cases.",
      "Marginal R-squared values represent the proportion of variation in",
      "the resposne explained by the fixed effects and",
      "Conditional values represent the proportion explained by the full model.",
      "Models without random effects (i.e., simple linear regression) do not",
      "have a conditional R-squared value."
    )
  )
```

### Burden

TBD

### Infection Status

TBD

### Infection-burden hurdle

TBD

## Next steps

* Prediction: how can the expression module (smaller data) inform the larger 
things

  - can we map expression to something else? Proxy variables for expression.
  
  - use 200 mice to fine tune estimation
  
* Maybe look at proposal for machine learning hanta virus / disease?