---
title: "SEMs (continued)"
output: html_document
date: "2025-10-22"
---

```{r}
library(tidyverse)
library(DiagrammeR)
```

# Simplified SEMs

This chapter will build on the insight gained from the previous chapter to 
build the final SEMs for our manuscript. In total, there will be three SEMs.
The first SEM is a "full" model that explores all components of our proposed 
model (mouse traits and behaviors, mouse interactions with disease vectors, 
mouse gene expression, environment, and infection status). To ensure maximum 
power, this full model utilizes data for both *P leucopus* (PELE) and 
*P maniculatus* (PEMA). The two remaining SEMs will explore the specific 
differences in disease dynamics between PELE and PEMA. These models will have
identical structure and will only differ by the species with which they are 
fitted. In order to make comparisons between two identically-structured SEMs, 
some  variables from the full SEM are excluded from these species-specific 
models. This simplification is necessitated by low representation of PEMA for 
some combinations of variables. 
By structuring the analysis in this way, with one overarching model and smaller
species-specific models, we are able to holistically explore the factors
of disease dynamics in this system while also acknowledging and quantifying
differences among species where possible.

## Species identification

Because we are interested in species-specific effects and because 
field assignments are known to be error-prone, we re-assessed
species identification prior to this chapter. Specifically, we used the 
following algorithm to assign species:

1. First, any individuals for which genetic information are available were
assigned to the species indicated by genetic markers.

2. For individuals not covered by step 1, 
discriminant analysis (trained on individuals with genetic assignments) was 
used to assign *observations* for which ear length, hind foot length, and weight
were available. These 3 variables had the highest assignment accuracy for both
species (90% for PELE, 86% for PEMA). Individuals were assigned by 
averaging the predicted species across observations.

3. For individuals not covered by steps 1-2, 
discriminant analysis was used to assign *observations* for which ear length
and weight were available. These two variables also had a high assignment
accuracy for both species (90% for PELE, 84% for PEMA). Individuals were
assigned by averaging the predicted species across observations.

4. Individuals not covered by steps 1-3 that were captured at BLAN, HARV, SERC,
or ORNL were assigned PELE. This was done because genetic assessment identified
no PEMA at any of these sites. **Note**: HARV specifically had extremely poor
field identification, with the vast majority of individuals being labeled as
*Peromyscus spp.* and about half of the individuals identified to species were
(incorrectly) labeled as PEMA.

5. Any remaining individuals were given the field assignment that occurred most 
commonly across their full capture record. 

## Full SEM

The full SEM exploring relationships between mice characteristics and behaviors,
environmental variation, gene expression, and *Borrelia* infection 
was fitted to all available data and is visualized by 
Figure \@ref(fig:simpSemGraph). The model hypothesizes that 1) individual 
traits covary and influence individual behavior, contact with disease vectors 
(tick parasitism), expression of immunity genes, and infection status; 2) 
behaviors covary and influence contact with vectors and 
infection status; 3) tick parasitism influences immunity expression and 
infection status; 4) expression of immunity-associated genes are associated with 
infection status (influenced by resistance and influencing tolerance); and 5)
environmental factors influence all other factors.

Our results show that these hypothesis are largely met:

1.  Mouse phenotype is significantly and directly associated with behavior 
(sex, weight), parasitism rates (sex), resistance expression (sex, weight), 
tolerance expression (reproductive maturity), and infection rates 
(weight, reproductive maturity).

2. Mouse behavior is significantly and directly associated with parasitism 
rates (shift in capture time), but is only indirectly associated with
infection rates (through its influence on tick parasitism).

3. Tick parasitism is significantly and directly associated with resistance
expression and infection status, but is only indirectly associated with 
tolerance expression (through its effect on infection status).

4. Tolerance expression is significantly and directly associated with with
infection status, but resistance expression is not.

5. Weather is significantly and directly associated with mouse phenotype (
reproductive maturity), behavior (capture timing), parasitism rates, immunity
expression (resistance), and infection status. Longer-term climate also has
significant impacts on parasitism rates, likely through its influence on tick
populations.

```{r}
simple_model_objects <- readRDS(
  "infection-modeling/data/model-objects/simplified-SEM-objects.rds"
)

simple_out <- simple_model_objects$sem_mod$fit

## reformat the tables
simple_out$dTable <- simple_out$dTable %>%
  data.frame() %>% rename(sig = "Var.6") %>% tibble()
simple_out$coefficients <- simple_out$coefficients %>%
  data.frame() %>% rename(sig = "Var.9") %>% tibble()
```


```{r}
group_tiers <- c(
  "environment", "trait", "behavior", "parasitism", "expression", "infection"
)

group_cols <- RColorBrewer::brewer.pal(length(group_tiers) + 1, "Pastel2")[-2]

# group_cols <- c(
#   "green", "burlywood", "azure", "grey80", "darkorchid1", "hotpink1"
# )

var_table <- tibble(
  var = c(
    "expr_PC2", "Bb_infected", "expr_PC1", "ticks_attached", "capprop_shift",
    "cap_prop_night", "weight", "sex_mature", 
    "sex_male", "wthr_PC1", "clim_PC1"
  ),
  label = c(
    "Tolerance\nexpression", "Infection\nstatus", "Resistance\nexpression", 
    "Parasitism\n rate (ticks)", "Capture\ntime shift", 
    "Capture\ntime", "Weight", "Reproductive\nmaturity", 
    "Sex", "Weather\nPC", "Climate\nPC"
  ),
  response = c(
    TRUE, TRUE, TRUE, TRUE, TRUE, 
    TRUE, TRUE, TRUE, 
    FALSE, FALSE, FALSE
  ),
  group = c(
    "expression", "infection", "expression", "parasitism", "behavior",
    "behavior", "trait", "trait", 
    "trait", "environment", "environment"
  ) %>% 
    factor(levels = group_tiers)
)
```


```{r, include = FALSE}
scales::pal_area(range = c(0, 1))

pos_col = "cornflowerblue"
neg_col = "orange"

effect_cols <- c(
  pos_col, colorspace::lighten(pos_col, 0.4), colorspace::lighten(pos_col, 0.8),
  neg_col, colorspace::lighten(neg_col, 0.4), colorspace::lighten(neg_col, 0.8)
)

# scales::show_col(effect_cols)
```

```{r}
paths <- simple_out$coefficients %>% 
  select(Response, Predictor, DF, Std.Estimate, P.Value) %>% 
  mutate(
    # pos_effect = Std.Estimate >= 0,
    Response = factor(
      Response, levels = var_table$var, labels = var_table$label
    ),
    Predictor = factor(
      Predictor, levels = var_table$var, labels = var_table$label
    ),
    color = if_else(Std.Estimate >= 0, pos_col, neg_col),
    penwidth = scales::rescale(x = abs(Std.Estimate) ,to = c(1, 9)),
    style = if_else(P.Value <= 0.1, "solid", "dashed"),
    tooltip = round(Std.Estimate, 2)
    # headlabel = round(Std.Estimate, 2),
    # labelfontsize = 14
  )

nodes <- tibble(
  var = unique(
    c(simple_out$coefficients$Response, simple_out$coefficients$Predictor)
  )
) %>% 
  mutate(
    var = factor(var, levels = var_table$var, labels = var_table$label),
    id = row_number(),
    group = var_table$group[match(var, var_table$label)],
    # y = as.numeric(group),
    shape = "rectangle",
    fillcolor = group_cols[match(group, group_tiers)]
  )
```

```{r simpSemGraph, out.width="100%", fig.asp = 1, fig.cap=caption, warning=FALSE}
sem_graph <- create_graph(attr_theme = "bt") %>% 
  # add_global_graph_attrs("layout", "dot", "graph") %>%
  # add_global_graph_attrs("rankdir", "LR", "graph") %>%
  add_nodes_from_table(nodes, label_col = var) %>% 
  add_edges_from_table(
    paths, from_col = Predictor, to_col = Response, from_to_map = label
  ) %>% 
  set_node_attrs(width, 1.4) %>%
  set_node_attrs(height, .7) %>%
  set_node_attrs(fontsize, 16) %>%
  set_node_attrs(fontcolor, "black") %>% 
  set_node_attrs(color, "black")

sem_graph %>% render_graph() 

caption = paste(
  "PATH diagram for the full SEM of *Peromyscus*-*Borrelia* dynamics.",
  "Orange vectors represent negative effects, blue vectors represent positive",
  "effects, solid lines are statistically significant (P \U2264 0.1), and",
  "dashed lines are not. Line widths are proportional to the magnitude of",
  "the standardized coefficient estimates. Boxes are colored according to",
  "variable groupings (blue: phenotype; turquoise: envrionment;",
  "pink: behavior; green: parasitism; yellow: gene expression;",
  "brown: infection.",
  "Each component model used to build the SEM included species, site, and year",
  "as random effects to accout for covariation of observations within and",
  "among these groups.",
  "Table \\@ref(tab:fullSEMcoefTab) shows full model parameters and",
  "Table \\@ref(tab:SemR2N) contains more information about the individual",
  "component models."
)
```

```{r fullSEMcoefTab}
simple_out$coefficients %>% 
  knitr::kable(
    digits = 2,
    caption = "Omnibus SEM statistics."
  )
```


In addition to the direct effects outlined above, it is also possible to
quantify cumulative (direct and indirect) effects from the SEM. The cumulative
effects on parasitism rates, immunity expression, and infection status are
given in Figure \@ref(fig:fullCumuFig).

```{r}
cumu_tab <- readRDS(
  file.path(
    "infection-modeling/data/model-objects",
    "cumulative-effects-table_simplified-SEM.rds"
  )
)

filtab <- cumu_tab %>%
  filter(
    Response %in% c("ticks_attached", "expr_PC1", "Bb_infected", "expr_PC2"),
    !is.na(total), !is.na(Predictor)
  ) %>%
  mutate(
    Response = factor(
      Response,
      levels = c("ticks_attached", "expr_PC1", "Bb_infected", "expr_PC2"),
      labels = c("Ticks", "Resistance", "Infection", "Tolerance")
    ),
    Predictor = factor(
      Predictor,
      levels = c(
        "clim_PC1", "wthr_PC1", "sex_male", "sex_mature", "weight",
        "cap_prop_night", "capprop_shift", "ticks_attached",
        "expr_PC1", "Bb_infected"
      ),
      labels = c(
        "climate", "weather", "sex", "reproductive", "weight",
        "captime", "capshift", "ticks",
        "resistance", "infection"
      )
    )
  )
```

```{r fullCumuFig, fig.width=6, fig.height=6, fig.cap=caption}
# table of effects for Infection, Resistance, and ticks
filtab %>%
  ggplot(aes(y = Predictor, x = total)) +
  facet_wrap(~Response, scales = "free_x") +
  # geom_vline(xintercept = 0, col = "grey50", linetype = "solid") +
  geom_col(
    aes(fill = "total"),
    width = 0.9, linewidth = 0.5, col = "black"
  ) +
  geom_col(
    aes(x = indirect, fill = "indirect"),
    position = position_nudge(y = 0.15),
    width = 0.3, linewidth = 0.25, col = "black"
  ) +
  geom_col(
    aes(x = new_direct, fill = "direct", linetype = P.Value <= 0.1),
    position = position_nudge(y = -0.15),
    width = 0.3, linewidth = 0.25, col = "black"
  ) +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    legend.position = "inside", legend.position.inside = c(0.01, 0.99),
    legend.justification = c(0, 1),
    legend.background = element_rect(color = "black", linewidth = 0.3),
    panel.spacing.x = unit(0, "lines")
  ) +
  labs(x = "Standardized effect", fill = NULL) +
  scale_fill_manual(values = c("cornflowerblue", "orange", "grey50")) +
  scale_linetype_manual(values = c("dashed", "solid")) +
  guides(linetype = "none")

ggsave(
  "infection-modeling/graphics/cumulative-effect-fig.png",
  width = 6, height = 6, dpi = 300
)

caption = paste(
  "Cumulative effects on parasitism (Ticks), immunity expression",
  "(Resistance, Tolerance), and *Borrelia* infection.",
  "These effects are derived from the full SEM (Figure \\@ref(fig:simpSemGraph)).",
  "Dashed borders around blue bars indicate non-significant direct effects."
)
```

```{r SemR2N}
mod_list <- simple_model_objects$component_mods

sample_sizes <- lapply(mod_list, function(x)nrow(model.frame(x))) %>% unlist()

bind_cols(
  simple_out$R2 %>% 
    mutate(
      Response = factor(
        Response, levels = var_table$var, 
        labels = var_table$label %>% gsub("\n", " ", .)
      )
    ) %>% 
    select(-method), 
  n = sample_sizes
) %>% knitr::kable(
  col.names = c("Response", "Family", "Link", "Marg. R-sqr", "Cond. R-sqr", "N"),
  digits = 2,
  caption = paste(
    "Full SEM component model information. Columns represent response,",
    "distribution family, distribution link function, marginal R-squared",
    "(fixed effects only), conditional R-squared (fixed and random effects),",
    "and the total number of observations used for each component model.",
    "For binomial models, R-squared values are theoretical."
  )
)
```


```{r allCumuRough, fig.width = 6, fig.asp = 1, include = FALSE}
# look at all the effects
cumu_tab %>%
  ggplot(aes(y = Predictor, x = total)) +
  facet_wrap(~Response, scales = "free_x") +
  geom_vline(xintercept = 0, col = "grey50", linetype = "dotted") +
  geom_col(
    aes(fill = "total"),
    width = 0.9, linewidth = 0.5, col = "black"
  ) + geom_col(
    aes(x = indirect, fill = "indirect"),
    position = position_nudge(y = 0.15),
    width = 0.3, linewidth = 0.25, col = "black"
  ) +
  geom_col(
    aes(x = new_direct, fill = "direct", linetype = P.Value <= 0.1),
    position = position_nudge(y = -0.15),
    width = 0.3, linewidth = 0.25, col = "black"
  ) + 
  theme_bw() +
  theme(
    strip.background = element_blank(),
    legend.position = "inside", legend.position.inside = c(0.95, 0.05),
    legend.justification = c(1, 0),
    # legend.background = element_rect(color = "black", linewidth = 0.3),
    panel.spacing.x = unit(0, "lines")
  ) +
  labs(x = "Standardized effect", fill = NULL) +
  scale_fill_manual(values = c("cornflowerblue", "orange", "grey50")) +
  scale_linetype_manual(values = c("dashed", "solid")) +
  guides(linetype = "none")
```

## Species comparison




```{r}
pele_model_objects <- readRDS(
  "infection-modeling/data/model-objects/PELE-SEM-objects.rds"
)
pele_sem <- pele_model_objects$sem_mod$fit
pema_model_objects <- readRDS(
  "infection-modeling/data/model-objects/PEMA-SEM-objects.rds"
)
pema_sem <- pema_model_objects$sem_mod$fit

spec_coefs <- readRDS(
  "infection-modeling/data/model-objects/joint-species-coefficients_SEM.rds"
)

sp_cumeffs <- readRDS(
  file.path(
    "infection-modeling/data/model-objects",
    "species_cumulative-effects-table_simplified-SEM.rds"
  )
)
```

```{r PelePemaSem, out.width="100%", fig.asp = 1, fig.cap=caption, warning=FALSE}
pele_fp <- frameplot_psem(pele_sem$coefficients)
pema_fp <- frameplot_psem(pema_sem$coefficients)

format_edge_lab <- function(x, y){
  x.small = round(x, 2)
  y.small = round(y, 2)
  paste0(x.small, " (", y.small, ")")
}

spec_unif_coefs <- spec_coefs %>%
  select(Response, Predictor, species, CI_overlap, Std.Estimate, P.Value) %>% 
  pivot_wider(names_from = species, values_from = c(Std.Estimate, P.Value)) %>% 
  mutate(
    opposite_signs = if_else(
      Std.Estimate_PEMA * Std.Estimate_PELE < 0, TRUE, FALSE
    ), 
    label = format_edge_lab(Std.Estimate_PELE, Std.Estimate_PEMA),
    tooltip = label,
    style = if_else(P.Value_PELE <= 0.1, "solid", "dashed"),
    color = case_when(
      CI_overlap & Std.Estimate_PELE >= 0 ~ "cornflowerblue",
      CI_overlap & Std.Estimate_PELE < 0 ~ "orange",
      !CI_overlap & !opposite_signs & Std.Estimate_PELE >= 0 ~ "cornflowerblue",
      !CI_overlap & !opposite_signs & Std.Estimate_PELE < 0 ~ "orange",
      !CI_overlap & opposite_signs ~ "purple"
    ),
    fontcolor = color,
  ) %>% ungroup()

pele_fp$edges <- spec_unif_coefs %>% 
  mutate(
    Response = factor(
      Response, levels = var_table$var, labels = var_table$label
    ),
    Predictor = factor(
      Predictor, levels = var_table$var, labels = var_table$label
    ),
  )

pele_fp$nodes <- pele_fp$nodes %>% 
  mutate(
    name = factor(name, levels = var_table$var, labels = var_table$label)
  )


build_psem_plot(pele_fp) %>% 
  set_node_attrs("shape", "rectangle") %>% 
  set_node_attrs("width", 1.1) %>% 
  set_node_attrs("height", 0.5) %>% 
  render_graph()

caption = paste(
  "Path diagram comparing a simplified SEM between *P. leucopus* and",
  "*P. maniculatus*. Blue vectors indicate positive effects, orange indicate",
  "negative effects, purple indicate conflicting effect directions between",
  "the two species, solid lines are statistically significant (for *leucopus*)",
  "and dotted lines are not.",
  "Path labels represent standardized effect sizes for both species, with",
  "*P. maniculatus* in parentheses.",
  "See Table \\@ref(tab:spCoefTab) for more model differences and",
  "Figure \\@ref(fig:spCumInf) for the cumulative effects on infection for each",
  "species."
)
```

Because relatively few *P. maniculatus* were observed with a combination of
genetic, behavioral, and infection information, to compare PELE and PEMA with
identical models required simplification. Specificially, we removed behavior 
and immunity expression for the comparative SEM.

With this comparison, we found that the model differed significantly between
PELE and PEMA (Figure \@ref(fig:PelePemaSem)) and our general assumptions were 
only met for PELE and not PEMA due to no strong link between parasitism and 
infection among PEMA.

Specific hypotheses:

1. Mouse phenotype is significantly associated with parasitism via sex among
PELE but not PEMA and via weight and reproductive maturity among PEMA but not 
PELE. Phenotype is also linked to infection status via reproductive maturity
and weight in both species (though, PEMA's weight effects are opposite PELE's)
and via Sex among PEMA only.

2. NA - behavior not assessed

3. Tick parasitism is directly associated with infection status 
**only among PELE**

4. NA - immunity not assessed

5. Weather significantly impacted phenotype, parasistism, and infection whereas
long-term climate only impacted parasitism rates directly, as in the full model.

This means that, while both species exhibit intraspecific variation in
disease competence, interaction with ticks is less important for PEMA than for
PELE. This could indicate that PEMA is more resistant to infection than PELE. 


```{r spCoefTab}
spec_coefs %>% 
  select(-c(upper_low, lower_high, CI_overlap)) %>% 
  arrange(Response, Predictor, species) %>% 
  relocate(species, .after = Predictor) %>% 
  mutate(DF = as.integer(DF)) %>% 
  knitr::kable(
    caption = "Species-specific SEM",
    digits = 2
  )
```



```{r spCumInf, fig.cap=caption}
# TODO: move to .Rmd
## visualize
sp_cumeffs %>%
  filter(Response == "Bb_infected") %>%
  ggplot(aes(y = Predictor)) +
  geom_vline(xintercept = 0, linetype = "solid", color = "grey50") +
  facet_wrap(~Species) +
  geom_col(aes(x = total, fill = "total"), col = "black") +
  geom_col(
    aes(x = direct, fill = "direct"), width = 0.4, col = "black",
    position = position_nudge(y = 0.2)
    ) +
  geom_col(
    aes(x = indirect, fill = "indirect"), width = 0.4, col = "black",
    position = position_nudge(y = -0.2)
  ) +
  theme_bw() +
  labs(x = "Effect on infection status (standardized)")

caption = paste(
  "Cumulative standardized effects on infection status for *P. leucopus* PELE",
  "and *P. maniculatus* (PEMA)."
)
```



```{r}
# TODO: move to .Rmd
# ## Plot the coefficients by species (solid lines and bold colors are
# ## significantly different)
# spec_coefs %>%
#   mutate(
#     Response = factor(
#       Response,
#       levels = c("sex_mature", "weight", "ticks_attached", "Bb_infected"),
#       labels = c("Reproductive", "Weight", "Ticks", "Infection")
#     ),
#     Predictor = factor(
#       Predictor,
#       levels = c(
#         "wthr_PC1", "clim_PC1", "sex_male", "sex_mature",
#         "weight", "ticks_attached"
#       ),
#       labels = c("Weather", "Climate", "Sex", "Reproductive", "Weight", "Ticks")
#     )
#   ) %>%
#   ggplot(
#     aes(
#       x = Std.Estimate, y = Predictor, fill = species, linetype = CI_overlap,
#       alpha = P.Value <= 0.1
#     )
#   ) +
#   geom_vline(xintercept = 0, col = "grey50") +
#   facet_grid(~Response) +
#   geom_col(position = "dodge", color = "black", width = 0.66) +
#   theme_bw() +
#   theme(
#     legend.position = "inside", legend.position.inside = c(.22, 0.75),
#     legend.justification = c(0.5, 0.5),
#     legend.background = element_rect(color = "black", linewidth = 0.2),
#     panel.spacing.x = unit(0, "lines"),
#     strip.background = element_blank()
#   ) +
#   labs(fill = NULL, x = "Standardized effect") +
#   scale_linetype_manual(values = c("solid", "dashed")) +
#   scale_alpha_manual(values = c(0.1, 1)) +
#   guides(linetype = FALSE, alpha = FALSE)

```
