---
title: "Additional NEON Peromyscus models"
execute:
  echo: true
format:
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
---

## John's Email

Hey Clay,

Good to zoom just now – I feel like we’ve got a solid plan and I’m looking forward to getting this finished!

Here are a list of my quick stats thoughts – I might send along some more by Friday, but wanted to send you this initial list just to get a sense of what we might want to check out.

1. Pele model – not sure if it’ll become our primary model for the ms, but I think we’ll definitely want to develop one to the same extent we’ve developed the current Pele/pema model. Do we have a pele-only model that also has the bootstrapped estimates? 

2. In Marty’s comments, he mentioned the Bb burden data – I vaguely recall talking about this, but did we ever explore using it in the model (presumably as a response variable, possibly in a hurdle-like fashion?).  For example, it would be interesting if ticks were positively associated with infection, but size (or reproductive status) was correlated with Bb burden after infection

3. Small tidbit – in Table 1, the infection model looks like we might want to check it out since the parameter estimates look strange for resistance (coeff=0, p=1).  Maybe there was a conversion issue, something strange about the input data, something funky with variable standardization, or something else that will help us resolve the funky parameter estimates?

4. Can we check on the models where we tried the space-use data (avg distance moved, trap diversity, etc.) to see what the results looked like (just to provide more details on why we don’t include them)?  In the Brehm et al. 2025 paper, the metric that really stood out was average distance moved (in explaining parasitism), so I want to make sure we feel like we’re solid in terms of exploring it in the current pSEM analysis.  In case it’s useful to have it in one spot, I’ll paste the description of how Alli and I calculated the metrics below just so it’s easy to find.

5. I know it’ll complicate our figure legends/labels, but I think it’ll be useful to have a label for climate that reminds readers that it (unlike weather) also includes variability.  And, as a final triple-check, can you send along the values that we’re using for climate just so we can confirm that they are the same as in the other papers (this will also be important because we’ll probably have a few sentences discussing those papers)?

Let me know what you think about how long those things might take – in the meantime, I’ll devote more thought to any other statistical things we might want to consider!

Sincerely,

John

## Setup

```{r}
#| results: "hide"
library(tidyverse)
library(lme4)
library(lmerTest)
library(piecewiseSEM)
```


```{r}
# construct path to model data ojects
model_path = file.path(
  "infection-modeling/data/model-objects/unscaled-peros_SEM-objects.rds"
)

# read the objects
fullSEM_model_objects <- readRDS(model_path)

# extract list of component modles
modlist <- fullSEM_model_objects$component_mods

# extract the fit object
full_psem <- fullSEM_model_objects$sem_mod$fit

# Load raw data
Peros <- readRDS("infection-modeling/data/peromyscus-model-data.rds") %>%
  mutate(species = updated_taxa) %>% 
  filter(!is.na(species)) %>% 
  mutate(
    across(c(siteID, plotID, year, iid, species), ~as.factor(.x)) # factor vars
  )
```

## 1. PELE Only

As we've discussed before, the PELE-only model is extremely similar to the 
model that includes PEMA. This is entirely unsurprising, since the `species`
random effect accounts for the between-species variation and is one of the reasons
we decided to use the full model that included them both. @tbl-PELE-Cstat shows
the adequate goodness-of-fit statistic, @tab-PELE-sem shows the coefficients,
and @tbl-PELE-R2 shows the $R^2$ values. @fig-PELE-full-compare shows a compact
comparison of the two models.

I can run the bootstrap on the PELE-only model, if this is the one you'd like
to use.

```{r}
#| results: "hide"

# Update each component mod by removing the random effect os species and
## filtering the data to only include PELE
pele_mods <- lapply(modlist, function(x){
  update(x, . ~ . -(1|species), data = Peros %>% filter(species == "PELE"))
})

PELE_semSpec <- piecewiseSEM::as.psem(pele_mods)
PELE_psem <- summary(PELE_semSpec)
```

```{r}
#| label: "tbl-PELE-Cstat"
PELE_psem$Cstat
```

```{r}
#| label: "tbl-PELE-sem"
PELE_psem$coefficients %>% data.frame() %>% select(-Var.9) %>% 
  mutate(sig = ifelse(P.Value <= 0.1, "*", ""))
```

```{r}
#| label: "tbl-PELE-R2"
PELE_psem$R2
```

```{r}
#| echo: false
#| label: "fig-PELE-full-compare"
#| fig-cap: "Comparison of PELE/PEMA model to PELE-only model coefficients."
#| fig-width: 6.5
#| fig-asp: 0.8
PELE_psem$coefficients %>% data.frame() %>% 
  mutate(mod = "pele-only") %>% 
  full_join(
    full_psem$coefficients %>% data.frame() %>% mutate(mod = "both-spec")
  ) %>% 
  mutate(
    sig = P.Value <= 0.1, group = interaction(Response,Predictor),
    group = forcats::fct_reorder(group, Std.Estimate)
  ) %>% 
  ggplot(aes(x = group, y = Std.Estimate, shape = sig, col = mod)) +
  geom_hline(yintercept = 0, linetype = "dashed", col = "grey50") + 
  geom_point(size = 3) + geom_line(aes(group = mod)) +
  theme_bw() +
  theme(
    legend.position.inside = c(0.1, .9), legend.position = "inside",
    legend.justification = c(0, 1), legend.direction = "horizontal",
    axis.text.x = element_text(angle = 90)
  ) +
  labs(
    x = "Ordered coefficient (Response.Predictor)", y = "Coefficient Value",
    color = "Model", shape = "P < 0.1"
  ) +
  scale_shape_manual(values = c(1, 16))
```


## 2. Burden / Hurdle

As we've discussed before, burden was excluded because it added very little to the
model, while making interpretation more difficult (hurdle SEMs don't work).
@tlb-burden-fit shows the coefficients table for the effects on log-burden 
(PELE-only here, but true for both models). Note that burden only
differs by reproductive maturity. Similarly, @fig-tol-burd shows that 
tolerance is even less affected by burden than by infection status.

If you want me to add burden it to the SEM, we have to include infection
status as a predictor of burden for the model to fit (interpretation is more
difficult).

```{r}
#| label: "tbl-burden-fit"
burden_pele <- modlist$infection %>% formula() %>% 
  update(log_burden ~ . + cap_prop_night - (1|species)) %>% 
  lmer(data = Peros %>% filter(Bb_infected == 1, species == "PELE"))

summary(burden_pele)$coefficients
```

```{r}
#| label: "fig-tol-burd"
tol_burd <- modlist$tolerance %>% 
  update(
    . ~ . + log_burden - (1|species), 
    data = Peros %>% filter(species == "PELE")
  )

summary(tol_burd)$coefficient
```

```{r}
burden_inf = burden_pele %>% 
  update(
    . ~ Bb_infected + ., 
    data = Peros %>% filter(species == "PELE")
  )

summary(burden_inf)$coefficients
```


```{r}
#| results: "hide"
add_burden_spec = piecewiseSEM::as.psem(
  append(pele_mods, list(burden = burden_inf)),
)
sem_with_burden = summary(add_burden_spec)
```

```{r}
sem_with_burden$coefficients %>% data.frame() %>% select(-Var.9) %>% 
  mutate(sig = ifelse(P.Value <= 0.1, "*", ""))
```


## 3. Infection model resistance parameters

The values for the effect of resistance (expr_PC1) on infection status that you
reference (β = 0, P = 1) from Table 1 are correct but subject to 
rounding and the significant digits of that table. The true values are 
β = 0.0002 and P = 0.9997 (@tbl-resinf-coef).

```{r}
#| label: tbl-resinf-coef
summary(modlist$infection)$coefficients
```


## 4. Space use data

The reason we decided to not use our 3 space-use metrics (weighted trap 
diversity, weighted trapability, and average move distance between 
recaptures) was because they were not important for *Borrelia*
infections (or burden). @tble-space-use shows that the only things affected
by these space-use metrics were tolerance (`expr_PC2`) and capture time 
(`cap_prop_night`). Additionally, inclusion of space-use metrics obscured the
effects of weight on infection status 
(fig-space-infect vs. fig-original-infect).

Note that these results that contain space-use are from an older version of the 
full PELE/PEMA that has slightly different parameteterization, 
but if you decide you want these added to the PELE-only model, that can 
be accommodated. Because these models are so similar, I don't expect any 
different conclusions to arise.

```{r}
#| label: "tbl-space-use"
#| fig-cap: >- 
#|   Effects of space use in Peromyscus SEM (PELE and PEMA). This is a subset
#|   of the SEM coefficient table that only includes rows where the predictor
#|   variable is a space-use term.

space_sem = readRDS(
  "infection-modeling/data/model-objects/full-infection-SEM_maximal-cases.rds"
)

space_sem$coefficients %>% data.frame() %>% select(-Var.9) %>% 
  mutate(sig = ifelse(P.Value <= 0.1, "*", "")) %>% 
  filter(
    grepl("trap|move", Predictor), 
    # P.Value <= 0.1
  )
```

```{r}
#| label: "fig-space-infect"
space_sem$coefficients %>% data.frame() %>% select(-Var.9) %>% 
  mutate(sig = ifelse(P.Value <= 0.1, "*", "")) %>% 
  filter(Response == "Bb_infected")
```

```{r}
#| label: "fig-original-infect"
full_psem$coefficients %>% data.frame() %>% select(-Var.9) %>% 
  mutate(sig = ifelse(P.Value <= 0.1, "*", "")) %>% 
  filter(Response == "Bb_infected")
```

## 5. Climate PCA

I'm happy to amend any of the figures with whatever details you'd like. Just
let me know the specifics. As for climate PCA, they were constructed 
with average monthly temperature, precipitation, and relative humidity; 
coefficients of variation on those monthly averages; and the average 
proportion of days in a month on which some precipitation occurred (this 
variable is novel compared to Brehm 2025 and was included to better capture
whether precip. tended to be temporally concentrated see @fig-rainday-compare
for the comparisons). @tbl-clim-pc shows the PC loadings for each site based
on these climate values and @fig-clim-pca shows the variables in PC space.
Note that we use the inverse of the first PC in our
models, to align it with the interpretation of the weather PC (i.e., 
positive = warmer and wetter).

```{r}
#| label: "tbl-clim-pca-data"
clim_pca <- readRDS("infection-modeling/data/climate-PCA.rds")

clim_pca$data
```

```{r}
#| fig-asp: 1
#| fig-width: 3.5 
#| layout-ncol: 2
#| label: "fig-rainday-compare"
clim_pca$data %>% 
  ggplot(
    aes(x = site_pct_monthly_raindays, y = site_CV_monthly_precip, col = siteID)
  ) +
  geom_point(size = 3, show.legend = FALSE) + theme_bw()

clim_pca$data %>% 
  ggplot(
    aes(x = site_pct_monthly_raindays, y = site_avg_monthly_precip, col = siteID)
  ) +
  geom_point(size = 3) + theme_bw() +
  theme(
    legend.position = "inside", legend.position.inside = c(0.99, 0.01), 
    legend.justification = c(1, 0), legend.background = 
      element_rect(color = "grey50")
  )
```

```{r}
#| label: "tbl-clim-pc"
Peros %>% 
  group_by(siteID) %>% 
  select(siteID, clim_PC1, clim_PC2) %>% distinct()
```

```{r}
#| label: "fig-clim-pca"
ordiplot(clim_pca$pca_obj, type = "text", display = "sp")
```

