---
title: "Additional NEON Peromyscus models"
execute:
  echo: false
  message: false
  warning: false
  collapse: true
format:
  pdf:
    toc: false
    number-sections: false
    colorlinks: true
    fontsize: "8"
    df-print: kable
    fig-pos: "h"
---

<!-- ## Setup -->

```{r}
#| output: false
library(tidyverse)
library(vegan)
library(lme4)
library(lmerTest)
library(piecewiseSEM)
```


```{r}
#| output: false

# construct path to model data ojects
model_path = file.path(
  "infection-modeling/data/model-objects/unscaled-peros_SEM-objects.rds"
)

# read the objects
fullSEM_model_objects <- readRDS(model_path)

# extract list of component modles
modlist <- fullSEM_model_objects$component_mods

# extract the fit object
full_psem <- fullSEM_model_objects$sem_mod$fit

# Load raw data
Peros <- readRDS("infection-modeling/data/peromyscus-model-data.rds") %>%
  mutate(species = updated_taxa) %>% 
  filter(!is.na(species)) %>% 
  mutate(
    across(c(siteID, plotID, year, iid, species), ~as.factor(.x)) # factor vars
  )
```

## PELE Only

> *Pele model – not sure if it’ll become our primary model for the ms, but 
> I think we’ll definitely want to develop one to the same extent we’ve 
> developed the current Pele/pema model. Do we have a pele-only model that 
> also has the bootstrapped estimates?* 

As we've discussed before, the PELE-only model is extremely similar to the 
model that includes PEMA. This is entirely unsurprising, since the `species`
random effect accounts for the between-species variation and is one of the 
reasons we decided to use the full model that included them both. 
@tbl-PELE-Cstat shows the adequate goodness-of-fit statistic, @tbl-PELE-sem 
shows the coefficients, and @tbl-PELE-R2 shows the $R^2$ values. 
@fig-PELE-full-compare shows a compact comparison of the two models.

I can run the bootstrap on the PELE-only model, if this is the one you'd like
to use.

```{r}
#| output: false

# Update each component mod by removing random species effect, only keep PELE
pele_mods <- lapply(modlist, function(x){
  update(x, . ~ . -(1|species), data = Peros %>% filter(species == "PELE"))
})

# construct and fit pSEM
PELE_semSpec <- piecewiseSEM::as.psem(pele_mods)
PELE_psem <- summary(PELE_semSpec)
```

```{r}
#| label: "tbl-PELE-Cstat"
PELE_psem$Cstat
```

```{r}
#| label: "tbl-PELE-sem"
PELE_psem$coefficients %>% data.frame() %>% select(-c(Var.9, Crit.Value)) %>% 
  mutate(sig = ifelse(P.Value <= 0.1, "*", ""), DF = round(DF))
```

```{r}
#| label: "tbl-PELE-R2"
PELE_psem$R2
```

```{r}
#| echo: false
#| label: "fig-PELE-full-compare"
#| fig-cap: "Comparison of PELE/PEMA model to PELE-only model coefficients."
#| fig-width: 6.5
#| fig-asp: 0.8
PELE_psem$coefficients %>% data.frame() %>% 
  mutate(mod = "pele-only") %>% 
  full_join(
    full_psem$coefficients %>% data.frame() %>% mutate(mod = "both-spec"),
    join_by(
      Response, Predictor, Estimate, Std.Error, DF, Crit.Value,
      P.Value, Std.Estimate, Var.9, mod
    )
  ) %>% 
  mutate(
    sig = P.Value <= 0.1, group = interaction(Response,Predictor),
    group = forcats::fct_reorder(group, Std.Estimate)
  ) %>% 
  ggplot(aes(x = group, y = Std.Estimate, shape = sig, col = mod)) +
  geom_hline(yintercept = 0, linetype = "dashed", col = "grey50") + 
  geom_point(size = 3) + geom_line(aes(group = mod)) +
  theme_bw() +
  theme(
    legend.position.inside = c(0.1, .9), legend.position = "inside",
    legend.justification = c(0, 1), legend.direction = "horizontal",
    axis.text.x = element_text(angle = 90)
  ) +
  labs(
    x = "Ordered coefficient (Response.Predictor)", y = "Coefficient Value",
    color = "Model", shape = "P < 0.1"
  ) +
  scale_shape_manual(values = c(1, 16))
```


## Burden / Hurdle

> *In Marty’s comments, he mentioned the Bb burden data – I vaguely recall 
> talking about this, but did we ever explore using it in the model (presumably 
> as a response variable, possibly in a hurdle-like fashion?).  For example, 
> it would be interesting if ticks were positively associated with infection, 
> but size (or reproductive status) was correlated with Bb burden after 
> infection*

As we've discussed before, burden was excluded because it added very little to 
the model, while making interpretation more difficult (hurdle SEMs don't work,
unfortunately).
@tbl-burden-fit shows the coefficients table for the effects on log-burden 
(PELE-only here, but true for both models). Note that burden only
differs by reproductive maturity. Similarly, @tbl-tol-burd shows that 
tolerance is even less affected by burden than by infection status.

If you want me to add burden to the SEM, we have to include infection
status as a predictor of burden for the model to fit (i.e., @tbl-burd-plus-inf,
@tbl-burd-plus-inf), 
which is legitimate, but makes interpretation is more difficult 
("duh, of course infection predicts burden because they are confounded"). In 
this case, we look to the other coefficients as explaining additional variance
in burden beyond what infection status already explains: reproductive maturity
explains some, for example (@tbl-burd-plus-inf).

```{r}
#| label: "tbl-burden-fit"
#| tbl-cap: "Effects on burden (PELE SEM component model)"
#| output: false

burden_pele <- modlist$infection %>% formula() %>% 
  update(log_burden ~ . + cap_prop_night - (1|species)) %>% 
  lmer(data = Peros %>% filter(Bb_infected == 1, species == "PELE"))

summary(burden_pele)$coefficients
```

```{r}
#| label: "tbl-tol-burd"
#| tbl-cap: "Effects on PELE tolerance (including infection status and burden)"
tol_burd <- modlist$tolerance %>% 
  update(
    . ~ . + log_burden - (1|species), 
    data = Peros %>% filter(species == "PELE")
  )

summary(tol_burd)$coefficient
```

```{r}
#| label: "tbl-burd-plus-inf"
#| tbl-cap: >- 
#|   Effects on burden, including infection status
#|   (PELE SEM component model)
burden_inf = burden_pele %>% 
  update(
    . ~ Bb_infected + ., 
    data = Peros %>% filter(species == "PELE")
  )

summary(burden_inf)$coefficients
```


```{r}
#| output: false
add_burden_spec = piecewiseSEM::as.psem(
  append(pele_mods, list(burden = burden_inf)),
)
sem_with_burden = summary(add_burden_spec)
```

```{r}
#| label: "tbl-sem-with-burden"
#| tbl-cap: "SEM coefficient table for PELE, with burden added."
sem_with_burden$coefficients %>% data.frame() %>% select(-Var.9, -Crit.Value) %>% 
  mutate(sig = ifelse(P.Value <= 0.1, "*", ""), DF = round(DF))
```

## Infection model resistance parameters

> *Small tidbit – in Table 1, the infection model looks like we might want 
> to check it out since the parameter estimates look strange for resistance 
> (coeff=0, p=1).  Maybe there was a conversion issue, something strange 
> about the input data, something funky with variable standardization, or 
> something else that will help us resolve the funky parameter estimates?*

The values for the effect of resistance (expr_PC1) on infection status that you
reference (β = 0, P = 1) from Table 1 are correct but subject to 
rounding and the significant digits of that table. The true values are 
β = 0.0002 and P = 0.9997 (@tbl-resinf-coef).

```{r}
#| label: "tbl-resinf-coef"
#| tbl-cap: "model of infection status (current PELE/PEMA SEM)"
summary(modlist$infection)$coefficients
```

## Space use data

> *Can we check on the models where we tried the space-use data (avg
> distance moved, trap diversity, etc.) to see what the results looked like
> (just to provide more details on why we don’t include them)?  In the
> Brehm et al. 2025 paper, the metric that really stood out was average
> distance moved (in explaining parasitism), so I want to make sure we feel
> like we’re solid in terms of exploring it in the current pSEM analysis.
> In case it’s useful to have it in one spot, I’ll paste the description
> of how Alli and I calculated the metrics below just so it’s easy to find.*

The reason we decided to not use our 3 space-use metrics (weighted trap 
diversity, weighted trapability, and average move distance between 
recaptures) was because they were not important for *Borrelia*
infections (or burden) or tick parasitism. This is true whether or not we 
standardize by number of captures (I strongly believe it is more appropriate in
this context to use the standardized values, though).
@tbl-space-use shows that the only things affected by these space-use metrics 
were tolerance (`expr_PC2`) and capture time (`cap_prop_night`). Additionally, 
inclusion of space-use metrics obscured the effects of weight on infection 
status (@tbl-space-infect vs. @tbl-original-infect).

Note that these results that contain space-use are from an older version of the 
full PELE/PEMA that has slightly different parameteterization, 
but if you decide you want these added to the PELE-only model, that can 
be accommodated. Because these models are so similar, I don't expect any 
different conclusions to arise.

```{r}
#| label: "tbl-space-use"
#| tbl-cap: >- 
#|   Effects of space use in Peromyscus SEM (PELE and PEMA). This is a subset
#|   of the SEM coefficient table that only includes rows where the predictor
#|   variable is a space-use term.

space_sem = readRDS(
  "infection-modeling/data/model-objects/full-infection-SEM_maximal-cases.rds"
)

space_sem$coefficients %>% data.frame() %>% 
  select(-Var.9, -Crit.Value, -Estimate, -Std.Error, -DF) %>% 
  mutate(sig = ifelse(P.Value <= 0.1, "*", "")) %>% 
  filter(
    grepl("trap|move", Predictor), 
    # P.Value <= 0.1
  )
```

```{r}
#| label: "tbl-space-infect"
#| tbl-cap: >- 
#|   Standardized effects of factors affecting infection status 
#|   (old PELE/PEMA SEM, including space use)
space_sem$coefficients %>% data.frame() %>% 
  select(-Var.9, -Crit.Value, -DF, -Estimate, -Std.Error) %>% 
  mutate(sig = ifelse(P.Value <= 0.1, "*", "")) %>% 
  filter(Response == "Bb_infected")
```

```{r}
#| label: "tbl-original-infect"
#| tbl-cap: >-
#|   Standardized effects of factors affecting infection status 
#|   (current PELE/PEMA SEM)
full_psem$coefficients %>% data.frame() %>% 
  select(-Var.9, -Crit.Value, -DF, -Estimate, -Std.Error) %>% 
  mutate(sig = ifelse(P.Value <= 0.1, "*", "")) %>% 
  filter(Response == "Bb_infected")
```

## Climate PCA

> *I know it’ll complicate our figure legends/labels, but I think it’ll be 
> useful to have a label for climate that reminds readers that it 
> (unlike weather) also includes variability.  And, as a final triple-check, 
> can you send along the values that we’re using for climate just so we can 
> confirm that they are the same as in the other papers (this will also be 
> important because we’ll probably have a few sentences discussing those 
> papers)?*

I'm happy to amend any of the figures with whatever details you'd like. Just
let me know the specifics. As for climate PCA, they were constructed 
with average monthly temperature, precipitation, and relative humidity; 
coefficients of variation on those monthly averages; and the average 
proportion of days in a month on which some precipitation occurred (this 
variable is novel compared to Brehm 2025 and was included to better capture
whether precip. tended to be temporally concentrated see @fig-rainday-compare-1 
and @fig-rainday-compare-1 for the comparisons). 
@tbl-clim-pc shows the PC loadings for each site based
on these climate values and @fig-clim-pca shows the variables in PC space.
Note that we use the inverse of the first PC in our
models, to align it with the interpretation of the weather PC (i.e., 
positive = warmer and wetter).

```{r}
#| label: "tbl-clim-pca-data"
#| tbl-cap: "Site average monthly climate data"
clim_pca <- readRDS("infection-modeling/data/climate-PCA.rds")

clim_pca$data %>% 
  setNames(gsub("site_(.*)_monthly_(.*)", "\\1_\\2", names(clim_pca$data)))
```

```{r}
#| fig-asp: 1
#| fig-width: 3.5 
#| layout-ncol: 2
#| label: "fig-rainday-compare"
#| fig-cap: 
#|  - "Rain days vs. precip CV"
#|  - "Rain days vs. avg precip"
clim_pca$data %>% 
  ggplot(
    aes(x = site_pct_monthly_raindays, y = site_CV_monthly_precip, col = siteID)
  ) +
  geom_point(size = 3, show.legend = FALSE) + theme_bw()

clim_pca$data %>% 
  ggplot(
    aes(x = site_pct_monthly_raindays, y = site_avg_monthly_precip, col = siteID)
  ) +
  geom_point(size = 3) + theme_bw() +
  theme(
    legend.position = "inside", legend.position.inside = c(0.99, 0.01), 
    legend.justification = c(1, 0), legend.background = 
      element_rect(color = "grey50")
  )
```

```{r}
#| label: "tbl-clim-pc"
#| tbl-cap: "Climate PC loadings"
Peros %>% 
  group_by(siteID) %>% 
  select(siteID, clim_PC1, clim_PC2) %>% distinct()
```

```{r}
#| label: "fig-clim-pca"
#| fig-cap: "Climate PC space"
ordiplot(clim_pca$pca_obj, type = "text", display = "sp")
```

