---
params:
  force: FALSE
---

```{r knit_setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, error = FALSE)
```

```{r cache_setup, include=FALSE}
cache_dir = "infection-modeling/R/rmarkdown/cache/resistance-tolerance"

if (!dir.exists(cache_dir)) {
  dir.create(cache_dir, recursive = TRUE)
}

knitr::opts_chunk$set(cache.path = R.utils::getAbsolutePath(cache_dir))

# function to run/cache objects
run_cache <- function(expr, rds_path, force = params$force){
  
  if (force || !file.exists(rds_path)) {
    x = eval(expr)
    saveRDS(x, rds_path)
  } else {
    x = readRDS(rds_path)
  }
  
  return(x)
}
```


```{r pkgs, message=FALSE}
library(tidyverse)
library(glmmTMB)
library(lme4)
```

# Resistance and Tolerance {-}

## Introduction

This document will track the analysis of the resistance to and tolerance of
*Borrelia* (B.b.) in *Peromyscus* mice at NEON.

## Data

First, we'll load in the data from Vania and rename a few of the columns.

```{r resTolData, echo = TRUE}
# read the excel file containing immune + infection data
df <- readxl::read_excel(
  "infection-modeling/data/Immune and burden Table.xlsx", na = c("", "NA")
) %>% tibble()

# clean it and keep only needed columns
df <- df %>% 
  select(
    # Linking ID variables
    plotID, tagID = tagID.x, collectDate, trapCoordinate, 
    # updated taxon information
    DFA_Taxon = `DFA ID`, Gel_Taxon = `Gel ID`, 
    # geneetic info
    genetics_ID = `RNA/DNA ID #`, 
    Bb_burden = `Bb burden (OspA/10kRpp30)`, Bb_status = `Bb Pos/Neg`,
    `IL-10`:`GATA-3`,
    # individual traits
  ) %>% distinct() %>% 
  relocate(`TGF-B`, .before = `GATA-3`)
```

Then we'll merge these data with the small mammal trapping data from NEON.

```{r mammals, echo = TRUE}
# read mammal trap data
mammals <- readRDS("infection-modeling/data/full-data.rds")
# count the trapping observations
n_mammals = nrow(mammals)
```

```{r unmatched, include = FALSE}
# rows in mammals that aren't in df
unmatched <- anti_join(df, mammals)

# save the unmatched as a table
if (FALSE) {
  unmatched %>% arrange(tagID) %>%
    write.csv(file = "infection-modeling/data/unmatched_rows_immunteTable.csv")
}

# get the rows which are duplicated
duplicates <- df %>% 
  group_by(tagID) %>% 
  tally() %>% 
  full_join(df) %>% 
  filter(n > 1) %>% 
  arrange(tagID, plotID, ymd(collectDate), trapCoordinate) %>% 
  select(-n)

# save the duplicates as a table
if (FALSE) {
  duplicates %>%
    write.csv(file = "infection-modeling/data/duplicated_tagIDs_immuneTable.csv")
}

# remove the duplicated sample (with two slightly different Bb_burdens)
df <- df %>% 
  filter( !(tagID == "NEON.MAM.D07.R5240" & Bb_burden > 25.2) )
```

```{r merge_mammals, echo = TRUE}
# join the tables together
mammals <- left_join(
  mammals, df, by = c("plotID", "trapCoordinate", "collectDate", "tagID")
)
# check that the rows haven't changed
stopifnot(nrow(mammals) == n_mammals)

mammals <- mammals %>% 
  mutate(Bb_infected = (Bb_status == "Positive"))
```

## Genetic correlations

First, let's look at how the genetic traits are correlated with each other. 
The figure below shows that the genetic traits are nearly all significantly
intercorrelated (the lone exception being TGF-B and IL-6, which are 
uncorrelated). Notably, the resistance traits (IL-10, IFN-y, IL-6, TLR-2) are
most strongly correlated with each other and, similarly, the tolerance traits
(TGF-B and GATA-3) are most strongly correlated with each each other. Both
TGF-B and GATA-3 appear to be zero inflated and all the genetic traits 
are highly right-skewed (i.e., many low values).

```{r gene-cors, fig.width=7.5, fig.asp=0.95, warning=FALSE, message=FALSE, fig.cap=caption, cache=TRUE}
gene_expr_lookup <- tibble(
  column = df %>% select(`IL-10`:`GATA-3`) %>% names(),
  group = c(rep("resistance", 4), rep("tolerance", 2))
)

library(GGally)

mammals %>% 
  select(one_of(gene_expr_lookup$column)) %>% 
  ggpairs(progress = FALSE) + 
  scale_x_continuous(breaks = c(500, 1500))

caption = paste(
  "Genetic correlations"
)
```

### PCA

The resistance and tolerance genetic traits also broadly group together
in a PCA (expression variables log(x + 1) transformed prior to ordination): 

```{r gene-PCA, fig.asp = 1, fig.width = 5, fig.cap=caption}
library(vegan)

infect_tab <- mammals %>% 
  select(uid, Bb_burden, Bb_status, all_of(gene_expr_lookup$column)) %>% 
  filter_all(any_vars(!is.na(.))) %>% 
  mutate(Bb_infected = as.numeric(Bb_status == "Positive")) %>% 
  select(-Bb_status) %>% 
  relocate(Bb_infected, .after = Bb_burden) %>% distinct()

resp = infect_tab %>%
  filter(complete.cases(.)) %>% 
  select(uid, Bb_burden, Bb_infected) %>% 
  mutate(across(-c(uid), ~scale(.x))) 

preds = infect_tab %>% 
  filter(complete.cases(.)) %>% 
  select(-c(Bb_burden, Bb_infected)) %>% 
  mutate(across(-uid, ~scale(log(.x + 1))))

genetic_pca <- pca(preds %>% select(-uid) %>% as.matrix())

gen_sites <- scores(genetic_pca, display = "sites") %>% 
  as.data.frame() %>% 
  bind_cols(preds) %>% 
  left_join(resp)
gen_specs <- scores(genetic_pca, display = 'species') %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "variable")

Bb_envfit <- envfit(genetic_pca, resp[, c(3, 2)], permutations = 2000)

ggplot(gen_sites, aes(x = PC1, y = PC2)) + 
  geom_hline(yintercept = 0, linetype = "dashed", col = "grey50") + 
  geom_vline(xintercept = 0, linetype = "dashed", col = "grey50") + 
  geom_point(
    shape = 21, size = 2, alpha = 0.8, stroke = 0.5, fill = "cornflowerblue"
    # aes(col = log(Bb_burden))
  ) + 
  geom_segment(
    col = "forestgreen",
    data = gen_specs, aes(x = 0, y = 0, xend = 0.93*PC1, yend = 0.93*PC2)
  ) + 
  geom_segment(
    col = "orange",
    data = data.frame(Bb_envfit$vectors$arrows),
    aes(x = 0, y = 0, xend = 0.93*PC1, yend = 0.93*PC2)
  ) + 
  geom_text(
    col = "orange", 
    data = data.frame(Bb_envfit$vectors$arrows) %>% rownames_to_column(),
    aes(x = PC1, y = PC2, label = rowname)
  ) +
  geom_text(
    data = gen_specs, col = "forestgreen", aes(label = variable)
  ) +
  theme_bw() + 
  scale_color_gradient2(midpoint = 3)

caption = paste(
  "Genetic PCA"
)
```

It appears that Bb infection status is significantly associated with our
expression PCs, while burden is only marginally significant (based on
permutation tests):

```{r}
Bb_envfit
```

And below are visualizations of the relationships between the individual PC
axes and our Bb variables:

```{r, fig.asp = 1, fig.width=4}
gen_sites %>% 
  ggplot(aes(x = PC1, y = Bb_infected)) + 
  geom_point() + 
  stat_smooth(method = "lm") +
  labs(x = "PC1 (resistance)", y = "Bb infection prob.")

gen_sites %>% 
  ggplot(aes(x = PC2, y = Bb_infected)) + 
  geom_point() + 
  geom_smooth(method = "lm") + 
  labs(x = "PC2 (tolerance)", y = "Bb infection prob.")

gen_sites %>% 
  filter(Bb_burden > 0) %>% 
  ggplot(aes(x = PC1, y = Bb_burden)) + 
  geom_point() + 
  geom_smooth(method = "lm") +
  labs(x = "PC1 (resistance)", y = "Bb burden (infected)")

gen_sites %>% 
  filter(Bb_burden > 0) %>% 
  ggplot(aes(x = PC2, y = Bb_burden)) + 
  geom_point() + 
  geom_smooth(method = "lm") +
  labs(x = "PC2 (tolerance)", y = "Bb burden (infected)")
```

A multivariate regression of the Bb variables reveals that the significance
that we saw before is almost entirely due to the tolerance axis (PC2):

```{r}
gen_sites %>% 
  lm(cbind(Bb_infected, Bb_burden) ~ PC1 + PC2, data = .) %>% 
  summary()
```

Note though, that the R-squared values are quite low.

### Transformations

Because of the distribution of these genetic trait values, transformations
are likely necessary such as the square-root transformation shown in the next
figure, but we will also group these values by quartile and decile later on.

```{r quantcut_fun, include=FALSE}
#' Break a vector into a factor by quantile
#'
#' @param x numeric vector
#' @param n number of groups (default is 4; quartiles)
#' @param ordered: logical: should the factor be ordered?
#'
#' @returns a factor of quantiles 1:n
#' @export
#'
#' @examples
quant_cut <- function(x, n = 4, ordered = TRUE){
  probs <- seq(0, 1, length.out = n + 1)
  quant <- quantile(x = x, probs = probs, na.rm = TRUE, names = FALSE)
  findInterval(
    x = x, vec = quant, all.inside = TRUE
  ) %>% factor(levels = seq_len(n), ordered = ordered)
}
```

```{r long_data}
# make a longer version of the data
infect_long <- infect_tab %>% 
  pivot_longer(
    -c(uid, Bb_burden, Bb_infected), names_to = "marker", values_to = "gene_expr"
  ) %>% 
  mutate(
    marker = factor(marker, levels = gene_expr_lookup$column),
    marker_group = gene_expr_lookup$group[
      match(marker, gene_expr_lookup$column)
    ]
  ) %>% 
  relocate(marker_group, .before = marker)

# break each expression variable into quantiles
infect_long <- infect_long %>% 
  group_by(marker) %>% 
  mutate(
    expr_quarts = quant_cut(gene_expr),
    expr_quints = quant_cut(gene_expr, n = 10)
  )
```

```{r, include = FALSE}
# check quartiles
infect_long %>% 
  filter(!is.na(expr_quarts)) %>% 
  ggplot(aes(x = expr_quarts, y = gene_expr)) + 
  facet_wrap(~marker, scales = "free_y") + 
  geom_point(color = "grey", size = 1) +
  stat_summary(fun.data = "mean_se") + 
  stat_summary(fun = "mean", geom = "line", aes(group = 1)) +
  theme_bw()

# and deciles
infect_long %>% 
  filter(!is.na(expr_quints)) %>% 
  ggplot(aes(x = expr_quints, y = gene_expr)) + 
  facet_wrap(~marker, scales = "free_y") + 
  geom_point(color = "grey", size = 1) +
  stat_summary(fun.data = "mean_se") + 
  stat_summary(fun = "mean", geom = "line", aes(group = 1)) + 
  theme_bw()
```

Square-root or log transforming the resistance variables does help their 
distributions a bit, though they are still not fully Guassian. Unfortunately,
a lot of data is lost with log-transformations (due to zeros), especially 
among tolerance traits.

```{r Gene-hists, fig.asp = 1.5, fig.width=5, fig.cap=caption}
infect_long %>% 
    mutate(
    x = gene_expr, 
    `sqrt(x)` = sqrt(gene_expr),
    `ln(x)` = log(gene_expr)
  ) %>% 
  pivot_longer(x:`ln(x)`, names_to = "transform") %>% 
  mutate(transform = factor(transform, levels = c("x", "sqrt(x)", "ln(x)"))) %>% 
  ggplot(aes(x = value, fill = transform)) + 
  facet_grid(marker ~ transform, scales = "free") + 
  geom_histogram(col = "black", show.legend = FALSE)

caption = paste(
 "Genetic traits histogram, before and after transformations" 
)
```

## B.b. variable distributions

*Borrelia* burden data are heavily zero-inflated, unsurprisingly, and there
are about equal numbers of B.b. positive and negative individuals.

```{r Bb-hists, fig.asp = 0.5, fig.width=4, fig.cap=caption}
# show a histogram of the borrelia data
infect_tab %>% 
  select(Bb_burden, Bb_infected) %>% 
  pivot_longer(everything()) %>% 
  ggplot(aes(x = value)) + 
  facet_wrap(~name, scales = "free", strip.position = "bottom") + 
  geom_histogram() + 
  theme_bw() + 
  theme(
    strip.background = element_blank(), strip.placement = "outside"
  ) + 
  labs(x = NULL)

caption = paste(
  "B.b. variable histograms"
)
```

Similar to the genetic traits, square-root transforming the burden data helps
with the skew. Log-transforming does an even better job, but can't handle the
zeros. This latter option is likely to be used with an infection hurdle model.

```{r Bb-trans, fig.asp = 0.38, fig.width = 5, fig.cap=caption}
infect_tab %>% 
  select(Bb_burden) %>% 
  filter(!is.na(Bb_burden)) %>% 
  mutate(
    x = Bb_burden, 
    `sqrt(x)` = sqrt(Bb_burden),
    `ln(x)` = log(Bb_burden)
  ) %>% 
  pivot_longer(x:`ln(x)`, names_to = "transform") %>% 
  mutate(transform = factor(transform, levels = c("x", "sqrt(x)", "ln(x)"))) %>% 
  ggplot(aes(x = value, fill = transform)) + 
  facet_wrap(~transform, scales = "free") + 
  geom_histogram(col = "black", show.legend = FALSE) + 
  labs(x = "B.b. burden") + 
  theme_bw()

caption = paste(
  "distribution of B.b. burden before and after transformations"
)
```

## Infection status modeling

Because of the zero-inflation of the B.b. burden data, we will look at the 
factors affecting B.b. in two parts (i.e., hurdle model). The first part is
to look at the factors affecting the probability of being infected (binary) and
the second part is to model the burden given that an individual is infected
(i.e., zeros removed).

The figures below assess the relationships between the genetic traits and 
B.b. infection status. 

The first shows that, there aren't any strong patterns between the untransformed
genetic traits and B.b. status. It also reveals that there are some rather 
extreme outliers among the genetic traits:

```{r infect-raw, fig.asp = 0.75, fig.width = 6, fig.cap=caption}
# raw relationships:
infect_long %>% 
  filter(is.finite(gene_expr)) %>% 
  ggplot(aes(x = gene_expr, y = Bb_infected)) + 
  facet_wrap(~marker, scales = "free", strip.position = "bottom") + 
  geom_point(
    col = "black", shape = 21, 
    aes(y = jitter(as.numeric(Bb_infected), amount = 0.1))
  ) + 
  geom_smooth(
    method = "glm", aes(group = 1), col = "blue", fill = "grey80",
    method.args = list(family = "binomial"),
    formula = "y ~ x"
  ) + 
  theme_bw() + 
  scale_y_continuous(breaks = seq(0, 1)) + 
  theme(
    strip.background = element_blank(), strip.placement = "outside", 
  ) + 
  labs(x = NULL, y = "B.b. infection (probability)")

caption = paste(
  "Relationships between genetic traits and B.b. infection"
)
```

The next figure shows the relationship between the transformed variables and
infection status, with the top 1% of the genetic trait values removed. Note
that we log-transform the resistance traits and square-root-transform the
tolerance traits. It appears that the transformed resistance traits are still
not strongly associated with B.b. status (true for both transformations), but
the transformed tolerance traits do appear to be **positively** associated
with B.b. infection status. 

```{r infect-sqrt, fig.asp = 0.7, fig.width = 6, fig.cap=caption}
infect_long %>% 
  group_by(marker) %>% 
  mutate(top_pct = quantile(gene_expr, 0.99, na.rm = TRUE)) %>% 
  ungroup() %>% 
  mutate(
    trans_val = if_else(
      marker %in% c("TGF-B", "GATA-3"),
      sqrt(gene_expr), sqrt(gene_expr)
    ),
    trans = if_else(marker %in% c("TGF-B", "GATA-3"), "sqrt", "sqrt"),
  ) %>% 
  ggplot(aes(x = trans_val, y = as.numeric(Bb_infected), col = trans)) + 
  facet_wrap(~marker, scales = "free_x", strip.position = "bottom") + 
    geom_point(
    col = "black", shape = 21, 
    aes(y = jitter(as.numeric(Bb_infected), amount = 0.1))
  ) + 
  geom_smooth(
    method = "glm", formula = "y~x", method.args=list(family="binomial"),
  ) +
  theme_bw() + 
  scale_y_continuous(breaks = seq(0, 1)) + 
  theme(
    strip.background = element_blank(), strip.placement = "outside", 
  ) + 
  labs(x = NULL, y = "B.b. infection (probability)", col = "trans(x)")


caption = paste(
  "Relationships between sqrt-transformed genetic traits and B.b. infection"
)
```

If we break the genetic traits into quartiles, we observe a slightly stronger
patterns. Again, the strongest and clearest relationships are with the
tolerance traits. Note that the blue points are mean proportion of infected
individuals ± 95% confidence intervals for each quartile:

```{r infect-quarts, fig.asp = 0.7, fig.width = 6, fig.cap=caption, cache=TRUE}
infect_long %>% 
  mutate(
    marker = factor(
      paste(marker, "quartile"), levels = paste(levels(marker), "quartile")
    )
  ) %>% 
  ggplot(aes(x = as.numeric(expr_quarts), y = Bb_infected)) + 
  facet_wrap(~marker, scales = "free_y", strip.position = "bottom") + 
  geom_point(
    col = "black", shape = 21,
    aes(
      x = jitter(as.numeric(expr_quarts), amount = 0.2),
      y = jitter(as.numeric(Bb_infected), amount = 0.1)
    )
  ) +
  geom_smooth(
    method = "glm", aes(group = 1), col = "blue", fill = "grey80",
    method.args = list(family = "binomial"), formula = "y ~ x"
  ) + 
  stat_summary(
    fun.data = mean_cl_normal, col = "blue", shape = 21, fill = "white"
  ) + 
  theme_bw() + 
  scale_x_continuous(breaks = seq(1, 10)) + 
  scale_y_continuous(breaks = seq(0, 1)) + 
  theme(
    strip.background = element_blank(), strip.placement = "outside",
  ) + 
  labs(x = NULL, y = "B.b. infection (probability ± 95% CI)")

caption = paste(
  "Relationships between genetic traits quartiles and B.b. infection"
)
```

And the same patterns exist when we look at deciles:

```{r infect-decs, fig.asp = 0.7, fig.width = 6, fig.cap=caption, cache=TRUE}
infect_long %>% 
  mutate(
    marker = factor(
      paste(marker, "decile"), levels = paste(levels(marker), "decile")
    )
  ) %>% 
  ggplot(aes(x = as.numeric(expr_quints), y = Bb_infected)) + 
  facet_wrap(~marker, scales = "free_y", strip.position = "bottom") + 
  geom_point(
    col = "black", shape = 21,
    aes(
      x = jitter(as.numeric(expr_quints), amount = 0.2),
      y = jitter(as.numeric(Bb_infected), amount = 0.1)
    )
  ) +
  # geom_count(shape = 21, fill = "grey80") + scale_size_area() + 
  geom_smooth(
    method = "glm", aes(group = 1), col = "blue", fill = "grey80",
    method.args = list(family = "binomial"), formula = "y ~ x"
  ) + 
  stat_summary(
    fun.data = mean_cl_normal, col = "blue", shape = 21, fill = "white"
  ) + 
  theme_bw() + 
  scale_x_continuous(breaks = seq(1, 10)) + 
  scale_y_continuous(breaks = seq(0, 1)) + 
  theme(
    strip.background = element_blank(), strip.placement = "outside",
  ) + 
  labs(x = NULL, y = "B.b. infection (probability ± 95% CI)")

caption = paste(
  "Relationships between genetic traits deciles and B.b. infection"
)
```

### Statistical models

While most bivariate relationships are weak, the multivariate relationships
may be more interesting. This section conducts multiple regression modeling
of infection status from these 6 genetic traits.

Unfortunately, a generalized linear mixed effects (logistic regression) model
using raw responses reveals that shows no significant effects (and very small
effect sizes):

```{r infect-mod-raw,}
# load mixed models package
library(glmmTMB)

# Fit a raw GLMM binomial model, with site and year as random effects
infect_glmm <- mammals %>% 
  filter(
    sex %in% c("M", "F"),
  ) %>% 
  mutate(sexM = as.numeric(sex == "M")) %>% 
  glmmTMB(
    Bb_infected ~ `IL-10` + `IFN-y` + `IL-6` + `TGF-B` + `GATA-3` + 
      (1|siteID) + (1|year),
    data = ., family = binomial
  )

# get as summary of the model
coefficients(summary(infect_glmm))$cond %>% 
  knitr::kable(
    col.names = c("Coef", "SE", "z", "P"), digits = 3,
    caption = paste(
      "Infection logistic GLMM model, with raw genetic traits."
    )
  )
```

square-root transforming the predictor variables does not improve the 
significance:

```{r infect-mod-trans}
# Fet a seperate model, with sqrt transforms (log transform on resistance failed)
infect_glmm2 <- mammals %>% 
  filter(
    sex %in% c("M", "F"),
    is.finite(`IL-10`), is.finite(`IFN-y`), is.finite(`IL-6`),
    is.finite(`TGF-B`), is.finite(`GATA-3`)
  ) %>% 
  mutate(
    sexM = as.numeric(sex == "M"),
    across(c(`IL-10`:`IL-6`), ~sqrt(.x), .names = "sqrt_{.col}"),
    across(c(`TGF-B`:`GATA-3`), ~sqrt(.x), .names = "sqrt_{.col}")
  ) %>% 
  glmmTMB(
    Bb_infected ~ `sqrt_IL-10` + `sqrt_IFN-y` + `sqrt_IL-6` + 
      `sqrt_TGF-B` + `sqrt_GATA-3` + 
      (1|siteID) + (1|year),
    data = ., family = binomial
  )

# get the summary of that model
coefficients(summary(infect_glmm2))$cond %>% 
    knitr::kable(
      col.names = c("Coef", "SE", "z", "P"), digits = 3,
      caption = paste(
        "Infection logistic GLMM model, with sqrt-transformed",
        "genetic traits."
      )
    )
```

Nor does standardizing the predictor variables:

```{r infect-mod-scaled}
# Fet a seperate model, with sqrt transforms (log transform on resistance failed)
infect_glmm3 <- mammals %>% 
  filter(
    sex %in% c("M", "F"),
    is.finite(`IL-10`), is.finite(`IFN-y`), is.finite(`IL-6`),
    is.finite(`TGF-B`), is.finite(`GATA-3`)
  ) %>% 
  mutate(
    sexM = as.numeric(sex == "M"),
    across(c(`IL-10`:`GATA-3`), ~scale(.x), .names = "scaled_{.col}"),
  ) %>% 
  glmmTMB(
    Bb_infected ~ `scaled_IL-10` + `scaled_IFN-y` + `scaled_IL-6` + 
      `scaled_TGF-B` + `scaled_GATA-3` + 
      (1|siteID) + (1|year),
    data = ., family = binomial
  )

# get the summary of that model
coefficients(summary(infect_glmm3))$cond %>% 
    knitr::kable(
      col.names = c("Coef", "SE", "z", "P"), digits = 3,
      caption = paste(
        "Infection logistic GLMM model, with z-scaled",
        "genetic traits."
      )
    )
```

However, if we exclude the top 5% of all the genetic traits, then the two
tolerance variables (TGF-B and GATA-3) are significantly associated with
B.b. infection. Probability of infection *incraeses* with increasing TGF-B 
expression and *decreases* with GATA-3, after accounting for the effects of
the other variables:

```{r infect-mod-95pct}
# load mixed models package
library(glmmTMB)

# Fit a raw GLMM binomial model, with site and year as random effects
infect_glmm <- mammals %>% 
  mutate(
    sexM = as.numeric(sex == "M"),
    across(`IL-10`:`GATA-3`, ~quant_cut(.x, 100), .names = "grp_{.col}")
  ) %>% 
  filter(
    sex %in% c("M", "F"),
    is.finite(`IL-10`), is.finite(`IFN-y`), is.finite(`IL-6`),
    is.finite(`TGF-B`), is.finite(`GATA-3`),
    `grp_IL-10` <= 95, `grp_IFN-y` <= 95, `grp_IL-6` <= 95, 
    `grp_TGF-B` <= 95, `grp_GATA-3` <= 95
  ) %>% 
  glmmTMB(
    Bb_infected ~ `IL-10` + `IFN-y` + `IL-6` + `TGF-B` + `GATA-3` + 
      (1|siteID) + (1|year),
    data = ., family = binomial
  )

# get as summary of the model
coefficients(summary(infect_glmm))$cond %>% 
    knitr::kable(
      col.names = c("Coef", "SE", "z", "P"), digits = 3,
      caption = paste(
        "Infection logistic GLMM model, with extreme",
        "genetic traits removed."
      )
    )
```

The same is likely true for using quartiles/deciles, but I haven't fit those
models yet.

```{r, include = FALSE}
#TODO: now I need to look at the binary relationship. Hurdle models might best explain things
```

## B.b. burden

TBD: Because the infection status did not seem to be associated with any of the
genetic variables, I am going to hold off on running the second part of the 
analysis ($P(\text{burden}|\text{infected})$).

## Weather

One important component of the SEMs will be a measure of climate. PCA allows
us to take a series of climatic variables and simplify them to a single
index variable that encompasses them. Here we include both the mean and 
coefficient of variation for temperature, precipitation, and relative
humidity. We will also perform the PCA at three different scales: the site 
scale (average climate across 8 years), the annual scale (average climate
within a single year at a site), and the month scale (average climate within
a single month at a site).

```{r sumamrize_temperatures}
# Load temperature data
daily_temps <- readRDS("neon-data/data/eight_sites/daily-air-temperatures.rds")

# calculate yearly and monthly summaries of daily temperature data
temp_summaries <- daily_temps %>% 
  filter(year %in% 2017:2024) %>% 
  group_by(siteID, year, month) %>% 
  mutate(
    # daily
    month_min_daily_temp = min(temp_day_mean, na.rm = TRUE),
    month_avg_daily_temp = mean(temp_day_mean, na.rm = TRUE),
    month_max_daily_temp = max(temp_day_mean, na.rm = TRUE),
    month_sd_daily_temp = sd(temp_day_mean, na.rm = TRUE),
  ) %>% 
  group_by(siteID, year) %>% 
  mutate(
    # daily
    year_avg_daily_temp = mean(temp_day_mean, na.rm = TRUE),
    year_min_daily_temp = min(temp_day_mean, na.rm = TRUE),
    year_max_daily_temp = max(temp_day_mean, na.rm = TRUE),
    year_sd_daily_temp = sd(temp_day_mean, na.rm = TRUE),
    # monthly
    year_min_monthly_temp = min(month_avg_daily_temp, na.rm = TRUE),
    year_avg_monthly_temp = mean(month_avg_daily_temp, na.rm = TRUE),
    year_max_monthly_temp = max(month_avg_daily_temp, na.rm = TRUE),
    year_sd_monthly_temp = sd(month_avg_daily_temp, na.rm = TRUE),
    ) %>% 
  group_by(siteID) %>% 
  mutate(
    # daily
    site_min_daily_temp = min(temp_day_mean, na.rm = TRUE),
    site_avg_daily_temp = mean(temp_day_mean, na.rm = TRUE),
    site_max_daily_temp = max(temp_day_mean, na.rm = TRUE),
    site_sd_daily_temp = sd(temp_day_mean, na.rm = TRUE),
    # monthly
    site_min_monthly_temp = min(month_avg_daily_temp, na.rm = TRUE),
    site_avg_monthly_temp = mean(month_avg_daily_temp, na.rm = TRUE),
    site_max_monthly_temp = max(month_avg_daily_temp, na.rm = TRUE),
    site_sd_monthly_temp = sd(month_avg_daily_temp, na.rm = TRUE),
    # yearly
    site_min_yearly_temp = min(year_avg_daily_temp, na.rm = TRUE),
    site_avg_yearly_temp = mean(year_avg_daily_temp, na.rm = TRUE),
    site_max_yearly_temp = max(year_avg_daily_temp, na.rm = TRUE),
    site_sd_yearly_temp = sd(year_avg_daily_temp, na.rm = TRUE),
  ) %>% 
  ungroup() %>% 
  select(
    siteID, year, month, 
    starts_with("site_"), starts_with("year_"), starts_with("month_")
  ) %>% 
  distinct()
```

```{r summarize_precips}
# Load precipitation data
daily_precips <- readRDS("neon-data/data/eight_sites/all-daily-precips.rds")

# calculate yearly and monthly summaries of precip data.
## Note that this averages across the different precip colleciton methods
precip_summaries <- daily_precips %>% 
  filter(year %in% 2017:2024) %>% 
  group_by(siteID, date, year, month) %>% 
  group_by(siteID, year, month) %>% 
  mutate(
    # total
    month_total_precip = sum(daily_precip, na.rm = TRUE),
    # daily
    month_avg_daily_precip = mean(daily_precip, na.rm = TRUE),
    month_sd_daily_precip = sd(daily_precip, na.rm = TRUE),
    month_pct_daily_raindays = mean(daily_precip > 0, na.rm = TRUE),
    month_sd_daily_raindays = sd(daily_precip > 0, na.rm = TRUE)
  ) %>% 
  group_by(siteID, year) %>% 
  mutate(
    # total
    year_total_precip = sum(daily_precip, na.rm = TRUE),
    # daily
    year_avg_daily_precip = mean(daily_precip, na.rm = TRUE),
    year_sd_daily_precip = sd(daily_precip, na.rm = TRUE),
    year_pct_daily_raindays = mean(daily_precip > 0, na.rm = TRUE),
    year_sd_daily_raindays = sd(daily_precip > 0, na.rm = TRUE),
    # monthly
    year_avg_monthly_precip = mean(month_avg_daily_precip, na.rm = TRUE),
    year_sd_monthly_precip = sd(month_avg_daily_precip, na.rm = TRUE),
    year_pct_monthly_raindays = mean(month_pct_daily_raindays, na.rm = TRUE),
    year_sd_monthly_raindays = sd(month_pct_daily_raindays, na.rm = TRUE),
  ) %>% 
  group_by(siteID) %>% 
  mutate(
    # total
    site_total_precip = sum(daily_precip, na.rm = TRUE),
    # daily
    site_avg_daily_precip = mean(daily_precip, na.rm = TRUE),
    site_sd_daily_precip = sd(daily_precip, na.rm = TRUE),
    site_pct_daily_raindays = mean(daily_precip > 0, na.rm = TRUE),
    site_sd_daily_raindays = sd(daily_precip > 0, na.rm = TRUE),
    # monthly
    site_avg_monthly_precip = mean(month_avg_daily_precip, na.rm = TRUE),
    site_sd_monthly_precip = sd(month_avg_daily_precip, na.rm = TRUE),
    site_pct_monthly_raindays = mean(month_pct_daily_raindays, na.rm = TRUE),
    site_sd_monthly_raindays = sd(month_pct_daily_raindays, na.rm = TRUE),
    # yearly
    site_avg_yearly_precip = mean(year_avg_daily_precip, na.rm = TRUE),
    site_sd_yearly_precip = sd(year_avg_daily_precip, na.rm = TRUE),
    site_pct_yearly_raindays = mean(year_pct_daily_raindays, na.rm = TRUE),
    site_sd_yearly_raindays = sd(year_pct_daily_raindays, na.rm = TRUE),
  ) %>% 
  ungroup() %>% 
  select(
    siteID, year, month, 
    starts_with("site_"), starts_with("year_"), starts_with("month_")
  ) %>% 
  distinct()
```

```{r summarize_humidities}
# Relative humidity
daily_humids <- readRDS(
  "neon-data/data/eight_sites/daily-relative-humidities.rds"
)

humid_summaries <- daily_humids %>% 
  filter(year %in% 2017:2024) %>% 
  group_by(siteID, year, month) %>% 
  mutate(
    # daily
    month_min_daily_RH = min(RH_day_mean, na.rm = TRUE),
    month_avg_daily_RH = mean(RH_day_mean, na.rm = TRUE),
    month_max_daily_RH = max(RH_day_mean, na.rm = TRUE),
    month_sd_daily_RH = sd(RH_day_mean, na.rm = TRUE),  
  ) %>% 
  group_by(siteID, year) %>% 
  mutate(
    # daily
    year_min_daily_RH = min(RH_day_mean, na.rm = TRUE),
    year_avg_daily_RH = mean(RH_day_mean, na.rm = TRUE),
    year_max_daily_RH = max(RH_day_mean, na.rm = TRUE),
    year_sd_daily_RH = sd(RH_day_mean, na.rm = TRUE),  
    # monthly
    year_min_monthly_RH = min(month_avg_daily_RH, na.rm = TRUE),
    year_avg_monthly_RH = mean(month_avg_daily_RH, na.rm = TRUE),
    year_max_monthly_RH = max(month_avg_daily_RH, na.rm = TRUE),
    year_sd_monthly_RH = sd(month_avg_daily_RH, na.rm = TRUE),  
  ) %>% 
  group_by(siteID) %>% 
  mutate(
    # daily
    site_min_daily_RH = min(RH_day_mean, na.rm = TRUE),
    site_avg_daily_RH = mean(RH_day_mean, na.rm = TRUE),
    site_max_daily_RH = max(RH_day_mean, na.rm = TRUE),
    site_sd_daily_RH = sd(RH_day_mean, na.rm = TRUE),  
    # monthly
    site_min_monthly_RH = min(month_avg_daily_RH, na.rm = TRUE),
    site_avg_monthly_RH = mean(month_avg_daily_RH, na.rm = TRUE),
    site_max_monthly_RH = max(month_avg_daily_RH, na.rm = TRUE),
    site_sd_monthly_RH = sd(month_avg_daily_RH, na.rm = TRUE),     
    # yearly
    site_min_yearly_RH = min(year_avg_daily_RH, na.rm = TRUE),
    site_avg_yearly_RH = mean(year_avg_daily_RH, na.rm = TRUE),
    site_max_yearly_RH = max(year_avg_daily_RH, na.rm = TRUE),
    site_sd_yearly_RH = sd(year_avg_daily_RH, na.rm = TRUE),
  ) %>% 
  ungroup() %>% 
  select(
    siteID, year, month, 
    starts_with("site_"), starts_with("year_"), starts_with("month_")
  ) %>% 
  distinct()
```

Here is a quick look at the site-scale weather PCA, where PC1 explains
65% of the total variation and PC2 explains an additional 27%:

```{r site-climate-PCA, fig.asp = 1, fig.width = 4}
# Join the site-level summaries to use for PCA
site_climate_summaries <- temp_summaries %>% 
  select(siteID, starts_with("site_")) %>% distinct() %>% 
  mutate(
    site_CV_monthly_temp = site_sd_monthly_temp / site_avg_monthly_temp
  ) %>% 
  select(
    siteID, site_avg_monthly_temp, site_CV_monthly_temp
  ) %>% 
  full_join(
    precip_summaries %>% 
      select(siteID, starts_with("site_")) %>% distinct() %>% 
      mutate(
        site_CV_monthly_precip = site_sd_monthly_precip / site_avg_monthly_precip
      ) %>% 
      select(
        siteID, site_avg_monthly_precip, site_CV_monthly_precip
      )
  ) %>% 
  full_join(
    humid_summaries %>% 
      select(siteID, starts_with("site_")) %>% distinct() %>% 
      mutate(
        site_CV_monthly_RH = site_sd_monthly_RH / site_avg_monthly_RH
      ) %>% 
      select(
        siteID, site_avg_monthly_RH, site_CV_monthly_RH
      )
  )

# Conduct the PCA
site_climate_PCA <- site_climate_summaries %>% 
  select(-siteID) %>% 
  pca(scale = TRUE)

# rough plot of the PCA
site_climate_PCA %>% 
  ordiplot(xlim = c(-3.3, 2.1), ylim = c(-2, 2.1)) %>% 
  orditorp(display = "species")

summary(site_climate_PCA)$cont$importance
```
```{r calculate-sitePCA}
climSites <- scores(site_climate_PCA)$sites %>% data.frame() %>% 
  bind_cols(site_climate_summaries) %>% 
  mutate(nudge_x = 0, nudge_y = 0)
climSpecs <- scores(site_climate_PCA)$species %>% data.frame() %>% 
  rownames_to_column(var = "variable") %>% 
  mutate(nudge_x = 0, nudge_y = 0)
```

```{r, fig.asp = 1, fig.width=6, eval = FALSE}
climSpecs[grepl("CV_RH", climSpecs$variable), "nudge_y"] <- 0.05
climSpecs[grepl("avg_temp", climSpecs$variable), "nudge_y"] <- -0.05

climSites[grepl("UNDE", climSites$siteID), "nudge_y"] <- -0.05
climSites[grepl("SERC", climSites$siteID), "nudge_y"] <- 0.15

ggplot(data = climSpecs, aes(x = PC1, y = PC2)) + 
  geom_hline(yintercept = 0, linetype = "dashed", col = "grey50") + 
  geom_vline(xintercept = 0, linetype = "dashed", col = "grey50") + 
  geom_segment(
    aes(x = 0, xend = PC1, y = 0, yend = PC2), col = "black"
  ) + 
  geom_label(
    aes(label = gsub("(\\w*)_(\\w*)_(\\w*)", "\\3 \\2", variable), x = PC1 + nudge_x, y = PC2 + nudge_y)
  ) + 
  geom_label(
    data = climSites, show.legend = FALSE,
    aes(label = siteID, col = siteID, x = PC1 + nudge_x, y = PC2 + nudge_y)
  ) + 
  labs(x = "PC1 (65% of variation)", y = "PC2 (27% of variation)") + 
  theme_bw()

summary(site_climate_PCA)
```

Here is a breakdown of PC1 across the sites and the loadings of the different
climate variables:

```{r, fig.asp = 1, fig.width = 3.8}
climSites %>% 
  mutate(siteID = fct_reorder(siteID, desc(PC1))) %>% 
  ggplot(aes(x = siteID, y = PC1, fill = siteID)) + 
  geom_col(show.legend = FALSE) + 
  theme_bw() + 
  labs(x = "NEON site", y = "Climate PC1 (2017-2024)")

climSpecs %>% 
  select(variable, PC1) %>% 
  arrange(desc(PC1)) %>% 
  knitr::kable(digits = 3)
## sites with positive PC1 values are: warmer and wetter
```

Because this site-scale PCA seems to place relative humidity and precipitation on
opposite sides of the PC1 axis, so I'd like to visualize the relationship of 
these two variables with temperature. From this, we can see that at sites with
the highest lowest average temperatures, average humidity and precipitation are 
less linked than at the sites with more middling average temperatures:

```{r, fig.asp = 0.66, fig.width = 5}
climSites %>% 
  ggplot(
    aes(x = site_avg_monthly_temp, y = scale(site_avg_monthly_precip))
  ) +
  geom_point(aes(col = "precip")) + geom_line(aes(col = "precip")) +
  geom_point(aes(y = scale(site_avg_monthly_RH), col = "humidity")) + 
  geom_line(aes(y = scale(site_avg_monthly_RH), col = "humidity")) + 
  labs(
    x = "average annual site temperature", y = "scaled response (precip or humidity)",
    color = "response"
  )
```

Here is the annual-scale PCA, where PC1 explains 48% of the variation and 
PC2 explains an additional 22%. 

```{r year-climate-PCA, fig.asp = 1, fig.width = 4}
# Join the year-level summaries to use for PCA
year_climate_summaries <- temp_summaries %>% 
  select(siteID, starts_with("year_")) %>% distinct() %>% 
  mutate(
    year_CV_monthly_temp = year_sd_monthly_temp / year_avg_monthly_temp
  ) %>% 
  select(
    siteID, year_avg_monthly_temp, year_CV_monthly_temp
  ) %>% 
  full_join(
    precip_summaries %>% 
      select(siteID, starts_with("year_")) %>% distinct() %>% 
      mutate(
        year_CV_monthly_precip = year_sd_monthly_precip / year_avg_monthly_precip
      ) %>% 
      select(
        siteID, year_avg_monthly_precip, year_CV_monthly_precip
      )
  ) %>% 
  full_join(
    humid_summaries %>% 
      select(siteID, starts_with("year_")) %>% distinct() %>% 
      mutate(
        year_CV_monthly_RH = year_sd_monthly_RH / year_avg_monthly_RH
      ) %>% 
      select(
        siteID, year_avg_monthly_RH, year_CV_monthly_RH
      )
  )

# Conduct the PCA
year_climate_PCA <- year_climate_summaries %>% 
  select(-siteID) %>% 
  pca(scale = TRUE)

# rough plot of the PCA
year_climate_PCA %>% 
  ordiplot() %>% 
  orditorp(display = "species")

summary(year_climate_PCA)$cont$importance
```

The monthly climate PCA (not visualized) further reduces the explanatory power of the PC axes
with PC1 explaining only 30% of the climate variation and PC2 explaining an
additional 24%:

```{r month-climate-PCA}
# Join the month-level climate variables for PCA
month_climate_summaries <- temp_summaries %>% 
  select(siteID, year, month, starts_with("month_")) %>% distinct() %>% 
  mutate(
    month_CV_daily_temp = month_sd_daily_temp / month_avg_daily_temp
  ) %>% 
  select(
    siteID, month_avg_daily_temp, month_CV_daily_temp
  ) %>% 
  full_join(
    precip_summaries %>% 
      select(siteID, year, month, starts_with("month_")) %>% distinct() %>% 
      mutate(
        month_CV_daily_precip = month_sd_daily_precip / month_avg_daily_precip
      ) %>% 
      select(
        siteID, month_avg_daily_precip, month_CV_daily_precip
      )
  ) %>% 
  full_join(
    humid_summaries %>% 
      select(siteID, year, month, starts_with("month_")) %>% distinct() %>% 
      mutate(
        month_CV_daily_RH = month_sd_daily_RH / month_avg_daily_RH
      ) %>% 
      select(
        siteID, month_avg_daily_RH, month_CV_daily_RH
      )
  )
# Conduct the PCA
month_climate_PCA <- run_cache(
  month_climate_summaries %>% 
    select(-siteID) %>% 
    filter(complete.cases(.)) %>% 
    pca(scale = TRUE),
  file.path(cache_dir, "month-climate-PCA.rds")
)

summary(month_climate_PCA)$cont$importance
```

```{r, eval = FALSE}
# rough plot of the PCA
month_climate_PCA %>% 
  ordiplot() %>% 
  orditorp(display = "species")
```

## Resistance/Tolerance PCA

Next, we will try to reduce the resistance and tolerance genetic expression 
variables into PCA axes, as we did with weather. In this case, we will fit
separate PCAs for resistance (4 variables) and tolerance (2 variables) and will
use the first PC axis as the index.

```{r load_capturetimes}
# library(lme4)
library(glmmTMB)
library(DHARMa)
library(piecewiseSEM)

# read the capture time data
cap_times <- readRDS(
  "activity-timing/data/ibuttons/capture-time-trapping-data_2022-2024.rds"
)

# merge it with the mammals table
mammals <- left_join(mammals, cap_times)

mammals <- left_join(
  mammals,
  climSites %>% select(
    siteID, siteClimatePC1 = PC1, siteClimatePC2 = PC2, 
  ), by = "siteID"
)
stopifnot(nrow(mammals) == n_mammals)
```

The resistance PC1 explains 77% of the variation among the (log-transformed) 
resistance variables, and is well correlated with those variables 
(only one shown):

```{r resistance_PCA, fig.asp=1, fig.width=3.5}
# Resistance PCA
resistance_tab <- infect_tab %>% 
  select(uid, `IL-10`:`TLR-2`) %>% 
  filter(complete.cases(.)) %>% 
  distinct() %>% 
  mutate(across(-uid, ~log(.x + 1), .names = "trans_{.col}")) %>% 
  select(-c(`IL-10`:`TLR-2`))

res_pca <- vegan::pca(resistance_tab %>% select(-uid))

res_scores <- scores(res_pca)

resistance_tab <- bind_cols(resistance_tab, data.frame(res_scores$sites)) %>% 
  rename(res_PC1 = PC1, res_PC2 = PC2)

resistance_tab %>% 
  ggplot(aes(x = res_PC1, y = `trans_IL-10`)) + 
  geom_point() + geom_smooth(method = "lm") +
  geom_smooth(aes(y = `trans_TLR-2`), col = "green", method = "lm") +
  theme_bw()

summary(res_pca)$cont$importance %>% 
  data.frame() %>% round(3)
```

The tolerance PC1 explains 97% of the variation in the two (log-transformed) 
tolerance variables:

```{r tolerance_PCA, fig.asp=1, fig.width=3.5}
# Tolerance PCA
tolerance_tab <- infect_tab %>% 
  select(uid, `TGF-B`, `GATA-3`) %>% 
  filter(complete.cases(.)) %>% 
  distinct() %>% 
  mutate(across(-uid, ~log(.x + 1), .names = "trans_{.col}")) %>% 
  select(-c(`TGF-B`, `GATA-3`))

tol_pca <- vegan::pca(tolerance_tab %>% select(-uid))

tol_scores <- scores(tol_pca)

tolerance_tab <- bind_cols(tolerance_tab, data.frame(tol_scores$sites)) %>% 
  mutate(PC1 = -1*PC1) %>% 
  rename(tol_PC1 = PC1, tol_PC2 = PC2)

tolerance_tab %>% 
  ggplot(aes(x = tol_PC1, y = `trans_GATA-3`)) + 
  geom_point() + geom_smooth(method = "lm") +
  geom_smooth(aes(y = `trans_TGF-B`), method = "lm", col = "green") +
  theme_bw()

summary(tol_pca)$cont$importance %>% 
  data.frame() %>% round(3)

gen_tab <- full_join(resistance_tab, tolerance_tab, by = "uid")
```

```{r}
mammals <- left_join(mammals, gen_tab, by = "uid")
stopifnot(nrow(mammals) == n_mammals)
```

## SEMs:

In this section, we will look at SEMs that involve resistance and tolerance. 
Specifically, we are interested in resistance and tolerance as *latent* 
variables for tick attachment and *Borrelia* infection. 

### Tick attachment

First, we'll build a model whose ultimate response variable is tick attachment.
These models will include resistance and tolerance as predictors of tick 
attachment as well as factors that might predict resistance and tolerance. We
will also try to incorporate mouse behavior (e.g., capture time) into these
models.

```{r}
Peros <- mammals %>% 
  filter(
    taxonID %in% c("PELE", "PEMA")
  ) %>% 
  mutate(
    sex = if_else(sex == "U", NA, sex),
    sex_male = as.numeric(sex == "M"),
    cap_prop_rescaled = scales::rescale(cap_prop_night, to = c(0.025, 0.975)),
    ticks_attatched = if_else(
      larvalTicksAttached | nymphalTicksAttached | adultTicksAttached,
      1, 0
    ), 
    sex_mature = vagina %in% c("neither", "swollen", "plugged") |
      nipples %in% c("enlarged") |
      testes %in% c("scrotal"),
    sex_mature = as.numeric(sex_mature)
  )

mod_dat <- Peros %>% 
  filter(taxonID == "PELE") %>% 
  select(
    siteID, tagID, year, cap_prop_night, sex_male, weight, ticks_attatched, 
    siteClimatePC1, sex_mature,
    res_PC1, tol_PC1, 
    # Bb_burden, Bb_infected,
  ) %>% 
  # mutate(Bb_infected = as.numeric(Bb_infected), log_Bb_burden = log(Bb_burden + 1)) %>% 
  filter(complete.cases(.))
```

#### model 1

The first model tries to replicate the model that Alli built, but adding an
additional year of data. This model is not identical to the one Alli fit:
rather than fitting the sub-models (lmer and glmer) to the *individual* 
scale (by summarizing variables across recaptures), we use fit to the individual
*observation* scale (including individual ID as a random effect). This prevents
the need to summarize and maximizes the available information (and power).

```{r}
# Alli's lmer models:
# 1. logit_prop_elapsed ~ sex + mean_weight + (1 | siteID)
# 2. propticsattached_logit ~ PC3_Climate + prop_elapsed_logit + sex*prop.reproductive + (1|siteID)
# 3. PC1_Resistance ~ PC3_Climate*year + prop.reproductive + mean_weight + sex + (1|siteID)
# 4. PC2_Tolerance ~ mean_weight + year + sex*prop_elapsed + propticksattached*prop.reproductive + (1|siteID)

# variables
# prop_elapsed = proportion of the night 
# mean_weight = individual's average body weight (why?)
# propticksattached = proportion of captures that ticks were attached (why?)

#1. ----
act_fm <- lmer(
  formula = cap_prop_night ~ sex_male + weight + (1|siteID) + (1|tagID), 
  data = Peros
)

#2. ----
tick_fm <- glmer(
  formula = ticks_attatched ~ siteClimatePC1 + cap_prop_night + sex_male*sex_mature + (1|siteID) + (1|tagID),
  data = Peros, family = "binomial"
)

#3. ----
resist_fm <- lmer(
  formula = res_PC1 ~ siteClimatePC1*year + sex_mature + weight + sex_male + (1|siteID) + (1|tagID),
  data = Peros,
)

#4. ----
# 4. PC2_Tolerance ~ mean_weight + year + sex*prop_elapsed + propticksattached*prop.reproductive + (1|siteID)
toler_fm <- lmer(
  formula = tol_PC1 ~ weight + year + sex_male*cap_prop_night + ticks_attatched*sex_mature + (1|siteID) + (1|tagID),
  data = Peros
)
```

The results from this model indicate that there are some important paths missing
from the model (the effect of sexual maturity on capture timing and the 
correlation between tolerance and resistance). 

The results are somewhat interesting:

* The probability (results on logit-link scale) of having a tick attached is associated with site 
level climate (P=0.003), capture time (P=0.4), and there is a significant
interaction between sex and sexual maturity (P=0.4). Notably the main effects
for sex and sexual maturity are not significant.

* Capture time is affected by both sex and weight of the individual

* Resistance is differs by year, sex, and weight

* Tolerance differs only by year


```{r}
# construct the pSEM
fsem1 <- psem(
  act_fm,
  tick_fm,
  resist_fm,
  toler_fm,
  data = Peros
)

# analyze the pSEM
smry_fsem1 <- run_cache(
  summary(fsem1, .progressBar = FALSE),
  file.path(cache_dir, "psem1-model-summary.rds")
)


smry_fsem1
```

Below is a visualization of the sex-maturity interaction:

```{r fig.asp = 0.9, fig.width=3.5}
tmp <- expand_grid(
  sex = c(0, 1), mature = c(0, 1), sex_eff = 0.28, mature_eff = 0.01, inx = 0.5,
) %>% 
  mutate(
    resp = sex*sex_eff + mature*mature_eff + sex*mature*inx,
    prob = boot::inv.logit(resp)
  )

tmp %>% 
  ggplot(aes(x = mature, y = prob, col = factor(sex))) +
  geom_point() + geom_line() + 
  labs(x = "Mature", col = "Sex", y = "Predicted P(tick attachment)") +
  theme_bw() + 
  scale_x_continuous(breaks = c(0, 1), labels = c("no", "yes")) + 
  scale_color_discrete(labels = c("F", "M")) + 
  theme(
    legend.position = "inside", legend.position.inside = c(1, 0.5),
    legend.justification = c(1, 0.5), legend.background = element_blank()
  )
```

Next, we update the model to include the suggested missing components. 
Non-significant paths have not yet been removed.

```{r}
# construct the pSEM
fsem1.corr <- psem(
  act_fm %>% update(. ~ . + year), 
  tick_fm, 
  resist_fm,
  toler_fm %>% update(. ~ . + res_PC1), 
  data = Peros
)

# analyze the pSEM
smry_fsem1.corr <- run_cache(
  summary(fsem1.corr, .progressBar = FALSE),
  file.path(cache_dir, "psem1Corr-model-summary.rds")
)

smry_fsem1.corr
```

Here's a visualization of the model (significant paths only):

```{r, fig.width = 3}
graphic_path <- R.utils::getAbsolutePath(
  file.path(
    "infection-modeling/graphical-models",
    "resistance-tolerance_model1.drawio.png"
    ) 
  )

knitr::include_graphics(graphic_path, rel_path = FALSE, error = FALSE)
```

There are a few problems with this model, from my perspective:

* There are some missing links that seem obvious (i.e., the effect of climate
on behavior).

* Year is treated as a numeric variable, which expects a constant trend through
time. This assumption is not reasonable. Better approaches would be to either 
1) include year as a random effect (note the lack of an interaction with year)
or 2) conduct a group comparison, whereby the model is parameter differently
for each year and testing which paths differ among years.

* The resistance and tolerance variables have a directed relationship. It is 
more appropriate to specify that these variables are adirectionally correlated.
I tried fitting the model with this specification multiple times and it would
not successfully fit.

* There are no significant links between ticks and resistance or tolerance, 
rather it is a model of the common factors that affect ticks, resistance, and 
tolerance. This is not completely uninteresting, but it does seem like there 
should be a direct link. 

```{r, eval = FALSE}
library(DiagrammeR)

tmp <- smry_fsem1$coefficients %>% 
  select(Response, Predictor, Std.Estimate, P.Value) %>% 
  filter(P.Value <= 0.1)

nodes <- tibble(
  label = unique(c(tmp$Response, tmp$Predictor)),
  id = seq_len(length(label))
)

gph <- DiagrammeR::create_graph() %>% 
  add_nodes_from_table(nodes, label_col = label) %>% 
  add_edges_from_table(tmp, from_col = Predictor, to_col = Response, from_to_map = label)


render_graph(gph)
# DiagrammeR::create_edge_df()
```

#### model 2

[**UNDER CONSTRUCTION**]

```{r, include=FALSE}
#1. ----
act_fm <- glmer(
  formula = cap_prop_night ~ siteClimatePC1 + sex_male + weight + (1|siteID) + (1|tagID) + (1|year), 
  data = mod_dat, family = "gaussian"
)
summary(act_fm)

#2. ----
tick_fm <- glmer(
  formula = ticks_attatched ~ siteClimatePC1 + cap_prop_night + sex_male*sex_mature + (1|siteID) + (1|tagID) + (1|year),
  data = mod_dat, family = "binomial"
)
summary(tick_fm)

#3. ----
resist_fm <- lmer(
  formula = res_PC1 ~ siteClimatePC1 + sex_mature + weight + sex_male + (1|siteID) + (1|tagID) + (1|year),
  data = mod_dat, #family = "gaussian"
)
summary(resist_fm)

#4. ----
# 4. PC2_Tolerance ~ mean_weight + year + sex*prop_elapsed + propticksattached*prop.reproductive + (1|siteID)
toler_fm <- lmer(
  formula = tol_PC1 ~ res_PC1 + weight + sex_male*cap_prop_night + ticks_attatched*sex_mature + (1|siteID) + (1|tagID) + (1|year),
  data = mod_dat, #family = "gaussian"
)
summary(toler_fm)
```

```{r}
res_tol_sem <- piecewiseSEM::psem(
  act_fm,
  tick_fm,
  resist_fm,
  toler_fm,
  data = mod_dat
)

res_tol_sem_smry <- run_cache(
  summary(res_tol_sem, .progressBar = FALSE),
  file.path(cache_dir, "psem2-model-summary.rds")
)

res_tol_sem_smry
```

```{r}
res_tol_sem_smry$coefficients %>% knitr::kable()

alt_psem <- psem(
  glmer(
    formula = cap_prop_night ~ res_PC1 + tol_PC1 + siteClimatePC1 + sex_male + sex_mature + weight + 
      (1|siteID) + (1|tagID) + (1|year), 
    data = mod_dat, family = "gaussian"
  ),
  glmer(
    formula = ticks_attatched ~ res_PC1 + tol_PC1 + siteClimatePC1 + sex_mature + sex_male +
      (1|siteID) + (1|tagID) + (1|year),
    data = mod_dat, family = "binomial"
  ),
  lmer(
    formula = res_PC1 ~ siteClimatePC1 + sex_mature + weight + sex_male +
      (1|siteID) + (1|tagID) + (1|year),
    data = mod_dat
  ),
  lmer(
    formula = tol_PC1 ~ weight + sex_mature + sex_male + 
      (1|siteID) + (1|tagID) + (1|year),
    data = mod_dat
  ),
  # lmer(
  #   log_Bb_burden ~ ticks_attatched + tol_PC1 + res_PC1 + weight + siteClimatePC1 + (1|siteID) + (1|tagID) + (1|year),
  #   data = mod_dat
  # ),
  # glmer(
  #   Bb_infected ~ ticks_attatched + tol_PC1 + res_PC1 + weight + siteClimatePC1 + cap_prop_night  + (1|siteID) + (1|tagID) + (1|year),
  #   data = mod_dat, family = "binomial"      
  # ),
  res_PC1 %~~% tol_PC1,
  data = mod_dat
)

alt_psem_smry <- run_cache(
  summary(alt_psem, .progressBar = FALSE),
  file.path(cache_dir, "psem2Alt-model-summary.rds")
)

alt_psem_smry
```

```{r}
alt_psem_smry$coefficients %>% knitr::kable(digits = 2)

alt_psem_smry$dTable
```

## Question: What coverage do we have for recaptures within a capture event?

In total there are nearly 8000 individual *Peromyscus* individuals that were
captured multiple times within a trapping event:

```{r, fig.asp = 0.9, fig.width = 6}
recaps <- mammals %>% 
  group_by(year, siteID, eventID, tagID) %>% 
  mutate(event_cap_count = n()) %>% 
  filter(event_cap_count > 1, !is.na(tagID))

nrow(recaps)

recaps %>% 
  filter(taxonID %in% c("PELE", "PEMA", "PELEPEMA")) %>% 
  group_by(year, siteID, eventID) %>% 
  summarize(recap_individuals = length(unique(tagID))) %>% 
  arrange(siteID, year, eventID) %>% 
  group_by(year, siteID) %>% 
  mutate(event_num = row_number()) %>% 
  ggplot(aes(x = event_num, y = recap_individuals, label = recap_individuals)) +
  facet_wrap(~siteID, scales = "free") + 
  geom_col(aes(fill = factor(year))) + 
  labs(
    y = "Peromyscus individuals recaptured",
    fill = "Year", 
    x = "Trap event number (within year)",
    title = "All recaptured Peromyscus"
  ) + 
  theme_bw()
 
# total individuals 
recaps %>% 
  filter(taxonID %in% c("PELE", "PEMA", "PELEPEMA")) %>% 
  group_by(year, siteID, eventID) %>% 
  summarize(recap_individuals = length(unique(tagID))) %>% 
  pull(recap_individuals) %>% sum()
```

And of those individuals that were recaptured, over 1000 were also captured (and
recaptured) in traps with ibuttons, allowing for estimation of capture time:

```{r, fig.asp = 0.9, fig.width = 6}
mammals %>% 
  group_by(year, siteID, eventID, tagID) %>% 
  mutate(event_cap_count = n()) %>% 
  filter(
    event_cap_count > 1, !is.na(tagID),
    !is.na(cap_prop_night), taxonID %in% c("PELE", "PEMA", "PELEPEMA")
  ) %>% 
  group_by(year, siteID, eventID) %>% 
  summarize(recap_individuals = length(unique(tagID))) %>% 
  arrange(siteID, year, eventID) %>% 
  group_by(year, siteID) %>% 
  mutate(event_num = row_number()) %>% 
  ggplot(aes(x = event_num, y = recap_individuals, label = recap_individuals)) +
  facet_wrap(~siteID, scales = "free") + 
  geom_col(aes(fill = factor(year))) + 
  labs(
    y = "Peromyscus individuals recaptured",
    fill = "Year", 
    x = "Trap event number (within year)",
    title = "Recaptured Peromyscus with estimated capture times"
  ) + 
  theme_bw() + 
  theme(
    legend.position = "inside", legend.position.inside = c(.9, 0), legend.justification = c(1, 0)
  )

# total individuals with capture times
mammals %>% 
  group_by(year, siteID, eventID, tagID) %>% 
  mutate(event_cap_count = n()) %>% 
  filter(
    event_cap_count > 1, !is.na(tagID),
    !is.na(cap_prop_night), taxonID %in% c("PELE", "PEMA", "PELEPEMA")
  ) %>% 
  group_by(year, siteID, eventID) %>% 
  summarize(recap_individuals = length(unique(tagID))) %>% 
  pull(recap_individuals) %>% sum()
```

Below is a table containing the counts of *Peromyscus* recaptures, for which 
we have capture time estimates and multiple captures within a session, that 
breaks down whether or not an individual was recaptured with the same tick 
status (`changed_status`) as it had the first time it was captured 
(`initial_status`) for each tick life stage (`tickStage`). Note that 
`tickStage == "any"` refers to any tick regardless of life stage. Overall, 
relatively few individuals were recaptured with a different tick status
than their first capture:

```{r}
cap_tick_tab <- mammals %>% 
  group_by(year, siteID, eventID, tagID) %>% 
  mutate(event_cap_count = n()) %>% 
  filter(
    event_cap_count > 1, !is.na(tagID),
    !is.na(cap_prop_night), taxonID %in% c("PELE", "PEMA", "PELEPEMA")
  ) %>% 
  arrange(year, siteID, eventID, tagID, collectDate) %>% 
  mutate(
    cap_num = row_number(),
    cap_time_change = cap_prop_night - first(cap_prop_night)
  ) %>% 
  rowwise() %>% 
  mutate(
    anyTicksAttached = any(
      larvalTicksAttached, nymphalTicksAttached, adultTicksAttached
    )
  ) %>% 
  ungroup() %>% 
  filter(!is.na(anyTicksAttached)) %>% 
  pivot_longer(
    c(
      larvalTicksAttached, nymphalTicksAttached, adultTicksAttached, 
      anyTicksAttached
    ), names_to = "tickStage", values_to = "attachStatus"
  ) %>% 
  mutate(
    tickStage = factor(
      tickStage, levels = c(
        "larvalTicksAttached", "nymphalTicksAttached", "adultTicksAttached", 
        "anyTicksAttached"
      ), labels = c("larva", "nymph", "adult", "any")
    )
  ) %>% 
  group_by(year, siteID, eventID, tagID, tickStage) %>% 
  arrange(year, siteID, eventID, tagID, tickStage, collectDate) %>% 
  mutate(
    initial_status = first(attachStatus),
    changed_status = attachStatus != initial_status
  )

cap_tick_tab %>% 
  filter(cap_num > 1) %>% 
  group_by(tickStage, initial_status, changed_status) %>% 
  tally() %>% 
  filter(!is.na(initial_status)) %>% 
  knitr::kable(
    caption = paste(
      "Recaptured Peromyscus tick attachment status counts,",
      "by tick lifestage"
    )
  )

```

```{r, fig.asp = 1, fig.width = 4.5}
cap_tick_tab %>% 
  filter(cap_num > 1, tickStage == "any") %>% 
  ggplot(aes(x = attachStatus, y = cap_time_change)) + 
  geom_hline(yintercept = 0, linetype = "dashed", col = "grey80") + 
  # geom_boxplot(aes(fill = initial_status)) +
  stat_summary(
    fun.data = mean_cl_normal, aes(fill = initial_status),
    position = position_dodge(0.8), shape = 21, col = "black",
    size = 0.5
  ) + 
  labs(
    y = "Change in capture time (± 95% CI)",
    x = "Ticks attached (recapture)",
    fill = "Ticks attached (initial)"
  ) + 
  theme_bw() + 
  theme(
    legend.position = "inside", legend.position.inside = c(0, 0),
    legend.justification = c(0, 0), legend.background = element_blank()
  )
```

There does appear to be be a trend whereby individuals that are recaptured
later tend to have a higher probability of being observed with a tick:

```{r, fig.asp=1, fig.width=5}
cap_tick_tab %>% 
  filter(cap_num > 1, tickStage == "any") %>% 
  mutate(started = factor(
    initial_status, levels = c(FALSE, TRUE),
    labels = c("without ticks", "with ticks"))
  ) %>% 
  ggplot(aes(x = cap_time_change, y = as.numeric(attachStatus), col = started)) + 
  geom_hline(yintercept = 0.5, col = "grey80", linetype = "dashed") + 
  geom_vline(xintercept = 0.0, col = "grey80", linetype = "dashed") + 
  # facet_wrap(~started) + 
  geom_point(shape = 21, size = 1) + 
  geom_smooth(
    method = "glm", method.args = list(family = "binomial"),
    fill = "grey80"
  ) + 
  geom_smooth(
    aes(group = 1, col = "either"),
    method = "glm", method.args = list(family = "binomial"),
    fill = "grey50"
  ) + 
  labs(
    y = "P(ticks on recapture)", x = "Change in capture time",
    fill = "Initially captured", col = "Initially captured"
  ) + 
  theme_bw() + 
  theme(
    legend.position = "inside", legend.position.inside = c(0.45, 0.58),
    legend.justification = c(0.5, 0.5), 
    legend.background = element_blank()
  ) + 
  scale_color_viridis_d(option = "plasma", end = 0.8, direction = -1)

ggsave("infection-modeling/graphics/res-tol-analysis/ticks-by-capturetimeshift.jpg",
       width = 5, height = 1*5, dpi = 300)
```

But that relationship is not significant 
(accounting for year, site, event, and individual as random effects). The only
factor that strongly predicts whether a recapture has a tick is if the 
initial capture had a tick:

```{r, echo = TRUE}
library(lmerTest)  

tick_time_fm <- glmer(
  attachStatus ~ cap_time_change * initial_status + 
    (1|year) + (1|siteID) + (1|eventID) + (1|tagID),
  data = cap_tick_tab %>% 
    filter(cap_num > 1, tickStage == "any") %>% 
    mutate(
      attachStatus = as.numeric(attachStatus),
      initial_status = as.numeric(initial_status)
    )
)  

# fixed effects
summary(tick_time_fm)$coefficients

# random effects
VarCorr(tick_time_fm)

# Type II ANOVA
car::Anova(tick_time_fm)
```
And if we only look at those that started without ticks, the effect is still
nonsignificant:

```{r, echo = TRUE}
upd <- update(
  tick_time_fm, . ~ . - initial_status - cap_time_change:initial_status,
  data = cap_tick_tab %>% 
    filter(cap_num > 1, tickStage == "any", !initial_status) %>% 
    mutate(
      attachStatus = as.numeric(attachStatus),
      initial_status = as.numeric(initial_status)
    ) 
  )

car::Anova(upd)
```

**Is tick status (and the change thereof) associated with change in relative
capture timing?**
