---
title: "SEM cookbook"
author: "Clay Morrow"
date: "2025-11-17"
output: 
  bookdown::html_document2: default
  bookdown::pdf_document2: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
library(glmmTMB)
library(lme4)
library(lmerTest)
library(piecewiseSEM)
library(semEff)
```

# Intro

This document outlines the basics of piecewiseSEMs. To render this notebook
use the following code (or similar).

```{r, eval = FALSE}
rmarkdown::render("infection-modeling/R/rmarkdown/SEM-cookbook.Rmd")
```

# Basic Usage

First, we will load the `keeley` dataset from the `piecewiseSEM` package to use
as our example:

```{r loadKeeley}
# load example data
data("keeley", package = "piecewiseSEM")

# scale and center all columns
keeley_scaled <- keeley %>% tibble() %>% 
  mutate(across(c(everything()), ~as.vector(scale(.x))))
```

## Initial Fit

Next we'll fit the component models that will comprise our SEM. These can be
many kinds of model fit objects (`lm`, `glm`, `lme4`, `glmmTMB`, etc.). The
important things to keep in mind here are that each model represents all the 
*direct* paths into a response variable and that **there can be no loops** in
the model (i.e., if Y is a response of X, then no direct or indirect paths from
Y can go back to X). In the case of our example SEM, we will fit 3 models: 
1) post-fire species richness dependent on ground cover, 2) ground cover
dependent on fire severity, and 3) fire severity dependent upon pre-fire stand
age. For this example, we'll just use simple linear regressions.

```{r compMods}
# Top level: richness affected by plant cover
rich_mod <- lm(rich ~ cover, data = keeley)

# middle level: plant cover affected by fire severity
cover_mod <- lm(cover ~ firesev, data = keeley)

# bottom level: fire severity affected by stand age prior to fire 
fire_mod <- lm(firesev ~ age, data = keeley)
```

Next we will combine these component models into the SEM, using the `psem`
function. Then we will actually perform the fit using the `summary.psem` method:

```{r pSEM}
# specify SEM
rich_psem <- psem(rich_mod, cover_mod, fire_mod, data = keeley)

# fit the SEM
rich_fit <- summary(rich_psem, .progressBar = FALSE)
```

The fitted SEM object contains many components (use `names(rich_fit)` to see 
what they are called). We'll go through the key ones in this example. First, 
let's take a look at the tests of d-separated independence 
(Table \@ref(tab:dTab)), which evaluates
whether there are any paths that might be important for our model that we did
not include. Here, there are 3 unconsidered paths: the effect of age on cover,
the effect of age on richness, and the effect of fire severity on richness. 
The p-values indicate that the age->richness path is significant (P < 0.05) 
and should be included in the SEM. The other p-values are also marginally
significant (P < 0.1), so these could also be included. For the sake of this
example, we will only add the age->richness path.

```{r dTab, fig.cap=caption}
rich_fit$dTable # %>% 
  # knitr::kable(
  #   digits = 3,
  #   caption = "D-separated independence claims"
  # )
caption = paste()
```

## Update fit

To do so, we can update our component models with the new variables:

```{r reFit}
# add age and firesev to rich mod
new_rich <- update(rich_mod, . ~ . + age)

# updated SEM spec
new_psem <- psem(new_rich, cover_mod, fire_mod, data = keeley)

# updated SEM spec
new_rich_fit <- summary(new_psem, .progressBar = FALSE)
```

Then we again check for any additional missing paths (\@ref(tab:dSepNew), 
no change):

```{r dSepNew}
new_rich_fit$dTable #%>% 
  # knitr::kable(
  #   digits = 3,
  #   caption = "Second d-sep table"
  # )
```

Now that we have an updated SEM, let's check its overall fit. The `piecewiseSEM`
packages provides two goodness of fit tests Fischer's C test and Chi-squared 
test. Both test if the SEM is *different from an optimal model* (i.e., one that
describes all variation in the data). However, this means that well-fitting 
models should have p-values *greater* than 0.05 (indicating that they are not
different; 
[source link](https://jslefche.github.io/sem_book/local-estimation.html#)). 
Since both of our goodness-of-fit p-values are less than .05, we can conclude
that the overall fit is relatively poor. Our next steps would be to try and find
a model that better fits the data (add new variables, change path orders, etc)
but for the simplicity of this example, we will proceed with the model as is.

```{r}
## Overall tests of model fit (explains variation in the full dataset)
new_rich_fit$Cstat %>% 
  bind_rows(new_rich_fit$ChiSq) %>% 
  mutate(
    stat = if_else(!is.na(Chisq), "Chisq", "Fisher.C"),
    val = if_else(!is.na(Chisq), Chisq, Fisher.C)
  ) %>% 
  select(stat, val, df, P.Value)
```

Additionally, we can look at the fits of the component models. The below table
shows that each component model has a reasonable $R^2$ value:

```{r}
new_rich_fit$R2
```

Note that in some instances, scaling and centering the data can sometimes help
with model fitting if the variables are on vastly different scales. But that
is not the case for this dataset:

```{r}
tmp <- psem(
  new_rich %>% update(. ~ ., data = keeley_scaled), 
  cover_mod %>% update(. ~ ., data = keeley_scaled), 
  fire_mod %>% update(. ~ ., data = keeley_scaled), 
  data = keeley_scaled
)
smry <- summary(tmp, .progressBar = FALSE)
smry$Cstat %>% 
  bind_rows(smry$ChiSq) %>% 
  mutate(
    stat = if_else(!is.na(Chisq), "Chisq", "Fisher.C"),
    val = if_else(!is.na(Chisq), Chisq, Fisher.C)
  ) %>% 
  select(stat, val, df, P.Value)
```

## Model coefficients

Finally, we can look at the model coefficients (\@ref(fig:coefTab)). It is
recommended to use P $\leq$ 0.1 for determining path significance. By this 
metric, all paths are significant.

```{r coefTab}
new_rich_fit$coefficients #%>% 
  # data.frame() %>% 
  # knitr::kable(
  #   digits = 2,
  #   col.names = c(
  #     "Resp.", "Pred.", "Coef. Est.", "SE", "DF", "Crit", "P", 
  #     "Std. Effect", "Sig."
  #   ),
  #   caption = paste(
  #     "SEM coefficient table: Response variable, Predictor variable",
  #     "Regression coefficient, Standard Error, Degrees of freedom,",
  #     "Critical value, P-value, Standardized effect size, and significance",
  #     "stars."
  #   )
  # )
```

And we can plot a simple version of the SEM using the `DiagrammeR` package:

```{r plotSEM}
library(DiagrammeR)

# copy data
path_tab <- new_rich_fit$coefficients %>% 
  data.frame()

# build table of nodes containing a node name and numeric ID
nodes_tab <- tibble(
  name = unique(c(path_tab$Response, path_tab$Predictor))
) %>% 
  mutate(id = row_number())

# add some plotting aesthetics to the path table (optional)
path_tab <- path_tab %>% 
  mutate(
    ## width of arrows proportional to effect size
    penwidth = scales::rescale(Std.Estimate, to = c(1, 5)),
    ## color of arrows by direction
    color = if_else(Std.Estimate <= 0, "red", "blue"),
    ## linetype of the arrow by significance
    style = if_else(P.Value <= 0.1, "solid", "dashed"),
    ## edge labels (near center)
    label = round(Std.Estimate, 2)
  )

# create the graph
sem_graph <- create_graph(attr_theme = "bt") %>% 
  add_nodes_from_table(nodes_tab, label_col = "name") %>% 
  add_edges_from_table(
    path_tab, from_col = "Predictor", to_col = "Response",
    from_to_map = "label"
  ) %>% 
  ## set some universal node attributes
  set_node_attrs("color", "black") %>% 
  set_node_attrs("fillcolor", "white")

# render the graph
render_graph(sem_graph)
```

## Cumulative effects

Next, let's show how to calculate the cummulative (direct + indirect) effects
in an SEM. Let's use the effects of age on richness as our example.
There are two paths from age to fire severity in the above SEM: 1) the direct
effect of age on richness (-0.21) and 2) the indirect effect of age on fire
severity *through* fire severity and ground cover. The cummulative indirect 
effect of age on richness is equal to the product of the direct effects of age
on fire severity (0.45), fire severity on cover (-0.44), and cover on 
richness (0.25). Then, the total effect of age on richness is the sum of this 
indirect effect and the direct effect 
(i.e., $-0.21 + (0.45 \times -0.44 \times -0.25) = -0.29$).

The following code calculates all the indirect and total effects and then 
plots them:

```{r}
## function to get all coefficients along a path
get_coefs_along_path <- function(p, edges, nodes){
  # early return for a missing path
  if (length(p) == 1 | all(is.na(p))){
    return(NA)
  }
  # convert the path numbers into names
  v <- nodes$name[p]
  # direct effect
  if (length(p) == 2) {
    ## return the direct effect
    edges %>%
      filter(Predictor == v[1], Response == v[2]) %>%
      pull(Std.Estimate) %>% return()
  }
  # indirect effects
  ## join into a lag table
  tab <- cbind(from = head(v, -1), to = tail(v, -1))
  ## go through all the rows of the table and find the matching coefficient
  apply(
    tab, 1, function(x) {
      pull(filter(edges, Predictor == x[1], Response == x[2]), Std.Estimate)
    }
  ) %>% return()
}
```

```{r}
# build a table to fill
breakdown <- expand_grid(
  Response = unique(path_tab$Response), Predictor = unique(path_tab$Predictor)
) %>% 
  ## response can't equal predictor
  filter(Response != Predictor) %>%
  ## add IDs for response and predictors
  mutate(
    resp_ID = nodes_tab$id[match(Response, nodes_tab$name)],
    pred_ID = nodes_tab$id[match(Predictor, nodes_tab$name)]
  ) %>% 
  ## add direct effects and associated P-value
  left_join(
    path_tab %>% 
      select(Response, Predictor, direct_P = P.Value, direct = Std.Estimate),
    by = c("Response", "Predictor")
  ) %>% 
  ## prepare to loop through each row
  rowwise() %>% 
  ## perform operations to each row...
  mutate(
    ## get a list of all possible paths from the predictor to the response
    all_paths = list(unique(get_paths(sem_graph, from = pred_ID, to = resp_ID))),
    ## get a list of coefficients along those paths
    all_coefs = list(
      lapply(all_paths, function(x)get_coefs_along_path(x, path_tab, nodes_tab))
    ),
    ## multiply the coefficients along a path
    coef_prods = list(lapply(all_coefs, prod)),
    ## get the total effect, by adding up the products
    total = sum(unlist(coef_prods), na.rm = TRUE),
    ## calculate indirect effects by subtracting the direct effects from the total
    indirect = if_else(!is.na(direct), total - direct, total - 0)
  ) %>% 
  ## remove the rowwise grouping
  ungroup()
```

```{r brkdwnPlot, warning=FALSE}
# plot the breakdown
breakdown %>% 
  select(Response, Predictor, direct_P, direct, indirect, total) %>% 
  mutate(
    Resposne = factor(Response, levels = nodes_tab$name),
    Predictor = factor(Predictor, levels = nodes_tab$name)
  ) %>% 
  ggplot(aes(x = Predictor)) + 
  facet_wrap(~Response, labeller = label_both) + 
  geom_col(aes(y = total, fill = "total"), col = "black", width = 1) + 
  geom_col(
    aes(y = direct, fill = "direct"), 
    col = "black", width = 0.4, position = position_nudge(x = 0.2)
  ) + 
  geom_col(
    aes(y = indirect, fill = "indirect"), 
    col = "black", width = 0.4, position = position_nudge(x = -0.2)
  ) + 
  theme_bw() + 
  labs(x = "Predictor", y = "Effect", fill = NULL)
```

# semEff

The `semEff` package can be used to obtain bootstrapped confidence intervals, 
and/or to automatically extract indirect and total effects:

```{r, cache = TRUE, warning = FALSE}
# the SEM as a list of models
modlist <- list(richness = new_rich, cover = cover_mod, firesev = fire_mod)

# boot strap the model (with 10 iterations), using default settings
sem_boot <- bootEff(modlist, R = 10)
```

```{r}
# calculate the effects from the bootsrapped model
sem_effs <- semEff(sem_boot, ci.conf = 0.90)

# print a summary of all effects
summary(sem_effs)
```

And to extract the indirect effects, use `getIndEff`. Below we can see that
there is an indirect effect of fireseverity on richness, but no indirect
effects on cover or fire. 

```{r}
# just extract indirect effects
getIndEff(sem_effs, type = "orig")
```

`getDirEff` and `getMedEff` can also be used to extract direct and mediator
effects, respectively. Mediator effects are simply the sum of all indirect 
paths which operate through the mediator variable.

## TODO:

* [x] look into mediators

* [x] binomial response / categorical predictors

* [ ] work through example together?

# Hannah Example

* 4 (blocks) fencinc manipulations - vegetative differences may be driving seed removal
* 36 depots (9 per depot) - `deployment_id`
* 3 different habitat types
* 215 acorns placed 
* 4 1m^2 quadrats, 10m radias trees (depot level)

```{r, eval = FALSE}
# load the acorn-squirrel data set
acorn_df <- read.csv("infection-modeling/R/Rmarkdown/acorn_squirrel.csv") %>% 
  rename(depotID = deployment_id) %>% 
  mutate(
    squirrels = log(SCCA_ppd + SCNI_ppd + 1),
    num.left = initial.ct - num.gone, prop.left = num.left / initial.ct,
    SCCA_ppd = log(SCCA_ppd + 1), SCNI_ppd = log(SCNI_ppd + 1), # log transform
    type = factor(
      type, levels = c("pine", "high", "med"), 
      labels = c("pine", "highOak", "medOak")
    ),
    type.num = as.numeric(type)
    # across(c(depotID, fence, type), ~factor(.x)) # convert to factor
  )
summary(acorn_df)
```

```{r, fig.asp = 0.85, fig.width = 3.5, eval = FALSE}
acorn_df %>% 
  ggplot(aes(x = type, y = prop.gone, col = type)) + 
  geom_boxplot(fill = NA) +  
  geom_point(size = 2) + 
  geom_smooth(method = "lm", aes(group = 1), fill = "grey80", col = "black") + 
  stat_summary(geom = "line", aes(group = 1), linetype = "dotted", col = "grey20") + 
  stat_summary(
    fun.data = mean_se, geom = "pointrange", 
    shape = 21, fill = "white", size = 1, linewidth = 1
  ) + 
  guides(col = "none") + 
  theme_bw()
```


## Together as lab group

```{r}

```

## Clay's try

```{r removal_binom, eval = FALSE}
# fit a binomial mixed model (with removals serving as our "success")
## Here, removal is a function of squirrels, canopy cover, bare ground, 
## forest type, forast basal area, and fencing type as well as
## random ID to control for group covariance
remove_fm <- glmer(
  cbind(num.gone, num.left) ~ squirrels + open.can +  
    type.num +
    (1|fence),
  family = "binomial", data = acorn_df
)
summary(remove_fm)

## Note that it does not make sense to include deployment_ID (depotID) as a
## random effect, because there are as many IDs as there are rows in these data
## (i.e., no repeated measures from any ID)
```

```{r, eval = FALSE}
squirrels_fm <- lmer(
  squirrels ~ open.can + type.num + (1|fence),
  data = acorn_df
)
```


```{r grey_pois, evall = FALSE}
# fit a model of grey squirrel density as a function of fox squirrel density,
## canopy cover, bare ground, forest basal area, forest type, fencing trt, 
## and random ID
grey_fm <- lmer(
  SCCA_ppd ~ open.can + type.num + 
    (1|fence),
  data = acorn_df
)
summary(grey_fm)
car::Anova(grey_fm)
```

```{r fox_pois, eval = FALSE}
# fox squirrel density as a function of canopy cover, bare ground, 
## forest basal area, forest type, fencing trt, and random ID
fox_fm <- lmer(
  SCNI_ppd ~ open.can + type.num + (1|fence),
  data = acorn_df
)
summary(fox_fm)
car::Anova(fox_fm)
```


```{r soil_lm, eval = FALSE}
# fit a model of bare soil as a function of canopy cover, BA, forest type, 
## and fencing trt
bare_fm <- lm(
  bare.soil ~ open.can,
  data = acorn_df
)
summary(bare_fm)
car::Anova(bare_fm)
```

```{r canopy_lm, eval = FALSE}
# fit a model of canopy cover as a function of BA, forest type, 
## and fencing trt
canopy_fm <- lm(
  open.can ~ total_BA,
  data = acorn_df
)
summary(canopy_fm)
car::Anova(canopy_fm)
```

```{r BA_fm, eval = FALSE}
basal_fm <- lm(
  total_BA ~ type.num, 
  data = acorn_df
)
summary(basal_fm)
car::Anova(basal_fm)
```

```{r SEM, eval = FALSE}
modlist = list(
  remove_fm, 
  squirrels_fm
  # bare_fm,
  # # basal_fm,
  # canopy_fm
)
```

```{r, message = TRUE}
removal_psem <- as.psem(object = modlist)
remove_psem_fit <- summary(removal_psem)
```

```{r, eval = FALSE}
remove_sem <- semEff(modlist, R = 100, data = acorn_df, ran.eff = "fence")

remove_sem$Summary
```
